### 2019-08-29 ml4ps_paper/paper.tex
Add initial paper.

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2019

% ready for submission
% \usepackage{neurips_2019}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2019}

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[preprint]{neurips_2019}

% to avoid loading the natbib package, add option nonatbib:
%     \usepackage[nonatbib]{neurips_2019}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\setcitestyle{numbers}
\setcitestyle{square}
\setcitestyle{comma}
\bibliographystyle{abbrvnat}
\setlength{\bibsep}{4pt plus 2pt minus 2pt}

\title{Scalable Extreme Deconvolution}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{
  James A. Ritchie\\
  School of Informatics\\
  University of Edinburgh\\
   \texttt{james.ritchie@ed.ac.uk} \\
  \And
  Iain Murray\\
  School of Informatics\\
  University of Edinburgh\\
   \texttt{i.murray@ed.ac.uk} \\
}

\begin{document}

\maketitle

\begin{abstract}

The Extreme Deconvolution method is a mixture-model based approach original intended to perform density estimation on astrometry datasets.
Newer much larger datasets are available to which the existing method for fitting the model cannot scale to.
We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via SGD.
We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.

\end{abstract}

\section{Introduction}

\section{Background}

\begin{equation}
\mathcal{L}(\theta) = \sum_i \log \sum_j \alpha_j\mathcal{N}(\mathbf{w}_i \mid \mathbf{m}_j, T_{ij})
\end{equation}

\section{Methods}

\subsection{Online Expectation-Maximisation}
The original method for fitting XD models used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
For the extreme deconvolution case, the expected sufficient statistics that need to be tracked are $q_{ij}$, $q_{ij}\mathbf{v}_i$, and $q_{ij}\mathbf{v}_i\mathbf{v}_i^T$.
The M-step comprises normalisation of the sums of expected sufficient statistics to get the mixture component parameters.

\citet{cappeOnlineExpectationMaximization2009} proposed an online variation of EM for latent data models.
Their method involves computing a stochastic approximation to the sums of expected sufficient statistics.
At each iteration $t$, the sums of expected sufficient statistics $s_t$ are computed over a minibatch, and the stochastic estimate $\hat{s}_t$ updated as
\begin{equation}
 \hat{s}_{t} = (1 - \lambda)\hat{s}_{t-1} + \lambda s_t
\end{equation}
where $\lambda$ is a sufficiently small step-size.
$\lambda$ should theoretically decrease at each iteration according to the Robbins-Monro conditions, but in practice we found that using a constant step-size was sufficient.
After each update of $\hat{s}_t$, we update the parameters.

\subsection{Stochastic Gradient Descent}

An alternative to EM-based methods is to optimise the log-likelihood directly.
The log-likelihood has some constraints on its parameters, namely that the weight $a_j$ sum to $1$ and that the covariances $V_j$ are positive definite.
Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$,
\begin{equation}
\alpha_j = \frac{e^{z_j}}{\sum_k^K e^{z_k}}.
\end{equation}
We parametrise the covariances $V_j$ via the Cholesky decomposition, 
\begin{equation}
  V_j = L_jL_j^T
\end{equation}
where $L_j$ is a lower-triangular matrix.
Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
For the extreme deconvolution model, this cannot be done, as we need to form the covariance $Tij$ for every datapoint.
In practice we found that using the Adam optimisation algorithm eliminates the need to do this~\cite{kingmaAdamMethodStochastic2014}.

\section{Experiments}

We implemented both of our methods in PyTorch.
The use of PyTorch allows us to take advantage of its automatic differentiation framework and in-built optimisers for the SGD method,
as well as allowing us to write code targeting both CPU and GPU-based computation.

To evaluate our methods, we used a subset of rows from the Gaia DR2 source table \cite{brownGaiaDataRelease2018}.
We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
In total there were approximately 2 million rows.
Where data were missing, we set the field to zero, the variance to a large value.
This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
We used the same number of components $K=128$ for each model, following 
In practice we would want to select a value of $K$ by cross-validation.

Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
Initial covariances were set to the identity matrix.
Early stopping was done based on a held-out validation set comprising 10\% of the rows.
Final model performance was evaluated on another held-out test set also comprising 10\% of the rows.

\begin{table}
  \caption{Validation and test log-likelihoods for the Gaia data subset.}
  \label{results-table}
  \centering
  \begin{tabular}{lcc}
    \toprule
    Method     & Validation     & Test \\
    \midrule
    Existing EM & -  & -     \\
    Online EM     & - & -      \\
    SGD     & -       & -  \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Discussion}

\bibliography{references}

\end{document}

### 2019-08-31 deconv/gmm/base.py
Update regularisation.

@@ -10,11 +10,11 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 class BaseGMM(ABC):
 
-    def __init__(self, components, dimensions, max_iters=100,
+    def __init__(self, components, dimensions, epochs=100,
                  w=1e-6, tol=1e-9, restarts=5, device=None):
         self.k = components
         self.d = dimensions
-        self.max_iters = max_iters
+        self.epochs = epochs
         self.w = w
         self.tol = tol
         self.restarts = restarts
@@ -47,7 +47,7 @@ class BaseGMM(ABC):
 
             prev_log_prob = torch.tensor(float('-inf'), device=self.device)
 
-            for i in range(self.max_iters):
+            for i in range(self.epochs):
                 log_prob, expectations = self._e_step(data)
                 if log_prob == n_inf:
                     break

### 2019-08-31 deconv/gmm/deconv_gmm.py

@@ -8,30 +8,12 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 class DeconvGMM(BaseGMM):
 
-    def __init__(self, components, dimensions, max_iters=1000, gamma=1,
-                 omega=None, eta=0, m_hat=0, w=1e-6, tol=1e-6, restarts=5,
-                 device=None):
-        super().__init__(components, dimensions, max_iters=max_iters, tol=tol,
+    def __init__(self, components, dimensions, epochs=1000,
+                 w=1e-6, tol=1e-6, restarts=5, device=None):
+        super().__init__(components, dimensions, epochs=epochs, tol=tol,
                          restarts=restarts, device=device)
 
-        if len(torch.tensor(gamma).shape) == 0:
-            self.gamma = gamma * torch.ones(self.k, 1, device=self.device)
-        else:
-            self.gamma = gamma
-
-        self.eta = eta
-        if len(torch.tensor(m_hat).shape) == 0:
-            self.m_hat = m_hat * torch.ones(self.d, device=self.device)
-        else:
-            self.m_hat = m_hat
-
-        if omega is None:
-            self.omega = (self.d + 1) / 2
-
-        if len(torch.tensor(w).shape) == 0:
-            self.w = w * torch.eye(self.d, device=self.device)
-        else:
-            self.w = w
+        self.w = w * torch.eye(self.d, device=self.device)
 
     def _init_expectations(self, data):
 

### 2019-08-31 deconv/gmm/online_deconv_gmm.py

@@ -7,21 +7,22 @@ from .util import minibatch_k_means
 
 class OnlineDeconvGMM(DeconvGMM):
 
-    def __init__(self, components, dimensions, max_iters=10000, gamma=1,
-                 omega=None, eta=0, w=1e-6, tol=1e-6, step_size=0.1,
-                 batch_size=100, restarts=5, max_no_improvement=20,
-                 device=None):
-        super().__init__(components, dimensions, max_iters=max_iters,
-                         gamma=gamma, omega=omega, eta=eta, w=w, tol=tol,
+    def __init__(self, components, dimensions, epochs=1000, w=1e-6,
+                 tol=1e-6, step_size=0.1, batch_size=100, restarts=5,
+                 max_no_improvement=20, k_means_factor=100,
+                 k_means_iters=10, device=None):
+        super().__init__(components, dimensions, epochs=epochs, w=w, tol=tol,
                          restarts=restarts, device=device)
         self.batch_size = batch_size
         self.step_size = step_size
         self.max_no_improvement = max_no_improvement
+        self.k_means_factor = k_means_factor
+        self.k_means_iters = k_means_iters
 
     def _init_sum_stats(self, loader, n):
 
         counts, centroids = minibatch_k_means(
-            loader, self.k, device=self.device
+            loader, self.k, max_iters=self.k_means_iters, device=self.device
         )
 
         self.weights = counts[:, None] / counts.sum()
@@ -46,11 +47,19 @@ class OnlineDeconvGMM(DeconvGMM):
             drop_last=True
         )
 
+        init_loader = data_utils.DataLoader(
+            data,
+            batch_size=self.k_means_factor * self.batch_size,
+            num_workers=4,
+            shuffle=True,
+            pin_memory=True
+        )
+
         n = len(data)
 
-        n_inf = torch.tensor(float('-inf'), device=self.device)
+        n_inf = float('-inf')
 
-        best_ll = torch.tensor(float('-inf'), device=self.device)
+        best_ll = float('-inf')
 
         for j in range(self.restarts):
 
@@ -58,50 +67,50 @@ class OnlineDeconvGMM(DeconvGMM):
             if val_data:
                 val_ll_curve = []
 
-            self._init_sum_stats(loader, n)
+            self._init_sum_stats(init_loader, n)
 
-            prev_ll = torch.tensor(float('-inf'), device=self.device)
-            max_val_ll = torch.tensor(float('-inf'), device=self.device)
[…]

### 2019-08-31 deconv/gmm/sgd_deconv_gmm.py

@@ -44,9 +44,10 @@ class SGDDeconvDataset(data_utils.Dataset):
 class SGDDeconvGMM(BaseSGDGMM):
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
-                 batch_size=64, tol=1e-6, restarts=5, device=None):
-        self.module = SGDDeconvGMMModule(components, dimensions, device)
+                 batch_size=64, tol=1e-6, w=1e-3, restarts=5, device=None):
+        self.module = SGDDeconvGMMModule(components, dimensions, w, device)
         super().__init__(
             components, dimensions, epochs=epochs, lr=lr,
-            batch_size=batch_size, tol=tol, restarts=restarts, device=device
+            batch_size=batch_size, w=w, tol=tol, restarts=restarts,
+            device=device
         )

### 2019-08-31 deconv/gmm/sgd_gmm.py

@@ -13,7 +13,7 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 class SGDGMMModule(nn.Module):
 
-    def __init__(self, components, dimensions, device=None):
+    def __init__(self, components, dimensions, w, device=None):
         super().__init__()
 
         self.k = components
@@ -33,6 +33,8 @@ class SGDGMMModule(nn.Module):
         self.d_idx = torch.eye(self.d, device=self.device).to(torch.bool)
         self.l_idx = torch.tril_indices(self.d, self.d, -1, device=self.device)
 
+        self.w = w * torch.eye(self.d, device=device)
+
     @property
     def L(self):
         L = torch.zeros(self.k, self.d, self.d, device=self.device)
@@ -42,7 +44,7 @@ class SGDGMMModule(nn.Module):
 
     @property
     def covars(self):
-        return torch.matmul(self.L, torch.transpose(self.L, -2, -1))
+        return torch.matmul(self.L, torch.transpose(self.L, -2, -1)) + self.w
 
     def forward(self, data):
 
@@ -64,13 +66,14 @@ class BaseSGDGMM(ABC):
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
                  batch_size=64, tol=1e-6, restarts=5, max_no_improvement=20,
-                 k_means_factor=100, k_means_iters=10, device=None):
+                 k_means_factor=100, w=1e-6, k_means_iters=10, device=None):
         self.k = components
         self.d = dimensions
         self.epochs = epochs
         self.batch_size = batch_size
         self.tol = 1e-6
         self.lr = lr
+        self.w = w
         self.restarts = restarts
         self.k_means_factor = k_means_factor
         self.k_means_iters = k_means_iters
@@ -226,12 +229,13 @@ class BaseSGDGMM(ABC):
         self.module.l_diag.data = nn.Parameter(torch.zeros(self.k, self.d, device=self.device))
         self.module.l_lower.data = torch.zeros(self.k, self.d * (self.d - 1) // 2, device=self.device)
 
+
 class SGDGMM(BaseSGDGMM):
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
-                 batch_size=64, tol=1e-6, device=None):
-        self.module = SGDGMMModule(components, dimensions, device)
+                 batch_size=64, tol=1e-6, w=1e-3, device=None):
+        self.module = SGDGMMModule(components, dimensions, w, device)
[…]

### 2019-08-31 ml4ps_paper/paper.tex

@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage[preprint]{neurips_2019}
+\usepackage{neurips_2019}
 
@@
 \usepackage{amsfonts}       % blackboard math symbols
+\usepackage{amsmath}
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
+\usepackage{microtype}
+\usepackage{color}      % microtypography
 
@@
 \setlength{\bibsep}{4pt plus 2pt minus 2pt}
+
+\newcommand{\bx}{\mathbf{x}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bm}{\mathbf{m}}
+\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
 
@@
 
-The Extreme Deconvolution method is a mixture-model based approach original intended to perform density estimation on astrometry datasets.
+The Extreme Deconvolution method is a mixture-model based approach originally intended to perform density estimation on noisy astronomy datasets with known noise covariances for each datapoint.
 Newer much larger datasets are available to which the existing method for fitting the model cannot scale to.
-We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via SGD.
+We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via SGD, both which can make use of GPU computation.
 We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
@@
 
+Many scientific applications require an estimate of some density $p(\bx)$ to be fitted using a set of samples $\bx_i$.
+In many cases the dataset is noisy, but we wish to fit our density estimate to the true underlying noiseless data.
+Extreme Deconvolution is a method that allows Gaussian Mixture Models (GMMs) to be fitted to data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
+The method was originally developed to perform density estimation on the dataset produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
+The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
+
+The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
+The existing method of fitting extreme deconvolution models cannot cope with such a dataset, as it requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.
+In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to such datasets, based on an online variation of the Expectation-Maximisation (EM) algorithm, and stochastic gradient descent (SGD) on the log-likelihood.
+We develop implementations that can leverage GPU-based computation, and show that they provide comparable density estimates to the existing method, whilst being much faster to train.
+
 \section{Background}
 
+The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a true value $\bv_i$,
 \begin{equation}
-\mathcal{L}(\theta) = \sum_i \log \sum_j \alpha_j\mathcal{N}(\mathbf{w}_i \mid \mathbf{m}_j, T_{ij})
+  \mathbf{x_i} = \bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
 \end{equation}
+We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
+\begin{equation}
+p(\bv_i \mid \theta) = \sum_j^K \alpha_j \mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K
+\end{equation}
+As both the noise model is Gaussian and the density estimation model is a mixture of Gaussians, the likelihood of $\bx_i$ is also a Gaussian mixture,
+and the total log-likehood of the dataset is
+\begin{equation}
+\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bv_i \mid \mathbf{m}_j, T_{ij}), \quad T_{ij} = V_j + S_i.
+\end{equation}
+The original model also includes a method for handling the case where $\bx_i$ has been transformed from $\bv_i$ via a per-datapoint projection matrix $R_i$, which we omit here for simplicity.
+Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
 
@@
 \subsection{Online Expectation-Maximisation}
-The original method for fitting XD models used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-For the extreme deconvolution case, the expected sufficient statistics that need to be tracked are $q_{ij}$, $q_{ij}\mathbf{v}_i$, and $q_{ij}\mathbf{v}_i\mathbf{v}_i^T$.
+The original method for fitting extreme deconvolution models used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
+In the E-step, the expected sufficient statistics for each datapoint that need to be computed are $s_{ij} = \{q_{ij}$, $q_{ij}\bv_i$, and $q_{ij}\bv_i\bv_i^T\}$.
 The M-step comprises normalisation of the sums of expected sufficient statistics to get the mixture component parameters.
@@
 Their method involves computing a stochastic approximation to the sums of expected sufficient statistics.
-At each iteration $t$, the sums of expected sufficient statistics $s_t$ are computed over a minibatch, and the stochastic estimate $\hat{s}_t$ updated as
+At each iteration $t$, the sums of expected sufficient statistics $s_t = \sum_i s_{ij}$ are computed over a minibatch using the current estimate of the parameters, and the stochastic estimate $\hat{s}_t$ updated as
 \begin{equation}
@@
 $\lambda$ should theoretically decrease at each iteration according to the Robbins-Monro conditions, but in practice we found that using a constant step-size was sufficient.
-After each update of $\hat{s}_t$, we update the parameters.
+After each update, the parameters of the model are updated by normalising $\hat{s}_t$ appropriately.
+
+\todo{More complete explanation if space.}
+
+A direct implementation suggests that we should update $V_j$ as
+\begin{equation}
+V_j  = \mathrm{E}(q_{ij} \bv_ij \bv_ij^T) - \bm_j \bm_j^T
+\end{equation}
+This is usually acceptable when using double-precision floats, but using GPU based computation restricts us to single precision.
+Performing the update of $V_j$ like this with single-precision floats will result in catastrophic cancellation if the means are sufficiently large compared to the variance.
+Instead, we compute smoothed versions of the sum of squared deviation products which avoids this issue.
+\begin{equation}
+\end{equation}
+
+\todo{Better justification of why this works, or alternate method.}
 
@@
 where $L_j$ is a lower-triangular matrix.
+We also constrain the diagonal elements of $L_j$ to be positive by taking the log of them.
 Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
 For the extreme deconvolution model, this cannot be done, as we need to form the covariance $Tij$ for every datapoint.
+Having removed the constraints, we can then take the gradient of the log-likelihood with respect to the parameters and optimise it using a standard minibatch-based optimiser.
+Pre-multiplying the gradient with the inverse Fisher information matrix may be necessary to keep the covariances stable during optimisation with standard stochastic gradient descent.
 In practice we found that using the Adam optimisation algorithm eliminates the need to do this~\cite{kingmaAdamMethodStochastic2014}.
@@
 In total there were approximately 2 million rows.
-Where data were missing, we set the field to zero, the variance to a large value.
+Where data were missing, we set the field to zero and the noise variance to a large value.
 This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used the same number of components $K=128$ for each model, following 
+We used the same number of components $K=128$ for each model, following \cite{andersonImprovingGaiaParallax2018}.
 In practice we would want to select a value of $K$ by cross-validation.
@@
 Early stopping was done based on a held-out validation set comprising 10\% of the rows.
-Final model performance was evaluated on another held-out test set also comprising 10\% of the rows.
+Final model performance was evaluated on another held-out test set comprising another 10\% of the rows.
+Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
 
 \begin{table}
-  \caption{Validation and test log-likelihoods for the Gaia data subset.}
+  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
   \label{results-table}
   \centering
-  \begin{tabular}{lcc}
+  \begin{tabular}{lccc}
     \toprule
-    Method     & Validation     & Test \\
+    Method     & Validation     & Test & Training Time (s) \\
     \midrule
@@
     Online EM     & - & -      \\
-    SGD     & -       & -  \\
+    SGD     & $-5.468 \times 10^6 \pm 2.46 \times 10^3$        & - & $1469 \pm 11$ \\
     \bottomrule
@@
 
+\todo{Add plot of log-likelihood results against dataset size.}
+
 \section{Discussion}
+
+Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst being able to work with datasets that cannot fit in memory.
+The results also show that using GPU-based computation speeds up training considerably.
+
+Further improvements to our approaches are possible.
+The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
+It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and online EM methods.
+Our approaches also add more free parameters to be selected, including learning rate and batch size.
+This adds scope for hyperparameter optimisation to improve the log-likelihood, potentially making use of Bayesian optimisation.
+
+The question remains of which of our two methods should be used for any particular problem?
+As the results show, the SGD method is faster than the online EM method, at a cost of having a slightly worse log-likelihood.
+The online EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
+(This issue also exists with the implementation of the existing method).
+The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
+In practice, we would suggest using the online EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.
+
+\todo{Add an example plot of online-EM going wrong if space?}
 

### 2019-08-31 deconv/gmm/deconv_gmm.py
Broadcast log_p evaluation in E-step.

@@ -37,9 +37,6 @@ class DeconvGMM(BaseGMM):
 
         X, noise_covars = data
 
-        n = X.shape[0]
-        log_resps = torch.empty(n, self.k, device=self.device)
-
         T = self.covars[None, :, :, :] + noise_covars[:, None, :, :]
         try:
             T_chol = torch.cholesky(T)
@@ -49,12 +46,11 @@ class DeconvGMM(BaseGMM):
             torch.eye(self.d, device=self.device), T_chol
         )
 
-        for j in range(self.k):
-            log_resps[:, j] = mvn(
-                loc=self.means[None, j, :],
-                scale_tril=T_chol[:, j, :, :]
-            ).log_prob(X)
-            log_resps[:, j] += torch.log(self.weights[j])
+        log_resps = mvn(
+            loc=self.means[None, :, :],
+            scale_tril=T_chol
+        ).log_prob(X[:, None, :])
+        log_resps += torch.log(self.weights[None, :, 0])
 
         diff = X[:, None, :] - self.means
 

### 2019-09-01 deconv/experiments/check_online_deconv_gmm.py
Update experiment files.

@@ -41,7 +41,8 @@ def check_online_deconv_gmm(D, K, N, plot=False, device=None, verbose=False):
         batch_size=500,
         step_size=1e-1,
         restarts=1,
-        w=1e-3
+        epochs=200,
+        w=1
     )
     gmm.fit(train_data, val_data=test_data, verbose=verbose)
 

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_em.py

@@ -3,12 +3,12 @@ import argparse
 import numpy as np
 import torch
 
-from deconv.gmm.plotting import plot_covariance
 from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
 from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
 
 
-def fit_gaia_lim_em(datafile, K, use_cuda):
+def fit_gaia_lim_em(datafile, K, batch_size, epochs, step_size, w_reg,
+                    k_means_iters, use_cuda):
     data = np.load(datafile)
 
     if use_cuda:
@@ -30,9 +30,12 @@ def fit_gaia_lim_em(datafile, K, use_cuda):
         K,
         7,
         device=device,
-        batch_size=1000,
+        batch_size=batch_size,
         restarts=1,
-        step_size=1e-1
+        w=w_reg,
+        step_size=step_size,
+        epochs=epochs,
+        k_means_iters=k_means_iters
     )
     gmm.fit(train_data, val_data=val_data, verbose=True)
     train_score = gmm.score_batch(train_data)
@@ -43,13 +46,27 @@ def fit_gaia_lim_em(datafile, K, use_cuda):
 
 
 if __name__ == '__main__':
-    
+
     parser = argparse.ArgumentParser()
 
     parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-b', '--batch-size', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-s', '--step-size', type=float)
+    parser.add_argument('-w', '--w_reg', type=float)
+    parser.add_argument('-k', '--k-means-iters', type=int)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
 
     args = parser.parse_args()
 
-    fit_gaia_lim_em(args.datafile, args.components, args.use_cuda)
+    fit_gaia_lim_em(
+        args.datafile,
+        args.components,
+        args.batch_size,
+        args.epochs,
+        args.step_size,
+        args.w_reg,
+        args.k_means_iters,
+        args.use_cuda
+    )

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_sgd.py

@@ -1,4 +1,6 @@
 import argparse
+import pickle
+import time
 
 import numpy as np
 import torch
@@ -6,7 +8,8 @@ import torch
 from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM, SGDDeconvDataset
 
 
-def fit_gaia_lim_sgd(datafile, K, use_cuda):
+def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
+                     w_reg, k_means_iters, use_cuda):
     data = np.load(datafile)
 
     if use_cuda:
@@ -28,17 +31,33 @@ def fit_gaia_lim_sgd(datafile, K, use_cuda):
         K,
         7,
         device=device,
-        batch_size=500,
-        epochs=1000,
+        batch_size=batch_size,
+        epochs=epochs,
+        w=w_reg,
         restarts=1,
-        lr=1e-1
+        k_means_iters=k_means_iters,
+        lr=lr
     )
+    start_time = time.time()
     gmm.fit(train_data, val_data=val_data, verbose=True)
+    end_time = time.time()
+
     train_score = gmm.score_batch(train_data)
     val_score = gmm.score_batch(val_data)
 
     print('Training score: {}'.format(train_score))
     print('Val score: {}'.format(val_score))
+
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+        'train_curve': gmm.train_loss_curve,
+        'val_curve': gmm.val_loss_curve
+    }
+    pickle.dump(results, file=open(str(output_prefix) + '_results.pkl', mode='wb'))
+    torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
 
 
 if __name__ == '__main__':
@@ -46,9 +65,23 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser()
 
     parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-b', '--batch-size', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-l', '--learning-rate', type=float)
+    parser.add_argument('-w', '--w_reg', type=float)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
+    parser.add_argument('output_prefix')
 
     args = parser.parse_args()
 
[…]

### 2019-09-01 deconv/experiments/gaia/preproccesing.py

@@ -1,3 +1,4 @@
+import argparse
 
 import numpy as np
 import pandas as pd
@@ -14,7 +15,7 @@ columns = [
     'pmra',
     'pmdec',
     'bp_rp',
-    'phot_g_mean_flux'
+    'phot_g_mean_mag'
 ]
 
 error_columns = [
@@ -23,7 +24,6 @@ error_columns = [
     'parallax_error',
     'pmra_error',
     'pmdec_error',
-    'phot_g_mean_flux_error'
 ]
 
 corr_map = {
@@ -44,22 +44,22 @@ def get_covar(row):
     return np.diag(row[error_columns].fillna(1e12).to_numpy(dtype=np.float32))
 
 
-def vot_to_pandas(vot_file, output):
+def vot_to_pandas(vot_file):
 
     tb = Table.read(vot_file)
 
-    df = tb[columns + error_columns + corr_map.keys()].to_pandas()
-    df.to_hdf(output, 'table')
+    df = tb[columns + error_columns + list(corr_map.keys())].to_pandas()
+    return df
 
 
-def pandas_to_numpy(pandas_hdf5_file, output_file):
-    df = pd.read_hdf(pandas_hdf5_file, 'table')
+def pandas_to_numpy(df, output_file):
+    df.insert(12, column='phot_g_mean_mag_error', value=0.01)
     df.insert(12, column='bp_rp_error', value=0.01)
+    error_columns.insert(5, 'phot_g_mean_mag_error')
     error_columns.insert(5, 'bp_rp_error')
-    print(error_columns)
 
     X = df[columns].fillna(0.0).to_numpy(dtype=np.float32)
-    C = np.zeros((len(df), 7, 7))
+    C = np.zeros((len(df), 7, 7), dtype=np.float32)
     diag = np.arange(7)
     C[:, diag, diag] = df[error_columns].fillna(1e6).to_numpy(
         dtype=np.float32
@@ -88,3 +88,14 @@ def pandas_to_numpy(pandas_hdf5_file, output_file):
         X_test=X_test,
         C_test=C_test
     )
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+    parser.add_argument('input_file')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    df = vot_to_pandas(args.input_file)
+    pandas_to_numpy(df, args.output_file)

### 2019-09-01 deconv/gmm/sgd_deconv_gmm.py
Add k-means options to sgd gmm.

@@ -44,10 +44,12 @@ class SGDDeconvDataset(data_utils.Dataset):
 class SGDDeconvGMM(BaseSGDGMM):
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
-                 batch_size=64, tol=1e-6, w=1e-3, restarts=5, device=None):
+                 batch_size=64, tol=1e-6, w=1e-3, restarts=5,
+                 k_means_factor=100, k_means_iters=10, device=None):
         self.module = SGDDeconvGMMModule(components, dimensions, w, device)
         super().__init__(
             components, dimensions, epochs=epochs, lr=lr,
             batch_size=batch_size, w=w, tol=tol, restarts=restarts,
+            k_means_factor=k_means_factor, k_means_iters=k_means_iters,
             device=device
         )

### 2019-09-01 deconv/experiments/check_online_deconv_gmm.py
Update experiments.

@@ -42,6 +42,7 @@ def check_online_deconv_gmm(D, K, N, plot=False, device=None, verbose=False):
         step_size=1e-1,
         restarts=1,
         epochs=200,
+        k_means_iters=20,
         w=1
     )
     gmm.fit(train_data, val_data=test_data, verbose=verbose)

### 2019-09-01 deconv/experiments/gaia/fit_em.sh

@@ -1,0 +1,13 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --gres=gpu:1
+#SBATCH --mem=14000  # memory in Mb
+#SBATCH --time=0-05:00:00
+#SBATCH -o output/em-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/em_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 128 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --use-cuda data/gaia_sample_mag.npz results/em_${SLURM_JOBID}

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_em.py

@@ -1,4 +1,6 @@
 import argparse
+import json
+import time
 
 import numpy as np
 import torch
@@ -7,7 +9,7 @@ from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
 from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
 
 
-def fit_gaia_lim_em(datafile, K, batch_size, epochs, step_size, w_reg,
+def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w_reg,
                     k_means_iters, use_cuda):
     data = np.load(datafile)
 
@@ -37,13 +39,31 @@ def fit_gaia_lim_em(datafile, K, batch_size, epochs, step_size, w_reg,
         epochs=epochs,
         k_means_iters=k_means_iters
     )
+
+    start_time = time.time()
     gmm.fit(train_data, val_data=val_data, verbose=True)
+    end_time = time.time()
+
     train_score = gmm.score_batch(train_data)
     val_score = gmm.score_batch(val_data)
 
     print('Training score: {}'.format(train_score))
     print('Val score: {}'.format(val_score))
 
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+        'train_curve': gmm.train_ll_curve,
+        'val_curve': gmm.val_ll_curve
+    }
+
+    json.dump(results, file=open(str(output_prefix) + '_results.json', mode='w'))
+    torch.save(
+        (gmm.weights, gmm.means, gmm.covars),
+        output_prefix + '_params.pkl'
+    )
 
 if __name__ == '__main__':
 
@@ -57,11 +77,14 @@ if __name__ == '__main__':
     parser.add_argument('-k', '--k-means-iters', type=int)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
+    parser.add_argument('output_prefix')
+
 
     args = parser.parse_args()
 
     fit_gaia_lim_em(
         args.datafile,
+        args.output_prefix,
         args.components,
         args.batch_size,
         args.epochs,

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_sgd.py

@@ -1,5 +1,5 @@
 import argparse
-import pickle
+import json
 import time
 
 import numpy as np
@@ -56,12 +56,11 @@ def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
         'train_curve': gmm.train_loss_curve,
         'val_curve': gmm.val_loss_curve
     }
-    pickle.dump(results, file=open(str(output_prefix) + '_results.pkl', mode='wb'))
+    json.dump(results, file=open(str(output_prefix) + '_results.json', mode='w'))
     torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
 
 
 if __name__ == '__main__':
-    
     parser = argparse.ArgumentParser()
 
     parser.add_argument('-c', '--components', type=int)
@@ -69,6 +68,7 @@ if __name__ == '__main__':
     parser.add_argument('-e', '--epochs', type=int)
     parser.add_argument('-l', '--learning-rate', type=float)
     parser.add_argument('-w', '--w_reg', type=float)
+    parser.add_argument('-k', '--k-means-iters', type=int)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
     parser.add_argument('output_prefix')
@@ -83,5 +83,6 @@ if __name__ == '__main__':
         args.epochs,
         args.learning_rate,
         args.w_reg,
+        args.k_means_iters,
         args.use_cuda
     )

### 2019-09-01 deconv/experiments/gaia/fit_baseline.sh
Update experiments.

@@ -1,0 +1,15 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --mem=14000  # memory in Mb
+#SBATCH --time=0-05:00:00
+#SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/baseline_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+
+export PYTHONPATH=${PYTHONPATH}:${HOME}/original_xd/py/
+
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 128 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_${SLURM_JOBID}

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_em.py

@@ -59,7 +59,7 @@ def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w
         'val_curve': gmm.val_ll_curve
     }
 
-    json.dump(results, file=open(str(output_prefix) + '_results.json', mode='w'))
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
     torch.save(
         (gmm.weights, gmm.means, gmm.covars),
         output_prefix + '_params.pkl'

### 2019-09-01 deconv/experiments/gaia/fit_gaia_lim_sgd.py

@@ -56,7 +56,7 @@ def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
         'train_curve': gmm.train_loss_curve,
         'val_curve': gmm.val_loss_curve
     }
-    json.dump(results, file=open(str(output_prefix) + '_results.json', mode='w'))
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
     torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
 
 

### 2019-09-01 deconv/gmm/online_deconv_gmm.py
Use recentered variance calcs.

@@ -35,8 +35,6 @@ class OnlineDeconvGMM(DeconvGMM):
 
         self.sum_cond_means = self.means * self.sum_resps
 
-        self.sum_dev_ps = self.covars * self.sum_resps[:, :, None]
-
     def fit(self, data, val_data=None, verbose=False, interval=1):
         loader = data_utils.DataLoader(
             data,
@@ -156,32 +154,61 @@ class OnlineDeconvGMM(DeconvGMM):
         if val_data:
             self.val_ll_curve = best_val_ll_curve
 
+    def _adjust(self, covar, scale, b, c):
+        result = scale[:, :, None] * covar
+        scale_sqrt = torch.sqrt(scale)
+
+        diffs = (scale_sqrt * b - c)
+        sums = (scale_sqrt * b + c)
+
+        result += 0.5 * diffs[:, :, None] * sums[:, None, :]
+        result += 0.5 * sums[:, :, None] * diffs[:, None, :]
+
+        return result
+
     def _m_step(self, expectations, n, step_size):
         log_resps, cond_means, cond_covars = expectations
         resps = torch.exp(log_resps)[:, :, None]    # n, j, 1
 
+        resps += 10 * torch.finfo(resps.dtype).eps
+
         sum_resps = resps.sum(dim=0)
 
+        sum_resps += 10 * torch.finfo(sum_resps.dtype).eps
+
         sum_cond_means = (resps * cond_means).sum(dim=0)
+        batch_means = sum_cond_means / sum_resps
 
-        diffs = cond_means - self.means
-
+        diffs = (cond_means - batch_means)
         outer_p = diffs[:, :, :, None] * diffs[:, :, None, :]
         outer_p += cond_covars
 
-        sum_dev_ps = torch.sum(
+        batch_covars = torch.sum(
             resps[:, :, :, None] * outer_p,
             dim=0
-        )
+        ) / sum_resps[:, :, None]
+
+        m_old = self.means.clone()
+        sum_resps_old = self.sum_resps.clone()
 
         self.sum_resps = (1 - step_size) * self.sum_resps + step_size * sum_resps
-        self.sum_cond_means = (1 - step_size) * self.sum_cond_means + step_size * sum_cond_means
 
+        self.sum_cond_means = (1 - step_size) * self.sum_cond_means + step_size * sum_cond_means
         self.means = self.sum_cond_means / self.sum_resps
 
-        self.sum_dev_ps = (1 - step_size) * self.sum_dev_ps + step_size * sum_dev_ps
[…]

### 2019-09-01 ml4ps_paper/paper.tex
Update paper.

@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
+\usepackage[preprint]{neurips_2019}
 
@@
 \newcommand{\bm}{\mathbf{m}}
+\newcommand{\bb}{\mathbf{b}}
+
 \newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
@@
 
-The Extreme Deconvolution method is a mixture-model based approach originally intended to perform density estimation on noisy astronomy datasets with known noise covariances for each datapoint.
-Newer much larger datasets are available to which the existing method for fitting the model cannot scale to.
-We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via SGD, both which can make use of GPU computation.
+The Extreme Deconvolution method fits a probability density to a dataset where every observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomy datasets.
+The existing method of fitting the model cannot scale to new larger astronomy datasets.
+We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via minibatch gradient descent, both of which can make use of GPU computation.
 We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
@@
 
-Many scientific applications require an estimate of some density $p(\bx)$ to be fitted using a set of samples $\bx_i$.
-In many cases the dataset is noisy, but we wish to fit our density estimate to the true underlying noiseless data.
-Extreme Deconvolution is a method that allows Gaussian Mixture Models (GMMs) to be fitted to data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform density estimation on the dataset produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
+Extreme deconvolution is a method that fits Gaussian Mixture Models (GMM)s to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
+The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
 The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
@@
 The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-The existing method of fitting extreme deconvolution models cannot cope with such a dataset, as it requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.
-In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to such datasets, based on an online variation of the Expectation-Maximisation (EM) algorithm, and stochastic gradient descent (SGD) on the log-likelihood.
+Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
+It requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.
+
+In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to the full dataset size, based on an online variation of the Expectation-Maximisation (EM) algorithm, and minibatch gradient descent on the log-likelihood.
 We develop implementations that can leverage GPU-based computation, and show that they provide comparable density estimates to the existing method, whilst being much faster to train.
@@
 
-The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a true value $\bv_i$,
+The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
 \begin{equation}
-  \mathbf{x_i} = \bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
+  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
 \end{equation}
@@
 \begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bv_i \mid \mathbf{m}_j, T_{ij}), \quad T_{ij} = V_j + S_i.
+\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
 \end{equation}
-The original model also includes a method for handling the case where $\bx_i$ has been transformed from $\bv_i$ via a per-datapoint projection matrix $R_i$, which we omit here for simplicity.
+
 Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
+For simplicity in the remainder of this work we will set $R_i$ to be the identity matrix and handle missing data via the covariance method.
 
@@
 
-\subsection{Online Expectation-Maximisation}
-The original method for fitting extreme deconvolution models used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-In the E-step, the expected sufficient statistics for each datapoint that need to be computed are $s_{ij} = \{q_{ij}$, $q_{ij}\bv_i$, and $q_{ij}\bv_i\bv_i^T\}$.
-The M-step comprises normalisation of the sums of expected sufficient statistics to get the mixture component parameters.
+\subsection{Minibatch Expectation-Maximisation}
 
-\citet{cappeOnlineExpectationMaximization2009} proposed an online variation of EM for latent data models.
-Their method involves computing a stochastic approximation to the sums of expected sufficient statistics.
-At each iteration $t$, the sums of expected sufficient statistics $s_t = \sum_i s_{ij}$ are computed over a minibatch using the current estimate of the parameters, and the stochastic estimate $\hat{s}_t$ updated as
+The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
+Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
+At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size N, using our current estimate of the parameters,
 \begin{equation}
- \hat{s}_{t} = (1 - \lambda)\hat{s}_{t-1} + \lambda s_t
+r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid \bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid \bm_k, T_{ik})}, \quad
+\bb_{ij} = m_j + V_jT_{ij}^{-1}(\bx_i - \bm_j), \quad
+B_{ij} = V_j - V_j T_{ij}^{-1}V_j.
 \end{equation}
-where $\lambda$ is a sufficiently small step-size.
-$\lambda$ should theoretically decrease at each iteration according to the Robbins-Monro conditions, but in practice we found that using a constant step-size was sufficient.
-After each update, the parameters of the model are updated by normalising $\hat{s}_t$ appropriately.
+The $r_{ij}$ term is just the posterior probability of datapoint $\bx_i$ coming from component $j$.
+The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
+The expected sufficient statistics are then summed together over the minibatch,
+\begin{equation}
+\phi_j^1 = \sum_i r_{ij}, \quad
+\phi_j^2 = \sum_i r_{ij} \bb_{ij}, \quad
+\phi_j^3 = \sum_i r_{ij} [\bb_{ij}\bb_{ij}^T + B_{ij}].
+\end{equation}
+Stochastic estimates $\hat{\phi}_{tj} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
+\begin{equation}
+\hat{\phi}_{tj} = (1 - \lambda)\hat{\phi}_{(t-1)j} + \lambda \phi_{tj}.
+\end{equation}
+Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters.
+\begin{equation}
+\alpha_j = \frac{\hat{\phi}_j^1}{N}, \quad
+\bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
+V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
+\end{equation}
+This procedure is repeated with new randomly ordered minibatches until convergence.
+If we set $\lambda=1$ and replace each minibatch with the entire dataset then we recover the original fitting method.
 
-\todo{More complete explanation if space.}
-
-A direct implementation suggests that we should update $V_j$ as
-\begin{equation}
-V_j  = \mathrm{E}(q_{ij} \bv_ij \bv_ij^T) - \bm_j \bm_j^T
-\end{equation}
-This is usually acceptable when using double-precision floats, but using GPU based computation restricts us to single precision.
-Performing the update of $V_j$ like this with single-precision floats will result in catastrophic cancellation if the means are sufficiently large compared to the variance.
-Instead, we compute smoothed versions of the sum of squared deviation products which avoids this issue.
-\begin{equation}
-\end{equation}
-
-\todo{Better justification of why this works, or alternate method.}
+\todo{Discuss numerical stability of the $V_j$ update.}
 
@@
 Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$,
+The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ as a lower triangular matrix $L_j$ via the Cholesky decomposition, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of the unconstrained element $\tilde{L}_q$,
 \begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_k^K e^{z_k}}.
+\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
+V_j = L_jL_j^T, \quad
+(L_j)_{qq} = \exp({\tilde{L}_q}).
 \end{equation}
-We parametrise the covariances $V_j$ via the Cholesky decomposition, 
-\begin{equation}
-  V_j = L_jL_j^T
-\end{equation}
-where $L_j$ is a lower-triangular matrix.
-We also constrain the diagonal elements of $L_j$ to be positive by taking the log of them.
 Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
@@
 Having removed the constraints, we can then take the gradient of the log-likelihood with respect to the parameters and optimise it using a standard minibatch-based optimiser.
-Pre-multiplying the gradient with the inverse Fisher information matrix may be necessary to keep the covariances stable during optimisation with standard stochastic gradient descent.
-In practice we found that using the Adam optimisation algorithm eliminates the need to do this~\cite{kingmaAdamMethodStochastic2014}.
 
@@
 We implemented both of our methods in PyTorch.
-The use of PyTorch allows us to take advantage of its automatic differentiation framework and in-built optimisers for the SGD method,
-as well as allowing us to write code targeting both CPU and GPU-based computation.
-
 To evaluate our methods, we used a subset of rows from the Gaia DR2 source table \cite{brownGaiaDataRelease2018}.
@@
 Early stopping was done based on a held-out validation set comprising 10\% of the rows.
-Final model performance was evaluated on another held-out test set comprising another 10\% of the rows.
+Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection performed based on this value.
 Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
+
+\todo{Add a 2D projection of an example solution?}
 
@@
     Existing EM & -  & -     \\
-    Online EM     & - & -      \\
-    SGD     & $-5.468 \times 10^6 \pm 2.46 \times 10^3$        & - & $1469 \pm 11$ \\
+    Minibatch EM     & - & -      \\
+    SGD     & $-5.468 (\pm 0.0025) \times 10^6$        & - & $1469 \pm 11$ \\
     \bottomrule
@@
 The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
-It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and online EM methods.
+It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
 Our approaches also add more free parameters to be selected, including learning rate and batch size.
@@
 The question remains of which of our two methods should be used for any particular problem?
-As the results show, the SGD method is faster than the online EM method, at a cost of having a slightly worse log-likelihood.
-The online EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
+As the results show, the SGD method is faster than the minibatch EM method, at a cost of having a slightly worse log-likelihood.
+The minibatch EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
 (This issue also exists with the implementation of the existing method).
 The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
-In practice, we would suggest using the online EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.
+In practice, we would suggest using the minibatch EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.
 
-\todo{Add an example plot of online-EM going wrong if space?}
+\todo{Add discussion of complexity}
+\todo{Add discussion of projections $R_{ij}$}
 

### 2019-09-02 deconv/experiments/gaia/fit_baseline.sh
Use multiple CPU cores in experiments.

@@ -2,6 +2,7 @@
 #SBATCH -N 1      # nodes requested
 #SBATCH -n 1      # tasks requested
 #SBATCH --mem=14000  # memory in Mb
+#SBATCH --cpus-per-task=8
 #SBATCH --time=0-05:00:00
 #SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
 #SBATCH -e output/baseline_error-%A_%a.txt  # send stderr to errfile

### 2019-09-02 deconv/experiments/gaia/fit_em.sh

@@ -3,6 +3,7 @@
 #SBATCH -n 1      # tasks requested
 #SBATCH --gres=gpu:1
 #SBATCH --mem=14000  # memory in Mb
+#SBATCH --cpus-per-task=8
 #SBATCH --time=0-05:00:00
 #SBATCH -o output/em-%A_%a.txt  # send stdout to outfile
 #SBATCH -e output/em_error-%A_%a.txt  # send stderr to errfile

### 2019-09-02 deconv/experiments/gaia/fit_sgd.sh

@@ -3,6 +3,7 @@
 #SBATCH -n 1      # tasks requested
 #SBATCH --gres=gpu:1
 #SBATCH --mem=14000  # memory in Mb
+#SBATCH --cpus-per-task=8
 #SBATCH --time=0-05:00:00
 #SBATCH -o output/sgd-%A_%a.txt  # send stdout to outfile
 #SBATCH -e output/sgd_error-%A_%a.txt  # send stderr to errfile

### 2019-09-03 deconv/experiments/gaia/fit_gaia_lim_sgd.py
Add LR scheduling to SGD.

@@ -9,7 +9,8 @@ from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM, SGDDeconvDataset
 
 
 def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
-                     w_reg, k_means_iters, use_cuda):
+                     w_reg, k_means_iters, lr_step, lr_gamma,
+                     use_cuda):
     data = np.load(datafile)
 
     if use_cuda:
@@ -36,7 +37,9 @@ def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
         w=w_reg,
         restarts=1,
         k_means_iters=k_means_iters,
-        lr=lr
+        lr=lr,
+        lr_step=lr_step,
+        lr_gamma=lr_gamma
     )
     start_time = time.time()
     gmm.fit(train_data, val_data=val_data, verbose=True)
@@ -69,6 +72,8 @@ if __name__ == '__main__':
     parser.add_argument('-l', '--learning-rate', type=float)
     parser.add_argument('-w', '--w_reg', type=float)
     parser.add_argument('-k', '--k-means-iters', type=int)
+    parser.add_argument('--lr-step', type=int)
+    parser.add_argument('--lr-gamma', type=float)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
     parser.add_argument('output_prefix')
@@ -84,5 +89,7 @@ if __name__ == '__main__':
         args.learning_rate,
         args.w_reg,
         args.k_means_iters,
+        args.lr_step,
+        args.lr_gamma,
         args.use_cuda
     )

### 2019-09-03 deconv/gmm/sgd_gmm.py

@@ -66,7 +66,8 @@ class BaseSGDGMM(ABC):
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
                  batch_size=64, tol=1e-6, restarts=5, max_no_improvement=20,
-                 k_means_factor=100, w=1e-6, k_means_iters=10, device=None):
+                 k_means_factor=100, w=1e-6, k_means_iters=10, lr_step=5,
+                 lr_gamma=0.1, device=None):
         self.k = components
         self.d = dimensions
         self.epochs = epochs
@@ -89,6 +90,11 @@ class BaseSGDGMM(ABC):
         self.optimiser = torch.optim.Adam(
             params=self.module.parameters(),
             lr=self.lr
+        )
+        self.scheduler = torch.optim.lr_scheduler.StepLR(
+            self.optimiser,
+            step_size=lr_step,
+            gamma=lr_gamma
         )
 
     @property
@@ -152,6 +158,8 @@ class BaseSGDGMM(ABC):
                 if val_data:
                     val_loss = self.score_batch(val_data)
                     val_loss_curve.append(val_loss)
+
+                self.scheduler.step()
 
                 if verbose and i % interval == 0:
                     if val_data:

### 2019-09-03 deconv/experiments/gaia/fit_em.sh
Try with 256 components.

@@ -2,13 +2,13 @@
 #SBATCH -N 1      # nodes requested
 #SBATCH -n 1      # tasks requested
 #SBATCH --gres=gpu:1
-#SBATCH --mem=14000  # memory in Mb
+#SBATCH --mem=8000  # memory in Mb
 #SBATCH --cpus-per-task=8
-#SBATCH --time=0-05:00:00
+#SBATCH --time=0-10:00:00
 #SBATCH -o output/em-%A_%a.txt  # send stdout to outfile
 #SBATCH -e output/em_error-%A_%a.txt  # send stderr to errfile
 #SBATCH --partition=apollo
 
 source ~/.bashrc
 conda activate deconv
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 128 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --use-cuda data/gaia_sample_mag.npz results/em_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 256 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --use-cuda data/gaia_sample_mag.npz results/em_256_${SLURM_JOBID}

### 2019-09-03 deconv/experiments/gaia/fit_gaia_baseline.py

@@ -42,6 +42,7 @@ def fit_gaia_baseline(datafile, output_prefix, K, epochs, w_reg,
         means,
         covars,
         w=w_reg,
+        logfile=str(output_prefix) + '_log',
         maxiter=epochs
     )
 

### 2019-09-03 ml4ps_paper/paper.tex
Updated results.

@@
 We implemented both of our methods in PyTorch.
-To evaluate our methods, we used a subset of rows from the Gaia DR2 source table \cite{brownGaiaDataRelease2018}.
+To evaluate our methods, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
 We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
-In total there were approximately 2 million rows.
+In total there were 2 million rows.
 Where data were missing, we set the field to zero and the noise variance to a large value.
 This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used the same number of components $K=128$ for each model, following \cite{andersonImprovingGaiaParallax2018}.
+We used the same number of components $K=128$ for each model, following \citet{andersonImprovingGaiaParallax2018}.
 In practice we would want to select a value of $K$ by cross-validation.
@@
 Initial covariances were set to the identity matrix.
+Training was done for a maximum of 20 epochs.
 Early stopping was done based on a held-out validation set comprising 10\% of the rows.
@@
     \toprule
-    Method     & Validation     & Test & Training Time (s) \\
+    Method     & Validation     & Test & Training Time (minutes) \\
     \midrule
-    Existing EM & -  & -     \\
-    Minibatch EM     & - & -      \\
-    SGD     & $-5.468 (\pm 0.0025) \times 10^6$        & - & $1469 \pm 11$ \\
+    Existing EM & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
+    Minibatch EM & $-5.183 (\pm 0.0005) \times 10^6$    & - & $48.7 \pm 0.39$      \\
+    SGD     & $-5.193 (\pm 0.0029) \times 10^6$        & - & $36.0 \pm 0.34$ \\
     \bottomrule

### 2019-09-03 ml4ps_paper/paper.tex
Add R to EM equations.

@@
 Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
-It requires all of the data to fit in memory, and must make a full pass over the dataset before it can update parameters.
+It must make a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
 
@@
 Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-For simplicity in the remainder of this work we will set $R_i$ to be the identity matrix and handle missing data via the covariance method.
 
@@
 \begin{equation}
-r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid \bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid \bm_k, T_{ik})}, \quad
-\bb_{ij} = m_j + V_jT_{ij}^{-1}(\bx_i - \bm_j), \quad
-B_{ij} = V_j - V_j T_{ij}^{-1}V_j.
+r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
+\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
+B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
 \end{equation}
@@
 This procedure is repeated with new randomly ordered minibatches until convergence.
-If we set $\lambda=1$ and replace each minibatch with the entire dataset then we recover the original fitting method.
-
-\todo{Discuss numerical stability of the $V_j$ update.}
+If we set $\lambda=1$, replace each minibatch with the entire dataset and factorise the update for $V_j$ then we recover the original fitting method.
+The update for $V_j$ could result in catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
+In Appendix~\ref{apx:variance-rewrite} we show how this update can be rewritten in a symbolically equivalent but more numerically stable form.
 
@@
 
-We implemented both of our methods in PyTorch.
-To evaluate our methods, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
-We selected the 5 primary astrometric features, along with the BP-RP colour and mean flux in the G-band.
+We implemented both of our methods in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
+To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
+We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
 In total there were 2 million rows.
 Where data were missing, we set the field to zero and the noise variance to a large value.
+We set the projection $R_i$ to the identity matrix for every sample.
 This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used the same number of components $K=128$ for each model, following \citet{andersonImprovingGaiaParallax2018}.
+We used the same number of components $K=256$ for each model.
 In practice we would want to select a value of $K$ by cross-validation.
@@
 Initial covariances were set to the identity matrix.
-Training was done for a maximum of 20 epochs.
-Early stopping was done based on a held-out validation set comprising 10\% of the rows.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection performed based on this value.
+Training was done for 20 epochs.
+We used a validation set comprising $10\%$ of the rows when developing our experiments.
+Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
 Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
@@
     \midrule
-    Existing EM & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
+    Existing EM (\citet{bovyExtremeDeconvolutionInferring2011}) & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
     Minibatch EM & $-5.183 (\pm 0.0005) \times 10^6$    & - & $48.7 \pm 0.39$      \\
@@
 
+\appendix
+
+\section{Stable Variance Update}
+\label{apx:variance-rewrite}
+
+\section{Experiment Details}
+
 \end{document}

### 2019-09-03 ml4ps_paper/paper.tex
Add plot of RA vs Dec.

@@
 \usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}
-\usepackage{color}      % microtypography
+\usepackage{microtype}      % microtypography
+\usepackage{color}
+\usepackage{graphicx}
 
@@
 
-\todo{Add a 2D projection of an example solution?}
-
 \begin{table}
-  \caption{Total validation and test log-likelihoods along with average training time for the Gaia data subset. Average over 5 runs with standard deviation.}
+  \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.
+Average over 5 runs with standard deviation.}
   \label{results-table}
   \centering
-  \begin{tabular}{lccc}
+  \begin{tabular}{lcccc}
     \toprule
-    Method     & Validation     & Test & Training Time (minutes) \\
+    Method     & K &  Validation     & Test & Time (minutes) \\
     \midrule
-    Existing EM (\citet{bovyExtremeDeconvolutionInferring2011}) & $-5.191 (\pm 0.0080) \times 10^6$  & - & $101.1 \pm 2.19$    \\
-    Minibatch EM & $-5.183 (\pm 0.0005) \times 10^6$    & - & $48.7 \pm 0.39$      \\
-    SGD     & $-5.193 (\pm 0.0029) \times 10^6$        & - & $36.0 \pm 0.34$ \\
+    Existing EM  & 128 & $-25.95 \pm 0.04$  & - & $101.1 \pm 2.19$    \\
+    (\citet{bovyExtremeDeconvolutionInferring2011}) & 256 & $-25.76 \pm 0.02$ & - & $195.3 \pm 4.08$ \\
+    \midrule
+    Minibatch EM & 128 & $-25.92 \pm 0.01$ & - & $44.5 \pm 1.0$      \\
+    & 256 & $-25.83 \pm 0.003$ & - & $51.6 \pm 3.0$ \\
+    \midrule
+    SGD  & 128  & $-25.94 \pm 0.01$ & - & $34.0 \pm 0.83$ \\
+    & 256 &$-25.84 \pm 0.01$ & - & $36.6 \pm 1.3$ \\
     \bottomrule
@@
 \end{table}
+
+\begin{figure}
+  \centering
+  \includegraphics{density.pdf}
+  \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
+  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way.}
+\end{figure}
 

### 2019-09-03 ml4ps_paper/figures/ra_dec_projection.py

@@ -1,0 +1,46 @@
+import numpy as np
+import torch
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+import corner
+
+sns.set()
+sns.set_context('paper')
+
+N = 100000
+
+weights, means, covars = torch.load(
+    '../results/em_256_507993_params.pkl',
+    map_location=torch.device('cpu')
+)
+
+X = torch.distributions.MultivariateNormal(
+    loc=means, covariance_matrix=covars
+).sample((N,)).numpy()
+
+j = torch.distributions.Categorical(
+    probs=weights.squeeze()
+).sample((N,)).numpy()
+
+X = X[np.arange(N), j, :]
+
+fig, ax = plt.subplots(figsize=(3, 2))
+
+figure = corner.hist2d(X[:, 0], X[:, 1], bins=100)
+
+ax.set_xlim(0, 360)
+ax.set_xticks(np.arange(0, 361, 60))
+
+ax.set_ylim(-90, 90)
+ax.set_yticks(np.arange(-90, 91, 30))
+
+ax.set_xlabel(r'Right Ascension ($\circ$)')
+ax.set_ylabel(r'Declination ($\circ$)')
+
+fig.tight_layout()
+
+fig.savefig('density.pdf')
+
+plt.show()

### 2019-09-04 deconv/experiments/checks/check_online_deconv_gmm.py
Update experiments.

@@ -43,7 +43,7 @@ def check_online_deconv_gmm(D, K, N, plot=False, device=None, verbose=False):
         restarts=1,
         epochs=200,
         k_means_iters=20,
-        w=1
+        w=1e-3
     )
     gmm.fit(train_data, val_data=test_data, verbose=verbose)
 

### 2019-09-04 deconv/experiments/gaia/fit_baseline.sh

@@ -1,7 +1,7 @@
 #!/bin/sh
 #SBATCH -N 1      # nodes requested
 #SBATCH -n 1      # tasks requested
-#SBATCH --mem=8000  # memory in Mb
+#SBATCH --mem=14000  # memory in Mb
 #SBATCH --cpus-per-task=8
 #SBATCH --time=0-10:00:00
 #SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
@@ -13,4 +13,4 @@ conda activate deconv
 
 export PYTHONPATH=${PYTHONPATH}:${HOME}/original_xd/py/
 
-python deconv/experiments/gaia/fit_gaia_baseline.py -c 256 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 64 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_64_${SLURM_JOBID}

### 2019-09-04 deconv/experiments/gaia/fit_em.sh

@@ -11,4 +11,4 @@
 
 source ~/.bashrc
 conda activate deconv
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 256 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --use-cuda data/gaia_sample_mag.npz results/em_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 64 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --use-cuda data/gaia_sample_mag.npz results/em_64_${SLURM_JOBID}

### 2019-09-04 deconv/experiments/gaia/fit_gaia_baseline.py

@@ -33,7 +33,7 @@ def fit_gaia_baseline(datafile, output_prefix, K, epochs, w_reg,
 
     weights = (counts / counts.sum()).numpy()
     means = centroids.numpy()
-    covars = np.array(128 * [np.eye(7)])
+    covars = np.array(K * [np.eye(7)])
 
     ll = extreme_deconvolution(
         data['X_train'],

### 2019-09-04 deconv/experiments/gaia/fit_sgd.sh

@@ -11,4 +11,4 @@
 
 source ~/.bashrc
 conda activate deconv
-python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 128 -b 500 -e 30 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_schedule_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 64 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_64_${SLURM_JOBID}

### 2019-09-04 ml4ps_paper/figures/learning.py
Update table and add plots.

@@ -1,0 +1,62 @@
+import os
+import numpy as np
+import torch
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+from results_table import produce_table
+
+sns.set()
+sns.set_context('paper')
+
+fig, axes = plt.subplots(1, 2, figsize=(6, 2)) 
+
+table = produce_table('../results/')
+
+axes[0].set_xlabel('Time-scaled Epoch')
+axes[0].set_ylabel('Avg. Log-likelihood')
+
+labels = ('Existing EM', 'Minibatch EM', 'SGD')
+
+result_files = os.listdir('../results/')
+baseline_curves = np.array([
+    np.loadtxt('../results/' + f) for f in result_files if f.startswith(
+        'baseline_256'
+    ) and f.endswith('loglike.log')
+])
+
+print(baseline_curves)
+
+for l, result in zip(labels, table.values()):
+    if l != 'Existing EM':
+        x = np.arange(20) * (result[256][1].mean() / 20)
+        y = result[256][2].mean(axis=0)
+        y_err = 1.96 * result[256][2].std(axis=0)
+    else:
+        x = np.arange(20) * (result[256][1].mean() / 20)
+        y = baseline_curves.mean(axis=0)
+        y_err = 1.96 * baseline_curves.std(axis=0)
+
+    axes[0].plot(x, y)
+    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5)
+
+k = np.array([64, 128, 256])
+
+axes[1].set_xlabel(r'Mixture Components $K$')
+axes[1].set_ylabel(r'Time (minutes)')
+axes[1].set_xticks(k)
+
+for l, result in zip(labels, table.values()):
+    t = np.array([[r[1].mean(), r[1].std()] for r in result.values()])
+    axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l)
+
+axes[1].legend(loc='lower left', bbox_to_anchor=(1.01, 0))
+
+axes[1].set_ylim(-1, 201)
+
+fig.tight_layout()
+
+fig.savefig('learning.pdf')
+
+plt.show()

### 2019-09-04 ml4ps_paper/figures/ra_dec_projection.py

@@ -26,7 +26,7 @@ j = torch.distributions.Categorical(
 
 X = X[np.arange(N), j, :]
 
-fig, ax = plt.subplots(figsize=(3, 2))
+fig, ax = plt.subplots(figsize=(4, 2))
 
 figure = corner.hist2d(X[:, 0], X[:, 1], bins=100)
 

### 2019-09-04 ml4ps_paper/figures/results_table.py

@@ -1,0 +1,71 @@
+import argparse
+import json
+import os
+
+import numpy as np
+
+MODELS = [
+    'baseline',
+    'em',
+    'sgd'
+]
+
+SIZES = [
+    64,
+    128,
+    256
+]
+
+def get_results(files, model, size, file_dir):
+    return [
+        json.load(open(str(file_dir) + f)) for f in files if f.startswith(
+            '{}_{}'.format(model, size)
+        ) and f.endswith('.json')
+    ]
+
+
+def produce_table(results_dir):
+
+    results_files = os.listdir(str(results_dir))
+
+    table = {}
+
+    for model in MODELS:
+        table[model] = {}
+        for size in SIZES:
+            results = get_results(results_files, model, size, results_dir)
+            scores = np.array([r['val_score'] for r in results]) / 200000
+            times = np.array(
+                [r['end_time'] - r['start_time'] for r in results]
+            ) / 60
+            if model != 'baseline':
+                curve = np.array([r['train_curve'] for r in results]) / 1600000
+            else:
+                curve = None
+
+            if model == 'sgd':
+                scores = scores * -1
+                curve = curve * -1
+
+            table[model][size] = (scores, times, curve)
+
+            print('{} {}: {:.2f} +- {:.2f}, {:.1f} +- {:.2f}'.format(
+                model,
+                size,
+                scores.mean(),
+                scores.std(),
+                times.mean(),
+                times.std()
+            ))
+
+    return table
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+    parser.add_argument('results_dir')
+    args = parser.parse_args()
+
+    print(args)
+
+    produce_table(args.results_dir)

### 2019-09-04 ml4ps_paper/paper.tex
Update table.
Update table and add plots.
Update paper.

@@
 Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
-At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size N, using our current estimate of the parameters,
+At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
 \begin{equation}
-r_{ij} = \frac{\alpha_j \mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
+r_{ij} = \frac{\alpha_j \mathcal{M}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
 \bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
@@
 \begin{equation}
-\alpha_j = \frac{\hat{\phi}_j^1}{N}, \quad
+\alpha_j = \frac{\hat{\phi}_j^1}{M}, \quad
 \bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
@@
 This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used the same number of components $K=256$ for each model.
+We used a range of mixture component sizes $K$.
 In practice we would want to select a value of $K$ by cross-validation.
@@
 Training was done for 20 epochs.
+The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU.
 We used a validation set comprising $10\%$ of the rows when developing our experiments.
 Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
+Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
+Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
+Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
 
-\begin{table}
+\begin{table}{}
   \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.
@@
     \midrule
-    Existing EM  & 128 & $-25.95 \pm 0.04$  & - & $101.1 \pm 2.19$    \\
-    (\citet{bovyExtremeDeconvolutionInferring2011}) & 256 & $-25.76 \pm 0.02$ & - & $195.3 \pm 4.08$ \\
+    Existing EM                                     & 64  & $-26.12 \pm 0.03$ & - & $60.0 \pm 7.9$\\
+    (\citet{bovyExtremeDeconvolutionInferring2011}) & 128 & $-25.95 \pm 0.04$  & - & $101.1 \pm 2.19$    \\
+                                                    & 256 & $-25.76 \pm 0.02$ & - & $195.3 \pm 4.08$ \\
     \midrule
-    Minibatch EM & 128 & $-25.92 \pm 0.01$ & - & $44.5 \pm 1.0$      \\
+    Minibatch EM  & 64  & $-26.04 \pm 0.01$   & - & $26.8 \pm 1.71$ \\
+                  & 128 & $-25.92 \pm 0.01$   & - & $44.5 \pm 1.0$      \\
                   & 256 & $-25.83 \pm 0.003$  & - & $51.6 \pm 3.0$ \\
     \midrule
-    SGD  & 128  & $-25.94 \pm 0.01$ & - & $34.0 \pm 0.83$ \\
-    & 256 &$-25.84 \pm 0.01$ & - & $36.6 \pm 1.3$ \\
+    SGD & 64    & $-26.10 \pm 0.02$  & - &$23.5 \pm 2.34$ \\
+        & 128   & $-25.94 \pm 0.01$  & - & $34.0 \pm 0.83$ \\
+        & 256   & $-25.84 \pm 0.01$  & - & $36.6 \pm 1.3$ \\
     \bottomrule
@@
   \centering
+  \includegraphics{learning.pdf}
+  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
+Epochs rescaled by average training time.
+Error bars not visible.
+\textbf{Right}: Training time as a function of mixture components $K$.
+Error bars indicate $\pm$ 2 standard deviations.}
+  \label{fig:training}
+\end{figure}
+
+\begin{figure}
+  \centering
   \includegraphics{density.pdf}
   \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
-  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way.}
+  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
+  \label{fig:projection}
 \end{figure}
-
-\todo{Add plot of log-likelihood results against dataset size.}
 
@@
 Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst being able to work with datasets that cannot fit in memory.
-The results also show that using GPU-based computation speeds up training considerably.
+The results also show that using GPU-based computation speeds up training considerably as mixture component size $K$ increases.
 
@@
 
-The question remains of which of our two methods should be used for any particular problem?
-As the results show, the SGD method is faster than the minibatch EM method, at a cost of having a slightly worse log-likelihood.
-The minibatch EM method performed better, but during training would occasionally terminate early as a result of an update forming an invalid covariance matrix, necessitating a restart.
-(This issue also exists with the implementation of the existing method).
+The question remains of which method should be used for any particular problem?
+The existing EM method had comparable log-likelihoods for smaller sizes of $K$, and did marginally better for the largest size, but at the cost of a much longer training time.
+The SGD method is faster than the minibatch EM method, especially as $K$ increases, at a cost of having a slightly worse log-likelihood.
 The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
-In practice, we would suggest using the minibatch EM if the size of the dataset was small enough to permit several restarts, whilst an extremely large dataset would probably be better fitted with the SGD method.
-
-\todo{Add discussion of complexity}
-\todo{Add discussion of projections $R_{ij}$}
+In practice we would suggest using the minibatch EM method for relatively smaller datasets, and the SGD method on larger datasets.
 

### 2019-09-04 ml4ps_paper/figures/learning.py
Add figure pdfs.

@@ -26,8 +26,6 @@ baseline_curves = np.array([
     ) and f.endswith('loglike.log')
 ])
 
-print(baseline_curves)
-
 for l, result in zip(labels, table.values()):
     if l != 'Existing EM':
         x = np.arange(20) * (result[256][1].mean() / 20)
@@ -57,6 +55,6 @@ axes[1].set_ylim(-1, 201)
 
 fig.tight_layout()
 
-fig.savefig('learning.pdf')
+fig.savefig('figures/learning.pdf')
 
 plt.show()

### 2019-09-04 ml4ps_paper/figures/ra_dec_projection.py

@@ -41,6 +41,6 @@ ax.set_ylabel(r'Declination ($\circ$)')
 
 fig.tight_layout()
 
-fig.savefig('density.pdf')
+fig.savefig('figures/density.pdf')
 
 plt.show()

### 2019-09-04 ml4ps_paper/paper.tex
Update figures to right place.
Iain tweaks

@@
 
-The Extreme Deconvolution method fits a probability density to a dataset where every observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomy datasets.
-The existing method of fitting the model cannot scale to new larger astronomy datasets.
-We propose two extensions to the extreme deconvolution method based on an online variation of the EM algorithm, and direct optimisation of the log-likelihood via minibatch gradient descent, both of which can make use of GPU computation.
+The Extreme Deconvolution method fits a probability density to a dataset where every observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
+% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
+The existing fitting method is batch EM, which wouldn't normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
+We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
 We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
@@
 
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMM)s to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
+Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
 The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
@@
 The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-Previous work using an extreme deconvolution model on the Gaia catalogue had to work with a subset of the data, as the existing method of fitting the models cannot work with very large datasets \cite{andersonImprovingGaiaParallax2018}.
-It must make a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
+Previous work using an extreme deconvolution model on the Gaia catalogue %had to
+worked with a subset of the data,
+%as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
+but the intention is to fit models with the full dataset
+\cite{andersonImprovingGaiaParallax2018}.
+The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
+To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
 
-In this work, we propose two new methods for fitting the extreme deconvolution model that can scale to the full dataset size, based on an online variation of the Expectation-Maximisation (EM) algorithm, and minibatch gradient descent on the log-likelihood.
-We develop implementations that can leverage GPU-based computation, and show that they provide comparable density estimates to the existing method, whilst being much faster to train.
+In this work, we develop two minibatch methods for fitting the extreme deconvolution model
+%that can scale to the full dataset size, % -- not demonstrated here?
+based on 1) an online variation of the Expectation-Maximisation (EM) algorithm, and 2) a gradient optimizer.
+Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
 
@@
 \end{equation}
-The $r_{ij}$ term is just the posterior probability of datapoint $\bx_i$ coming from component $j$.
+The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
 The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
@@
 V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
+\label{eqn:mstep}
 \end{equation}
-This procedure is repeated with new randomly ordered minibatches until convergence.
-If we set $\lambda=1$, replace each minibatch with the entire dataset and factorise the update for $V_j$ then we recover the original fitting method.
-The update for $V_j$ could result in catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how this update can be rewritten in a symbolically equivalent but more numerically stable form.
+This procedure is repeated with new randomly-ordered minibatches until convergence.
+If we set $\lambda=1$ and replace each minibatch with the entire dataset,
+% and factorise the update for $V_j$ -- % unclear to someone new?
+then the update corresponds to the original batch fitting method.
+Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}.
+There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
+In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
+% symbolically equivalent but
+more numerically stable form.
 
@@
 An alternative to EM-based methods is to optimise the log-likelihood directly.
-The log-likelihood has some constraints on its parameters, namely that the weight $a_j$ sum to $1$ and that the covariances $V_j$ are positive definite.
+The log-likelihood has some constraints on its parameters, namely that the mixture weights $a_j$ are positive and sum to $1$ and that the covariances $V_j$ are positive definite.
 Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
@@
 \end{equation}
-Note that for a standard GMM we could bypass forming the covariance altogether and evaluate the log PDF of the Normal distribution directly with the Cholesky factor.
-For the extreme deconvolution model, this cannot be done, as we need to form the covariance $Tij$ for every datapoint.
-Having removed the constraints, we can then take the gradient of the log-likelihood with respect to the parameters and optimise it using a standard minibatch-based optimiser.
+Having removed the constraints, we optimise the likelihood using any standard minibatch gradient-based optimiser.
+
+For a standard Gaussian mixture model, gradient based optimization has a scaling advantage over EM\@.
+There is no need to form the $D\!\times\!D$ covariance matrix $V_j$, since the Gaussian density can be evaluated directly from the Cholesky factor $L_j$ in $O(D^2)$, whereas an EM update is $O(D^3)$.
+Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
 
@@
 
-We implemented both of our methods in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
+We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
 To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
@@
 We set the projection $R_i$ to the identity matrix for every sample.
-This is only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
+This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
 We used a range of mixture component sizes $K$.
@@
 
-Initialisation of all models was performed by first running a minibatch version of the K-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
+Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
 Initial covariances were set to the identity matrix.
+% TODO -- is data standardized?
 Training was done for 20 epochs.
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU.
+The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
+While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
 We used a validation set comprising $10\%$ of the rows when developing our experiments.
@@
   \centering
-  \includegraphics{learning.pdf}
+  \includegraphics{figures/learning.pdf}
   \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
@@
   \centering
-  \includegraphics{density.pdf}
+  \includegraphics{figures/density.pdf}
   \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.

### 2019-09-04 ml4ps_paper/paper.tex
Add inset figure.

@@
 \newcommand{\bb}{\mathbf{b}}
+\newcommand{\bc}{\mathbf{c}}
+\newcommand{\bd}{\mathbf{d}}
 
@@
 % The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which wouldn't normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
+The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
 We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
@@
   \centering
-  \includegraphics{figures/learning.pdf}
+  \includegraphics[width=\textwidth]{figures/learning.pdf}
   \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
@@
 
-\section{Stable Variance Update}
+\section{Stable Covariance Update}
 \label{apx:variance-rewrite}

### 2019-09-04 ml4ps_paper/figures/learning.py

@@ -1,8 +1,8 @@
 import os
 import numpy as np
-import torch
 
 import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes, mark_inset
 import seaborn as sns
 
 from results_table import produce_table
@@ -26,18 +26,33 @@ baseline_curves = np.array([
     ) and f.endswith('loglike.log')
 ])
 
+axins = inset_axes(
+    axes[0], width='40%', height='40%', loc='lower right', borderpad=2
+)
+
 for l, result in zip(labels, table.values()):
     if l != 'Existing EM':
         x = np.arange(20) * (result[256][1].mean() / 20)
         y = result[256][2].mean(axis=0)
         y_err = 1.96 * result[256][2].std(axis=0)
+
     else:
         x = np.arange(20) * (result[256][1].mean() / 20)
         y = baseline_curves.mean(axis=0)
         y_err = 1.96 * baseline_curves.std(axis=0)
 
-    axes[0].plot(x, y)
-    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5)
+    axes[0].plot(x, y, zorder=200)
+    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5, zorder=200)
+    axins.plot(x, y)
+    axins.fill_between(x, y + y_err, y - y_err, alpha=0.5)
+
+
+axins.set_ylim(-26.2, -25.7)
+axins.set_xlim(-1, 50)
+axins.set_xticks([])
+axins.set_yticks([])
+plt.setp(axins.spines.values(), color='grey')
+mark_inset(axes[0], axins, loc1=1, loc2=3, ec='grey')
 
 k = np.array([64, 128, 256])
 

### 2019-09-05 ml4ps_paper/figures/learning.py
Switch to logscale and remove inset on Figure 1.

@@ -26,9 +26,9 @@ baseline_curves = np.array([
     ) and f.endswith('loglike.log')
 ])
 
-axins = inset_axes(
-    axes[0], width='40%', height='40%', loc='lower right', borderpad=2
-)
+# axins = inset_axes(
+#     axes[0], width='40%', height='40%', loc='lower right', borderpad=2
+# )
 
 for l, result in zip(labels, table.values()):
     if l != 'Existing EM':
@@ -41,18 +41,12 @@ for l, result in zip(labels, table.values()):
         y = baseline_curves.mean(axis=0)
         y_err = 1.96 * baseline_curves.std(axis=0)
 
-    axes[0].plot(x, y, zorder=200)
-    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5, zorder=200)
-    axins.plot(x, y)
-    axins.fill_between(x, y + y_err, y - y_err, alpha=0.5)
+    axes[0].plot(x, y)
+    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5)
 
+axes[0].set_xscale('log')
+axes[0].set_ylim(-28, -25)
 
-axins.set_ylim(-26.2, -25.7)
-axins.set_xlim(-1, 50)
-axins.set_xticks([])
-axins.set_yticks([])
-plt.setp(axins.spines.values(), color='grey')
-mark_inset(axes[0], axins, loc1=1, loc2=3, ec='grey')
 
 k = np.array([64, 128, 256])
 

### 2019-09-05 deconv/experiments/gaia/fit_gaia_lim_em.py
Add step size decay option to online em.

@@ -10,7 +10,7 @@ from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
 
 
 def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w_reg,
-                    k_means_iters, use_cuda):
+                    k_means_iters, lr_step, lr_gamma, use_cuda):
     data = np.load(datafile)
 
     if use_cuda:
@@ -37,7 +37,9 @@ def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w
         w=w_reg,
         step_size=step_size,
         epochs=epochs,
-        k_means_iters=k_means_iters
+        k_means_iters=k_means_iters,
+        lr_step=lr_step,
+        lr_gamma=lr_gamma
     )
 
     start_time = time.time()
@@ -75,6 +77,8 @@ if __name__ == '__main__':
     parser.add_argument('-s', '--step-size', type=float)
     parser.add_argument('-w', '--w_reg', type=float)
     parser.add_argument('-k', '--k-means-iters', type=int)
+    parser.add_argument('--lr-step', type=int)
+    parser.add_argument('--lr-gamma', type=float)
     parser.add_argument('--use-cuda', action='store_true', help='Use GPU')
     parser.add_argument('datafile')
     parser.add_argument('output_prefix')
@@ -91,5 +95,7 @@ if __name__ == '__main__':
         args.step_size,
         args.w_reg,
         args.k_means_iters,
+        args.lr_step,
+        args.lr_gamma,
         args.use_cuda
     )

### 2019-09-05 deconv/gmm/online_deconv_gmm.py

@@ -10,7 +10,8 @@ class OnlineDeconvGMM(DeconvGMM):
     def __init__(self, components, dimensions, epochs=1000, w=1e-6,
                  tol=1e-6, step_size=0.1, batch_size=100, restarts=5,
                  max_no_improvement=20, k_means_factor=100,
-                 k_means_iters=10, device=None):
+                 k_means_iters=10, lr_step=10, lr_gamma=0.1,
+                 device=None):
         super().__init__(components, dimensions, epochs=epochs, w=w, tol=tol,
                          restarts=restarts, device=device)
         self.batch_size = batch_size
@@ -18,6 +19,8 @@ class OnlineDeconvGMM(DeconvGMM):
         self.max_no_improvement = max_no_improvement
         self.k_means_factor = k_means_factor
         self.k_means_iters = k_means_iters
+        self.lr_step = lr_step
+        self.lr_gamma = lr_gamma
 
     def _init_sum_stats(self, loader, n):
 
@@ -93,6 +96,9 @@ class OnlineDeconvGMM(DeconvGMM):
                 if val_data:
                     val_ll = self.score_batch(val_data)
                     val_ll_curve.append(val_ll)
+
+                if (i + 1) % self.lr_step == 0:
+                    self.step_size *= self.lr_gamma
 
                 if verbose and i % interval == 0:
                     if val_data:

### 2019-09-05 deconv/experiments/gaia/fit_baseline.sh
Bundle together model size fitting.

@@ -14,3 +14,6 @@ conda activate deconv
 export PYTHONPATH=${PYTHONPATH}:${HOME}/original_xd/py/
 
 python deconv/experiments/gaia/fit_gaia_baseline.py -c 64 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_64_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 128 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_128_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 256 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 512 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_512_${SLURM_JOBID}

### 2019-09-05 ml4ps_paper/figures/learning.py
Update results.

@@ -5,14 +5,14 @@ import matplotlib.pyplot as plt
 from mpl_toolkits.axes_grid1.inset_locator import inset_axes, mark_inset
 import seaborn as sns
 
-from results_table import produce_table
+from results_table import get_table
 
 sns.set()
 sns.set_context('paper')
 
 fig, axes = plt.subplots(1, 2, figsize=(6, 2)) 
 
-table = produce_table('../results/')
+table = get_table('../results/')
 
 axes[0].set_xlabel('Time-scaled Epoch')
 axes[0].set_ylabel('Avg. Log-likelihood')
@@ -26,18 +26,13 @@ baseline_curves = np.array([
     ) and f.endswith('loglike.log')
 ])
 
-# axins = inset_axes(
-#     axes[0], width='40%', height='40%', loc='lower right', borderpad=2
-# )
-
 for l, result in zip(labels, table.values()):
     if l != 'Existing EM':
-        x = np.arange(20) * (result[256][1].mean() / 20)
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
         y = result[256][2].mean(axis=0)
         y_err = 1.96 * result[256][2].std(axis=0)
-
     else:
-        x = np.arange(20) * (result[256][1].mean() / 20)
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
         y = baseline_curves.mean(axis=0)
         y_err = 1.96 * baseline_curves.std(axis=0)
 
@@ -48,7 +43,7 @@ axes[0].set_xscale('log')
 axes[0].set_ylim(-28, -25)
 
 
-k = np.array([64, 128, 256])
+k = np.array([64, 128, 256, 512])
 
 axes[1].set_xlabel(r'Mixture Components $K$')
 axes[1].set_ylabel(r'Time (minutes)')
@@ -56,11 +51,10 @@ axes[1].set_xticks(k)
 
 for l, result in zip(labels, table.values()):
     t = np.array([[r[1].mean(), r[1].std()] for r in result.values()])
+    print(t.shape)
     axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l)
 
 axes[1].legend(loc='lower left', bbox_to_anchor=(1.01, 0))
-
-axes[1].set_ylim(-1, 201)
 
 fig.tight_layout()
 

### 2019-09-06 deconv/gmm/online_deconv_gmm.py
Simplyify covar update.

@@ -202,19 +202,18 @@ class OnlineDeconvGMM(DeconvGMM):
         self.sum_cond_means = (1 - step_size) * self.sum_cond_means + step_size * sum_cond_means
         self.means = self.sum_cond_means / self.sum_resps
 
-        shrunk_covar = (1 - step_size) * self._adjust(
+        self.covars = (1 - step_size) * self._adjust(
             self.covars - self.w,
             sum_resps_old / self.sum_resps,
             m_old,
             self.means
-        )
-        new_covar = step_size * self._adjust(
+        ) + step_size * self._adjust(
             batch_covars,
             sum_resps / self.sum_resps,
             batch_means,
             self.means
-        )
-        self.covars = shrunk_covar + new_covar + self.w
+        ) + self.w
+
         self.weights = self.sum_resps / self.batch_size
 
     def score_batch(self, dataset):

### 2019-09-06 deconv/gmm/sgd_gmm.py
Add multistep schedule to SGD

@@ -91,9 +91,9 @@ class BaseSGDGMM(ABC):
             params=self.module.parameters(),
             lr=self.lr
         )
-        self.scheduler = torch.optim.lr_scheduler.StepLR(
+        self.scheduler = torch.optim.lr_scheduler.MultiStepLR(
             self.optimiser,
-            step_size=lr_step,
+            milestones=[lr_step, lr_step + 5],
             gamma=lr_gamma
         )
 

### 2019-09-08 deconv/gmm/sgd_gmm.py
Do regularisation properly.

@@ -44,7 +44,7 @@ class SGDGMMModule(nn.Module):
 
     @property
     def covars(self):
-        return torch.matmul(self.L, torch.transpose(self.L, -2, -1)) + self.w
+        return torch.matmul(self.L, torch.transpose(self.L, -2, -1))
 
     def forward(self, data):
 
@@ -105,7 +105,13 @@ class BaseSGDGMM(ABC):
     def covars(self):
         return self.module.covars.detach()
 
+    def reg_loss(self, n, n_total):
+        l = (n / n_total) * self.w / torch.diagonal(self.module.covars, dim1=-1, dim2=-2)
+        return l.sum()
+
     def fit(self, data, val_data=None, verbose=False, interval=1):
+
+        n_total = len(data)
 
         init_loader = data_utils.DataLoader(
             data,
@@ -148,10 +154,13 @@ class BaseSGDGMM(ABC):
                     self.optimiser.zero_grad()
 
                     loss = self.module(d)
+                    train_loss += loss.item()
+
+                    n = d[0].shape[0]
+                    loss += self.reg_loss(n, n_total)
+
                     loss.backward()
                     self.optimiser.step()
-
-                    train_loss += loss.item()
 
                 train_loss_curve.append(train_loss)
 

### 2019-09-08 deconv/experiments/gaia/fit_baseline.sh
Allow more time for baseline.

@@ -3,7 +3,7 @@
 #SBATCH -n 1      # tasks requested
 #SBATCH --mem=14000  # memory in Mb
 #SBATCH --cpus-per-task=8
-#SBATCH --time=0-10:00:00
+#SBATCH --time=0-20:00:00
 #SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
 #SBATCH -e output/baseline_error-%A_%a.txt  # send stderr to errfile
 #SBATCH --partition=apollo

### 2019-09-09 ml4ps_paper/figures/learning.py
Fix scale on Figure 1.

@@ -55,6 +55,7 @@ for l, result in zip(labels, table.values()):
     axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l)
 
 axes[1].legend(loc='lower left', bbox_to_anchor=(1.01, 0))
+axes[1].set_ylim(-1, 430)
 
 fig.tight_layout()
 

### 2019-09-09 ml4ps_paper/paper.tex
Update results.
Add Appendix B.

@@
 Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
+Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
 Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
@@
   \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.
-Average over 5 runs with standard deviation.}
+Average over 10 runs with standard deviation.}
   \label{results-table}
@@
       \midrule
-    Existing EM                                     & 64  & $-26.12 \pm 0.03$ & - & $60.0 \pm 7.9$\\
-    (\citet{bovyExtremeDeconvolutionInferring2011}) & 128 & $-25.95 \pm 0.04$  & - & $101.1 \pm 2.19$    \\
-                                                    & 256 & $-25.76 \pm 0.02$ & - & $195.3 \pm 4.08$ \\
+      Existing EM & 64 & $-26.10 \pm 0.03$ & - & $55.53 \pm 7.17$ \\
+      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - & $100.30 \pm 1.96$ \\
+       & 256 & $-25.76 \pm 0.02$ & - & $195.45 \pm 3.93$ \\
+       & 512 & $-25.67 \pm 0.01$ & - & $385.60 \pm 8.24$ \\
       \midrule
-    Minibatch EM  & 64  & $-26.04 \pm 0.01$   & - & $26.8 \pm 1.71$ \\
-                  & 128 & $-25.92 \pm 0.01$   & - & $44.5 \pm 1.0$      \\
-                  & 256 & $-25.83 \pm 0.003$  & - & $51.6 \pm 3.0$ \\
+      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - & $30.98 \pm 0.51$ \\
+       & 128 & $-25.91 \pm 0.01$ & - & $47.01 \pm 0.67$ \\
+       & 256 & $-25.83 \pm 0.00$ & - & $55.18 \pm 2.89$ \\
+       & 512 & $-25.80 \pm 0.00$ & - & $100.47 \pm 3.89$ \\
       \midrule
-    SGD & 64    & $-26.10 \pm 0.02$  & - &$23.5 \pm 2.34$ \\
-        & 128   & $-25.94 \pm 0.01$  & - & $34.0 \pm 0.83$ \\
-        & 256   & $-25.84 \pm 0.01$  & - & $36.6 \pm 1.3$ \\
+      SGD & 64 & $-25.89 \pm 0.02$ & - & $28.02 \pm 0.77$ \\
+       & 128 & $-25.77 \pm 0.02$ & - & $36.44 \pm 0.44$ \\
+       & 256 & $-25.67 \pm 0.02$ & - & $40.54 \pm 1.48$ \\
+       & 512 & $-25.59 \pm 0.02$ & - & $64.77 \pm 2.91$ \\
       \bottomrule
@@
 The question remains of which method should be used for any particular problem?
-The existing EM method had comparable log-likelihoods for smaller sizes of $K$, and did marginally better for the largest size, but at the cost of a much longer training time.
-The SGD method is faster than the minibatch EM method, especially as $K$ increases, at a cost of having a slightly worse log-likelihood.
-The answer will therefore depend on exactly what dataset we wish to perform density estimation on, and how accurate an estimation we require.
-In practice we would suggest using the minibatch EM method for relatively smaller datasets, and the SGD method on larger datasets.
+The SGD method provides marginally better log-likelihood values, is faster to train, and scales better with component size $K$.
+In addition, we found it to be more numerically stable during training, making it the obvious choice.
 
@@
 \section{Experiment Details}
+\label{apx:repro}
+
+Here we provide specific details of our experiments for reproducibility.
+
+\subsection{Dataset}
+
+From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
+Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $2,000,000$.
+Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
+Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
+For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
+
+\subsection{Initialisation}
+For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
+Covariances $V_j$ were set to the identity matrix.
+
+\subsection{Training}
+
+All methods were trained for a total of twenty epochs.
+Both minibatch methods used a batch size of $500$.
+For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
+For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
+
+For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$, following the scheme of the original publication using $W = wI$, where $w = 10^{-3}$ and $I$ is the identity matrix.
 

### 2019-09-09 ml4ps_paper/paper.tex
Remove times from results table.

@@
 Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method along with average training time.
+Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
 Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
@@
 \begin{table}{}
-  \caption{Average validation and test log-likelihoods with average training time for the Gaia data subset for different numbers of mixture components $K$.
+  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
 Average over 10 runs with standard deviation.}
@@
       \toprule
-      Method     & K &  Validation     & Test & Time (minutes) \\
+      Method     & K &  Validation     & Test\\
       \midrule
-      Existing EM & 64 & $-26.10 \pm 0.03$ & - & $55.53 \pm 7.17$ \\
-      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - & $100.30 \pm 1.96$ \\
-       & 256 & $-25.76 \pm 0.02$ & - & $195.45 \pm 3.93$ \\
-       & 512 & $-25.67 \pm 0.01$ & - & $385.60 \pm 8.24$ \\
+      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
+      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
+       & 256 & $-25.76 \pm 0.02$ & - \\
+       & 512 & $-25.67 \pm 0.01$ & - \\
       \midrule
-      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - & $30.98 \pm 0.51$ \\
-       & 128 & $-25.91 \pm 0.01$ & - & $47.01 \pm 0.67$ \\
-       & 256 & $-25.83 \pm 0.00$ & - & $55.18 \pm 2.89$ \\
-       & 512 & $-25.80 \pm 0.00$ & - & $100.47 \pm 3.89$ \\
+      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
+       & 128 & $-25.91 \pm 0.01$ & - \\
+       & 256 & $-25.83 \pm 0.00$ & - \\
+       & 512 & $-25.80 \pm 0.00$ & - \\
       \midrule
-      SGD & 64 & $-25.89 \pm 0.02$ & - & $28.02 \pm 0.77$ \\
-       & 128 & $-25.77 \pm 0.02$ & - & $36.44 \pm 0.44$ \\
-       & 256 & $-25.67 \pm 0.02$ & - & $40.54 \pm 1.48$ \\
-       & 512 & $-25.59 \pm 0.02$ & - & $64.77 \pm 2.91$ \\
+      SGD & 64 & $-25.89 \pm 0.02$ & - \\
+       & 128 & $-25.77 \pm 0.02$ & - \\
+       & 256 & $-25.67 \pm 0.02$ & - \\
+       & 512 & $-25.59 \pm 0.02$ & - \\
       \bottomrule

### 2019-09-09 ml4ps_paper/figures/results_table.py

@@ -55,7 +55,7 @@ def get_table(results_dir):
 def print_table(table):
 
     print(r'\toprule')
-    print(r'Method     & K &  Validation     & Test & Time (minutes) \\')
+    print(r'Method     & K &  Validation     & Test\\')
 
     labels = (
         'Existing EM',
@@ -67,12 +67,10 @@ def print_table(table):
         print(r'\midrule')
         for i, (k, r) in enumerate(m.items()):
 
-            s = ' & {} & ${:.2f} \pm {:.2f}$ & - & ${:.2f} \pm {:.2f}$ \\\\'.format(
+            s = ' & {} & ${:.2f} \pm {:.2f}$ & - \\\\'.format(
                 k,
                 r[0].mean(),
-                r[0].std(),
-                r[1].mean(),
-                r[1].std()
+                r[0].std()
             )
 
             if i == 0:

### 2019-09-09 ml4ps_paper/paper.tex
More paper fixes.

@@
 \newcommand{\bd}{\mathbf{d}}
+\newcommand{\bs}{\mathbf{s}}
 
@@
 Previous work using an extreme deconvolution model on the Gaia catalogue %had to
-worked with a subset of the data,
+worked with a subset of the data and restricted the number of mixture components,
 %as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
@@
 \subsection{Minibatch Expectation-Maximisation}
-
+\label{sec:minibatch-em}
 The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
@@
 \begin{equation}
-r_{ij} = \frac{\alpha_j \mathcal{M}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
+r_{ij} = \frac{\alpha_{j} \mathcal{M}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
 \bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
@@
 \begin{equation}
-\phi_j^1 = \sum_i r_{ij}, \quad
-\phi_j^2 = \sum_i r_{ij} \bb_{ij}, \quad
-\phi_j^3 = \sum_i r_{ij} [\bb_{ij}\bb_{ij}^T + B_{ij}].
+q_{jt} = \sum_i r_{ijt}, \quad
+\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
+S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
 \end{equation}
-Stochastic estimates $\hat{\phi}_{tj} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
+Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
 \begin{equation}
-\hat{\phi}_{tj} = (1 - \lambda)\hat{\phi}_{(t-1)j} + \lambda \phi_{tj}.
+\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
+\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
+\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
 \end{equation}
-Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters.
+Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
 \begin{equation}
-\alpha_j = \frac{\hat{\phi}_j^1}{M}, \quad
-\bm_j = \frac{\hat{\phi}_j^2}{\hat{\phi}_j^1}, \quad
-V_j = \frac{\hat{\phi}_j^3}{\hat{\phi}_j^1} - \bm_j \bm_j^T.
+\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
+\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
+V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
 \label{eqn:mstep}
@@
 \end{equation}
-Having removed the constraints, we optimise the likelihood using any standard minibatch gradient-based optimiser.
+Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
 
@@
 \section{Experiments}
+\label{sec:experiments}
 
@@
 
-Initialisation of all models was performed by first running a minibatch version of the $K$-means clustering algorithm on the dataset to get estimates of the weights and means \cite{sculleyWebscaleKmeansClustering2010}.
-Initial covariances were set to the identity matrix.
-% TODO -- is data standardized?
-Training was done for 20 epochs.
 The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
@@
 Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
+
 Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
-Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$ and training time as function of mixture components $K$.
+The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
+Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
 Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
@@
 
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst being able to work with datasets that cannot fit in memory.
-The results also show that using GPU-based computation speeds up training considerably as mixture component size $K$ increases.
+Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster and being able to work with datasets that cannot fit in memory.
+The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
 
@@
 Our approaches also add more free parameters to be selected, including learning rate and batch size.
-This adds scope for hyperparameter optimisation to improve the log-likelihood, potentially making use of Bayesian optimisation.
+This adds scope for hyperparameter optimisation to improve the log-likelihood.
 
 The question remains of which method should be used for any particular problem?
-The SGD method provides marginally better log-likelihood values, is faster to train, and scales better with component size $K$.
+The SGD method provides comparable log-likelihood values, is faster to train, and scales better with component size $K$.
 In addition, we found it to be more numerically stable during training, making it the obvious choice.
@@
 
+In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
+Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
+For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
+
+First, we define an adjustment operation,
+\begin{align}
+  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
+  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
+\end{align}
+which can be thought of as recentering a scaled variance around a new mean.
+Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
+
+In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
+We also compute minibatch-specific parameters using exact sums over the minibatch:
+\begin{equation}
+q_b = \sum_i^M r_i, \quad
+\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
+\end{equation}{}
+We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
+\begin{align}
+V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
+&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
+&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
+&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
+&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
+\end{align}
+Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
+Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
+
 \section{Experiment Details}
@@
 Here we provide specific details of our experiments for reproducibility.
+Code used to run the experiments will be made available.
 
@@
 
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$, following the scheme of the original publication using $W = wI$, where $w = 10^{-3}$ and $I$ is the identity matrix.
+For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
+Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
+For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
+For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
+As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
+If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
 

### 2019-09-09 ml4ps_paper/figures/learning.py

@@ -26,7 +26,9 @@ baseline_curves = np.array([
     ) and f.endswith('loglike.log')
 ])
 
-for l, result in zip(labels, table.values()):
+linestyles = ('-', '--', '-.')
+
+for l, ls, result in zip(labels, linestyles, table.values()):
     if l != 'Existing EM':
         x = np.arange(1, 21) * (result[256][1].mean() / 20)
         y = result[256][2].mean(axis=0)
@@ -36,7 +38,7 @@ for l, result in zip(labels, table.values()):
         y = baseline_curves.mean(axis=0)
         y_err = 1.96 * baseline_curves.std(axis=0)
 
-    axes[0].plot(x, y)
+    axes[0].plot(x, y, linestyle=ls)
     axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5)
 
 axes[0].set_xscale('log')
@@ -49,10 +51,10 @@ axes[1].set_xlabel(r'Mixture Components $K$')
 axes[1].set_ylabel(r'Time (minutes)')
 axes[1].set_xticks(k)
 
-for l, result in zip(labels, table.values()):
+for l, ls, result in zip(labels, linestyles, table.values()):
     t = np.array([[r[1].mean(), r[1].std()] for r in result.values()])
     print(t.shape)
-    axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l)
+    axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l, linestyle=ls)
 
 axes[1].legend(loc='lower left', bbox_to_anchor=(1.01, 0))
 axes[1].set_ylim(-1, 430)

### 2019-09-13 deconv/experiments/gaia/score_test_sgd.py
Add SGD test set scoring.

@@ -1,0 +1,64 @@
+import argparse
+import json
+import os
+
+import numpy as np
+import torch
+
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM, SGDDeconvDataset
+
+
+def score_sgd(datafile, results_dir, output_file):
+    data = np.load(datafile)
+
+    test_data = SGDDeconvDataset(
+        torch.Tensor(data['X_test']),
+        torch.Tensor(data['C_test'])
+    )
+
+    rf = os.listdir(results_dir)
+
+    param_files = [
+        f for f in rf if f.startswith('sgd_512') and f.endswith('.pkl')
+    ]
+
+    gmm = SGDDeconvGMM(
+        512,
+        7,
+        batch_size=500
+    )
+
+    scores = []
+
+    for p in param_files:
+        state_dict = torch.load(
+            results_dir + p,
+            map_location=torch.device('cpu')
+        )
+        gmm.module.load_state_dict(state_dict)
+
+        test_score = gmm.score_batch(test_data)
+        print(test_score)
+
+        scores.append(test_score)
+
+    print('Test Score: {} +- {}'.format(
+        np.mean(scores),
+        np.std(scores)
+    ))
+
+    json.dump(scores, open(output_file, 'w'))
+
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('datafile')
+    parser.add_argument('results_dir')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    score_sgd(args.datafile, args.results_dir, arg.output_file)

### 2019-09-13 deconv/experiments/gaia/score_test_sgd.py
Typo fix

@@ -50,7 +50,6 @@ def score_sgd(datafile, results_dir, output_file):
     json.dump(scores, open(output_file, 'w'))
 
 
-
 if __name__ == '__main__':
 
     parser = argparse.ArgumentParser()
@@ -61,4 +60,4 @@ if __name__ == '__main__':
 
     args = parser.parse_args()
 
-    score_sgd(args.datafile, args.results_dir, arg.output_file)
+    score_sgd(args.datafile, args.results_dir, args.output_file)

### 2019-09-13 deconv/experiments/gaia/score_test_baseline.py
Add test scoring for baseline and em.

@@ -1,0 +1,57 @@
+import argparse
+import json
+import os
+
+import numpy as np
+
+from extreme_deconvolution import extreme_deconvolution
+
+
+def score_baseline(datafile, results_dir, output_file):
+    data = np.load(datafile)
+
+    rf = os.listdir(results_dir)
+
+    param_files = [
+        f for f in rf if f.startswith('baseline_512') and f.endswith('.npz')
+    ]
+
+    scores = []
+
+    for p in param_files:
+        params = np.load(results_dir + p)
+        weights = params['weights']
+        means = params['means']
+        covars = params['covar']
+
+        test_score = extreme_deconvolution(
+            data['X_test'],
+            data['C_test'],
+            weights,
+            means,
+            covars,
+            likeonly=True
+        )
+        print(test_score)
+
+        scores.append(test_score)
+
+    print('Test Score: {} +- {}'.format(
+        np.mean(scores),
+        np.std(scores)
+    ))
+
+    json.dump(scores, open(output_file, 'w'))
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('datafile')
+    parser.add_argument('results_dir')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    score_baseline(args.datafile, args.results_dir, args.output_file)

### 2019-09-13 deconv/experiments/gaia/score_test_em.py

@@ -1,0 +1,66 @@
+import argparse
+import json
+import os
+
+import numpy as np
+import torch
+
+from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
+
+
+def score_em(datafile, results_dir, output_file):
+    data = np.load(datafile)
+
+    test_data = SGDDeconvDataset(
+        torch.Tensor(data['X_test']),
+        torch.Tensor(data['C_test'])
+    )
+
+    rf = os.listdir(results_dir)
+
+    param_files = [
+        f for f in rf if f.startswith('em_512') and f.endswith('.pkl')
+    ]
+
+    gmm = OnlineDeconvGMM(
+        512,
+        7,
+        batch_size=500
+    )
+
+    scores = []
+
+    for p in param_files:
+        weights, means, covars = torch.load(
+            results_dir + p,
+            map_location=torch.device('cpu')
+        )
+        gmm.weights = weights
+        gmm.means = means
+        gmm.covars = covars
+
+        test_score = gmm.score_batch(test_data)
+        print(test_score)
+
+        scores.append(test_score)
+
+    print('Test Score: {} +- {}'.format(
+        np.mean(scores),
+        np.std(scores)
+    ))
+
+    json.dump(scores, open(output_file, 'w'))
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('datafile')
+    parser.add_argument('results_dir')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    score_em(args.datafile, args.results_dir, args.output_file)

### 2019-09-13 ml4ps_paper/figures/results_table.py
Add test scores to table script.

@@ -48,7 +48,24 @@ def get_table(results_dir):
                 scores = scores * -1
                 curve = curve * -1
 
-            table[model][size] = (scores, times, curve)
+            if size == 512:
+                test_scores = np.array(json.load(
+                    open(results_dir + '{}_test_scores.json'.format(
+                        model
+                    ))
+                ))
+
+                if model != 'baseline':
+                    test_scores /= 200000
+
+                if model == 'sgd':
+                    test_scores *= -1
+
+                table[model][size] = (
+                    scores, times, curve, test_scores
+                )
+            else:
+                table[model][size] = (scores, times, curve)
 
     return table
 
@@ -67,11 +84,20 @@ def print_table(table):
         print(r'\midrule')
         for i, (k, r) in enumerate(m.items()):
 
-            s = ' & {} & ${:.2f} \pm {:.2f}$ & - \\\\'.format(
-                k,
-                r[0].mean(),
-                r[0].std()
-            )
+            if k == 512:
+                s = ' & {} & ${:.2f} \pm {:.2f}$ & ${:.2f} \pm {:.2f}$ \\\\'.format(
+                    k,
+                    r[0].mean(),
+                    r[0].std(),
+                    r[3].mean(),
+                    r[3].std()
+                )
+            else:
+                s = ' & {} & ${:.2f} \pm {:.2f}$ & - \\\\'.format(
+                    k,
+                    r[0].mean(),
+                    r[0].std()
+                )
 
             if i == 0:
                 s = l + s

### 2019-09-15 ml4ps_paper/paper.tex
Add test scores to paper.
Small tweaks, and some comments (marked TODO)

@@
 \usepackage{graphicx}
+\usepackage[autolanguage]{numprint}
 
@@
 
-The Extreme Deconvolution method fits a probability density to a dataset where every observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
+The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
 % The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
@@
 We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
+% TODO do we really want to end with scaling to much larger datasets?
+Not seeing test of scaling with N.
+And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
+(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
+And if want to make other claims, plan future expts accordingly.
 
@@
 \begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K
+p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
 \end{equation}
-As both the noise model is Gaussian and the density estimation model is a mixture of Gaussians, the likelihood of $\bx_i$ is also a Gaussian mixture,
-and the total log-likehood of the dataset is
+Because the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
+The total log-likehood of the model is
 \begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
+\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
 \end{equation}
@@
 \begin{equation}
-r_{ij} = \frac{\alpha_{j} \mathcal{M}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
+r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
 \bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
@@
 An alternative to EM-based methods is to optimise the log-likelihood directly.
-The log-likelihood has some constraints on its parameters, namely that the mixture weights $a_j$ are positive and sum to $1$ and that the covariances $V_j$ are positive definite.
-Directly fitting it with standard gradient-based optimisers requires a transformation of the parameters to remove the constraints.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ as a lower triangular matrix $L_j$ via the Cholesky decomposition, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of the unconstrained element $\tilde{L}_q$,
+The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
+Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints.
+The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
 \begin{equation}
@@
 Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
+% TODO we didn't invent this style of approach! Should find a citation or two where this sort
+% of thing is done.
 
@@
 We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this value.
+Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
 Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
@@
        & 256 & $-25.76 \pm 0.02$ & - \\
-       & 512 & $-25.67 \pm 0.01$ & - \\
+       & 512 & $-25.67 \pm 0.01$ & $-25.66 \pm 0.01$ \\
       \midrule
@@
        & 256 & $-25.83 \pm 0.00$ & - \\
-       & 512 & $-25.80 \pm 0.00$ & - \\
+       & 512 & $-25.80 \pm 0.00$ & $-25.79 \pm 0.00$ \\
       \midrule
@@
        & 256 & $-25.67 \pm 0.02$ & - \\
-       & 512 & $-25.59 \pm 0.02$ & - \\
+       & 512 & $-25.59 \pm 0.02$ & $-25.57 \pm 0.02$ \\
       \bottomrule
@@
 
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster and being able to work with datasets that cannot fit in memory.
+Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
+% and being able to work with datasets that cannot fit in memory.
+% TODO I commented out as just an implementation detail? Not seeing an actual test of batch (can be accumulated in chunks) vs minibatch opt.
 The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
@@
 This adds scope for hyperparameter optimisation to improve the log-likelihood.
+% TODO for longer/final version. Is split-merge actually worthwhile? Seems worth
+% doing quick pilot study to see if actually worth investigating further.
 
-The question remains of which method should be used for any particular problem?
-The SGD method provides comparable log-likelihood values, is faster to train, and scales better with component size $K$.
-In addition, we found it to be more numerically stable during training, making it the obvious choice.
+In this preliminary study the SGD method provided the best log-likelihood values, was faster to train, and scaled better with component size $K$.
+In addition, we found SGD to be more numerically stable than minibatch-EM during training.
 
@@
 From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $2,000,000$.
+Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
 Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
@@
 For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
+% TODO cite scikit learn too if using their implementation?
 Covariances $V_j$ were set to the identity matrix.

### 2019-09-16 ml4ps_paper/paper.tex
add comment with more challenging of conclusions
Smal final changes.

@@
 
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
-
 \title{Scalable Extreme Deconvolution}
@@
 We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
+We demonstrate that these methods provide faster fitting, whilst being able to scale to much larger models for use with larger datasets.
 % TODO do we really want to end with scaling to much larger datasets?
@@
 And if want to make other claims, plan future expts accordingly.
+% Something like larger modesl required for larger datasets.
 
@@
 The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
-Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints.
+Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints~\cite{williams1996}.
 The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
@@
 Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
-% TODO we didn't invent this style of approach! Should find a citation or two where this sort
-% of thing is done.
 
@@
 Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
-% and being able to work with datasets that cannot fit in memory.
-% TODO I commented out as just an implementation detail? Not seeing an actual test of batch (can be accumulated in chunks) vs minibatch opt.
 The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
@@
 This adds scope for hyperparameter optimisation to improve the log-likelihood.
-% TODO for longer/final version. Is split-merge actually worthwhile? Seems worth
-% doing quick pilot study to see if actually worth investigating further.
 
@@
 In addition, we found SGD to be more numerically stable than minibatch-EM during training.
+Further investigation is required to establish exactly why the minibatch-EM method suffers from numerical instability, and whether this method can be tuned to match the performance of the SGD method.
+
+% TODO again, always aim to question claims and have deep understanding. Is fast
+% to train per epoch and/or convergence? Could Adam be applied to the
+% minibatch-EM updates?
+% z = (1-\lambda) z + \lambda z'
+%   = z - \lambda (z - z')
+% interpret (z-z') as a gradient?
+%
+% Do we understand why the scaling with K is better. Implementation detail that
+% might change, or some matrix operation that's intrinsically more annoying?
+%
+% Would be nice to pin down numerical issues further. Although could decide not
+% worth it, if it wouldn't be better even if infinite precision were free.
 
@@
 For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-% TODO cite scikit learn too if using their implementation?
 Covariances $V_j$ were set to the identity matrix.
@@
 Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
+For the minibatch-EM method, the step size $\lambda$ was started at $10^{-2}$ and reduced by a factor of two after ten epochs.
 For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.

### 2019-09-23 ml4ps_paper/paper.tex
Final edit.

@@
 % The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
+The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog containing noisy observations of a billion stars.
 We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
@@
 \end{equation}
-Because the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
+As the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
 The total log-likehood of the model is
@@
 In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-Further investigation is required to establish exactly why the minibatch-EM method suffers from numerical instability, and whether this method can be tuned to match the performance of the SGD method.
+Both minibatch methods will allow us to fit larger models going forwards.
 
@@
 For the minibatch-EM method, the step size $\lambda$ was started at $10^{-2}$ and reduced by a factor of two after ten epochs.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
+For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of ten for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
 

### 2019-09-23 aistats_paper/sample_paper.tex
Add AISTATS paper pack.

\documentclass[twoside]{article}

\usepackage{aistats2020}
% If your paper is accepted, change the options for the package
% aistats2020 as follows:
%
% \usepackage[accepted]{aistats2020}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
%\usepackage[round]{natbib}
%\renewcommand{\bibname}{References}
%\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for paper submissions to AISTATS 2020}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The \textbf{Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 1~inch (6~picas).
Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

\subsection{CITATIONS, FIGURES, REFERENCES}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). References should follow any style that
you are used to using, as long as their style is consistent throughout
the paper.  Be sure that the sentence reads correctly if the citation
is deleted: e.g., instead of ``As described by (Cheesman, 1985), we
first frobulate the widgets,'' write ``As described by Cheesman
(1985), we first frobulate the widgets.''  %Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title,
and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
\textbf{PART}  &\textbf{DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you
can include them in the supplementary material file.

\section{INSTRUCTIONS FOR CAMERA-READY PAPERS}

For the camera-ready paper, if you are using \LaTeX, please make sure
that you follow these instructions.  (If you are not using \LaTeX,
please make sure to achieve the same effect using your chosen
typesetting package.)

\begin{enumerate}
    \item Download \texttt{fancyhdr.sty} -- the
    \texttt{aistats2020.sty} file will make use of it.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2020\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages. The option \texttt{accepted} for the package
    \texttt{aistats2020.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative
    shorter version of the title of your paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of
    the authors of your paper, all separated by commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    \item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, don't change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}

Use the unnumbered third level heading for the acknowledgements.  All
acknowledgements go at the end of the paper.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Any choice of citation style is
acceptable as long as you are consistent.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

\begin{thebibliography}{}
\setlength{\itemindent}{-\leftmargin}
\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
    \newblock Performance of a stochastic learning microchip.
    \newblock In D. S. Touretzky (ed.),
    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
    San Mateo, Calif.: Morgan Kaufmann.

\bibitem{} F.~Rosenblatt (1962).
    \newblock \textit{Principles of Neurodynamics.}
    \newblock Washington, D.C.: Spartan Books.

\bibitem{} G.~Tesauro (1989).
    \newblock Neurogammon wins computer Olympiad.
    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
\end{thebibliography}
\end{document}

### 2019-09-25 aistats_paper/sections/background.tex
Initial work on aistats paper.

--- 2019-09-23 ml4ps_paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2019
-
-% ready for submission
-% \usepackage{neurips_2019}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2019}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[autolanguage]{numprint}
-
-\setcitestyle{numbers}
-\setcitestyle{square}
-\setcitestyle{comma}
-\bibliographystyle{abbrvnat}
-\setlength{\bibsep}{4pt plus 2pt minus 2pt}
-
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bm}{\mathbf{m}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
-\newcommand{\bd}{\mathbf{d}}
-\newcommand{\bs}{\mathbf{s}}
-
-\title{Scalable Extreme Deconvolution}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{
-  James A. Ritchie\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{james.ritchie@ed.ac.uk} \\
-  \And
-  Iain Murray\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{i.murray@ed.ac.uk} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
-% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog containing noisy observations of a billion stars.
-We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting, whilst being able to scale to much larger models for use with larger datasets.
-% TODO do we really want to end with scaling to much larger datasets?
-Not seeing test of scaling with N.
-And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
-(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
-And if want to make other claims, plan future expts accordingly.
-% Something like larger modesl required for larger datasets.
-
-\end{abstract}
-
-\section{Introduction}
-
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
-The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
-
-The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-Previous work using an extreme deconvolution model on the Gaia catalogue %had to
-worked with a subset of the data and restricted the number of mixture components,
-%as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
-but the intention is to fit models with the full dataset
-\cite{andersonImprovingGaiaParallax2018}.
-The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
-To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
-
-In this work, we develop two minibatch methods for fitting the extreme deconvolution model
-%that can scale to the full dataset size, % -- not demonstrated here?
-based on 1) an online variation of the Expectation-Maximisation (EM) algorithm, and 2) a gradient optimizer.
-Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
-
 \section{Background}
@@
 The total log-likehood of the model is
-\begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
-\end{equation}
-
+\begin{align}
+\mathcal{L}(\theta) &= \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \\
+T_{ij} &= R_iV_jR_i^T + S_i.
+\end{align}
 Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-
-\section{Methods}
-
-\subsection{Minibatch Expectation-Maximisation}
-\label{sec:minibatch-em}
-The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
-At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
-\begin{equation}
-r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
-\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
-B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
-\end{equation}
-The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
-The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
-The expected sufficient statistics are then summed together over the minibatch,
-\begin{equation}
-q_{jt} = \sum_i r_{ijt}, \quad
-\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
-S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
-\end{equation}
-Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
-\begin{equation}
-\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
-\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
-\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
-\end{equation}
-Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
-\begin{equation}
-\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
-\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
-V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
-\label{eqn:mstep}
-\end{equation}
-This procedure is repeated with new randomly-ordered minibatches until convergence.
-If we set $\lambda=1$ and replace each minibatch with the entire dataset,
-% and factorise the update for $V_j$ -- % unclear to someone new?
-then the update corresponds to the original batch fitting method.
-Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}.
-There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
-% symbolically equivalent but
-more numerically stable form.
-
-\subsection{Stochastic Gradient Descent}
-
-An alternative to EM-based methods is to optimise the log-likelihood directly.
-The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
-Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints~\cite{williams1996}.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
-\begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
-V_j = L_jL_j^T, \quad
-(L_j)_{qq} = \exp({\tilde{L}_q}).
-\end{equation}
-Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
-
-For a standard Gaussian mixture model, gradient based optimization has a scaling advantage over EM\@.
-There is no need to form the $D\!\times\!D$ covariance matrix $V_j$, since the Gaussian density can be evaluated directly from the Cholesky factor $L_j$ in $O(D^2)$, whereas an EM update is $O(D^3)$.
-Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
-We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
-In total there were 2 million rows.
-Where data were missing, we set the field to zero and the noise variance to a large value.
-We set the projection $R_i$ to the identity matrix for every sample.
-This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used a range of mixture component sizes $K$.
-In practice we would want to select a value of $K$ by cross-validation.
-
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
-While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
-We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
-Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
-
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
-The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
-Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
-Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
-
-\begin{table}{}
-  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
-Average over 10 runs with standard deviation.}
-  \label{results-table}
-  \centering
-  \begin{tabular}{lcccc}
-      \toprule
-      Method     & K &  Validation     & Test\\
-      \midrule
-      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
-      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
-       & 256 & $-25.76 \pm 0.02$ & - \\
-       & 512 & $-25.67 \pm 0.01$ & $-25.66 \pm 0.01$ \\
-      \midrule
-      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
-       & 128 & $-25.91 \pm 0.01$ & - \\
-       & 256 & $-25.83 \pm 0.00$ & - \\
-       & 512 & $-25.80 \pm 0.00$ & $-25.79 \pm 0.00$ \\
-      \midrule
-      SGD & 64 & $-25.89 \pm 0.02$ & - \\
-       & 128 & $-25.77 \pm 0.02$ & - \\
-       & 256 & $-25.67 \pm 0.02$ & - \\
-       & 512 & $-25.59 \pm 0.02$ & $-25.57 \pm 0.02$ \\
-      \bottomrule
-  \end{tabular}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{figures/learning.pdf}
-  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
-Epochs rescaled by average training time.
-Error bars not visible.
-\textbf{Right}: Training time as a function of mixture components $K$.
-Error bars indicate $\pm$ 2 standard deviations.}
-  \label{fig:training}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics{figures/density.pdf}
-  \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
-  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
-  \label{fig:projection}
-\end{figure}
-
-\section{Discussion}
-
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
-The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
-
-Further improvements to our approaches are possible.
-The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
-It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
-Our approaches also add more free parameters to be selected, including learning rate and batch size.
-This adds scope for hyperparameter optimisation to improve the log-likelihood.
-
-In this preliminary study the SGD method provided the best log-likelihood values, was faster to train, and scaled better with component size $K$.
-In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-Both minibatch methods will allow us to fit larger models going forwards.
-
-% TODO again, always aim to question claims and have deep understanding. Is fast
-% to train per epoch and/or convergence? Could Adam be applied to the
-% minibatch-EM updates?
-% z = (1-\lambda) z + \lambda z'
-%   = z - \lambda (z - z')
-% interpret (z-z') as a gradient?
-%
-% Do we understand why the scaling with K is better. Implementation detail that
-% might change, or some matrix operation that's intrinsically more annoying?
-%
-% Would be nice to pin down numerical issues further. Although could decide not
-% worth it, if it wouldn't be better even if infinite precision were free.
-
-\bibliography{references}
-
-\appendix
-
-\section{Stable Covariance Update}
-\label{apx:variance-rewrite}
-
-In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
-Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
-For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
-
-First, we define an adjustment operation,
-\begin{align}
-  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
-  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
-\end{align}
-which can be thought of as recentering a scaled variance around a new mean.
-Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
-
-In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
-We also compute minibatch-specific parameters using exact sums over the minibatch:
-\begin{equation}
-q_b = \sum_i^M r_i, \quad
-\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
-\end{equation}{}
-We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
-\begin{align}
-V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
-&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
-&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
-&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
-\end{align}
-Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
-Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
-
-\section{Experiment Details}
-\label{apx:repro}
-
-Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
-
-\subsection{Dataset}
-
-From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
-Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
-Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
-For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
-
-\subsection{Initialisation}
-For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-Covariances $V_j$ were set to the identity matrix.
-
-\subsection{Training}
-
-All methods were trained for a total of twenty epochs.
-Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was started at $10^{-2}$ and reduced by a factor of two after ten epochs.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of ten for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
-
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
-Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
-For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
-For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
-As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
-If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
-
-\end{document}

### 2019-09-25 aistats_paper/sections/introduction.tex
Initial work on aistats paper.

--- 2019-09-15 ml4ps_paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2019
-
-% ready for submission
-% \usepackage{neurips_2019}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2019}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[autolanguage]{numprint}
-
-\setcitestyle{numbers}
-\setcitestyle{square}
-\setcitestyle{comma}
-\bibliographystyle{abbrvnat}
-\setlength{\bibsep}{4pt plus 2pt minus 2pt}
-
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bm}{\mathbf{m}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
-\newcommand{\bd}{\mathbf{d}}
-\newcommand{\bs}{\mathbf{s}}
-
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
-
-\title{Scalable Extreme Deconvolution}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{
-  James A. Ritchie\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{james.ritchie@ed.ac.uk} \\
-  \And
-  Iain Murray\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{i.murray@ed.ac.uk} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
-% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
-We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
-% TODO do we really want to end with scaling to much larger datasets?
-Not seeing test of scaling with N.
-And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
-(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
-And if want to make other claims, plan future expts accordingly.
-
-\end{abstract}
-
 \section{Introduction}
 
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
+Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\citep{bovyExtremeDeconvolutionInferring2011}.
+The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\citep{perrymanHipparcosCatalogue1997}.
 The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
 
-The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
+The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\citep{collaborationGaiaMission2016}.
 Previous work using an extreme deconvolution model on the Gaia catalogue %had to
@@
 but the intention is to fit models with the full dataset
-\cite{andersonImprovingGaiaParallax2018}.
+\citep{andersonImprovingGaiaParallax2018}.
 The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
-To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
+To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \citep{bottou2018}.
 
@@
 Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
-
-\section{Background}
-
-The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
-\begin{equation}
-  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
-\end{equation}
-We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
-\begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
-\end{equation}
-Because the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
-The total log-likehood of the model is
-\begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
-\end{equation}
-
-Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-
-\section{Methods}
-
-\subsection{Minibatch Expectation-Maximisation}
-\label{sec:minibatch-em}
-The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
-At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
-\begin{equation}
-r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
-\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
-B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
-\end{equation}
-The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
-The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
-The expected sufficient statistics are then summed together over the minibatch,
-\begin{equation}
-q_{jt} = \sum_i r_{ijt}, \quad
-\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
-S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
-\end{equation}
-Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
-\begin{equation}
-\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
-\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
-\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
-\end{equation}
-Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
-\begin{equation}
-\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
-\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
-V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
-\label{eqn:mstep}
-\end{equation}
-This procedure is repeated with new randomly-ordered minibatches until convergence.
-If we set $\lambda=1$ and replace each minibatch with the entire dataset,
-% and factorise the update for $V_j$ -- % unclear to someone new?
-then the update corresponds to the original batch fitting method.
-Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}.
-There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
-% symbolically equivalent but
-more numerically stable form.
-
-\subsection{Stochastic Gradient Descent}
-
-An alternative to EM-based methods is to optimise the log-likelihood directly.
-The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
-Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
-\begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
-V_j = L_jL_j^T, \quad
-(L_j)_{qq} = \exp({\tilde{L}_q}).
-\end{equation}
-Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
-% TODO we didn't invent this style of approach! Should find a citation or two where this sort
-% of thing is done.
-
-For a standard Gaussian mixture model, gradient based optimization has a scaling advantage over EM\@.
-There is no need to form the $D\!\times\!D$ covariance matrix $V_j$, since the Gaussian density can be evaluated directly from the Cholesky factor $L_j$ in $O(D^2)$, whereas an EM update is $O(D^3)$.
-Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
-We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
-In total there were 2 million rows.
-Where data were missing, we set the field to zero and the noise variance to a large value.
-We set the projection $R_i$ to the identity matrix for every sample.
-This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used a range of mixture component sizes $K$.
-In practice we would want to select a value of $K$ by cross-validation.
-
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
-While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
-We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
-Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
-
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
-The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
-Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
-Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
-
-\begin{table}{}
-  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
-Average over 10 runs with standard deviation.}
-  \label{results-table}
-  \centering
-  \begin{tabular}{lcccc}
-      \toprule
-      Method     & K &  Validation     & Test\\
-      \midrule
-      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
-      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
-       & 256 & $-25.76 \pm 0.02$ & - \\
-       & 512 & $-25.67 \pm 0.01$ & $-25.66 \pm 0.01$ \\
-      \midrule
-      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
-       & 128 & $-25.91 \pm 0.01$ & - \\
-       & 256 & $-25.83 \pm 0.00$ & - \\
-       & 512 & $-25.80 \pm 0.00$ & $-25.79 \pm 0.00$ \\
-      \midrule
-      SGD & 64 & $-25.89 \pm 0.02$ & - \\
-       & 128 & $-25.77 \pm 0.02$ & - \\
-       & 256 & $-25.67 \pm 0.02$ & - \\
-       & 512 & $-25.59 \pm 0.02$ & $-25.57 \pm 0.02$ \\
-      \bottomrule
-  \end{tabular}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{figures/learning.pdf}
-  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
-Epochs rescaled by average training time.
-Error bars not visible.
-\textbf{Right}: Training time as a function of mixture components $K$.
-Error bars indicate $\pm$ 2 standard deviations.}
-  \label{fig:training}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics{figures/density.pdf}
-  \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
-  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
-  \label{fig:projection}
-\end{figure}
-
-\section{Discussion}
-
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
-% and being able to work with datasets that cannot fit in memory.
-% TODO I commented out as just an implementation detail? Not seeing an actual test of batch (can be accumulated in chunks) vs minibatch opt.
-The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
-
-Further improvements to our approaches are possible.
-The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
-It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
-Our approaches also add more free parameters to be selected, including learning rate and batch size.
-This adds scope for hyperparameter optimisation to improve the log-likelihood.
-% TODO for longer/final version. Is split-merge actually worthwhile? Seems worth
-% doing quick pilot study to see if actually worth investigating further.
-
-In this preliminary study the SGD method provided the best log-likelihood values, was faster to train, and scaled better with component size $K$.
-In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-
-\bibliography{references}
-
-\appendix
-
-\section{Stable Covariance Update}
-\label{apx:variance-rewrite}
-
-In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
-Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
-For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
-
-First, we define an adjustment operation,
-\begin{align}
-  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
-  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
-\end{align}
-which can be thought of as recentering a scaled variance around a new mean.
-Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
-
-In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
-We also compute minibatch-specific parameters using exact sums over the minibatch:
-\begin{equation}
-q_b = \sum_i^M r_i, \quad
-\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
-\end{equation}{}
-We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
-\begin{align}
-V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
-&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
-&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
-&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
-\end{align}
-Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
-Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
-
-\section{Experiment Details}
-\label{apx:repro}
-
-Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
-
-\subsection{Dataset}
-
-From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
-Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
-Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
-For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
-
-\subsection{Initialisation}
-For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-% TODO cite scikit learn too if using their implementation?
-Covariances $V_j$ were set to the identity matrix.
-
-\subsection{Training}
-
-All methods were trained for a total of twenty epochs.
-Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
-
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
-Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
-For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
-For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
-As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
-If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
-
-\end{document}

### 2019-09-25 aistats_paper/sections/background.tex
Hack workshop paper into AISTATS format.

@@
+%!TEX root = ../aistats_paper.tex
 \section{Background}
@@
 We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
-\begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
-\end{equation}
+\begin{multline}
+p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \\ \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
+\end{multline}
 As the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.

### 2019-09-25 aistats_paper/sections/discussion.tex
Hack workshop paper into AISTATS format.

--- 2019-09-15 ml4ps_paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2019
-
-% ready for submission
-% \usepackage{neurips_2019}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2019}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[autolanguage]{numprint}
-
-\setcitestyle{numbers}
-\setcitestyle{square}
-\setcitestyle{comma}
-\bibliographystyle{abbrvnat}
-\setlength{\bibsep}{4pt plus 2pt minus 2pt}
-
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bm}{\mathbf{m}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
-\newcommand{\bd}{\mathbf{d}}
-\newcommand{\bs}{\mathbf{s}}
-
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
-
-\title{Scalable Extreme Deconvolution}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{
-  James A. Ritchie\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{james.ritchie@ed.ac.uk} \\
-  \And
-  Iain Murray\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{i.murray@ed.ac.uk} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
-% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
-We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
-% TODO do we really want to end with scaling to much larger datasets?
-Not seeing test of scaling with N.
-And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
-(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
-And if want to make other claims, plan future expts accordingly.
-
-\end{abstract}
-
-\section{Introduction}
-
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
-The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
-
-The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-Previous work using an extreme deconvolution model on the Gaia catalogue %had to
-worked with a subset of the data and restricted the number of mixture components,
-%as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
-but the intention is to fit models with the full dataset
-\cite{andersonImprovingGaiaParallax2018}.
-The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
-To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
-
-In this work, we develop two minibatch methods for fitting the extreme deconvolution model
-%that can scale to the full dataset size, % -- not demonstrated here?
-based on 1) an online variation of the Expectation-Maximisation (EM) algorithm, and 2) a gradient optimizer.
-Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
-
-\section{Background}
-
-The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
-\begin{equation}
-  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
-\end{equation}
-We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
-\begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
-\end{equation}
-Because the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
-The total log-likehood of the model is
-\begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
-\end{equation}
-
-Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-
-\section{Methods}
-
-\subsection{Minibatch Expectation-Maximisation}
-\label{sec:minibatch-em}
-The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
-At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
-\begin{equation}
-r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
-\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
-B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
-\end{equation}
-The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
-The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
-The expected sufficient statistics are then summed together over the minibatch,
-\begin{equation}
-q_{jt} = \sum_i r_{ijt}, \quad
-\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
-S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
-\end{equation}
-Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
-\begin{equation}
-\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
-\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
-\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
-\end{equation}
-Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
-\begin{equation}
-\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
-\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
-V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
-\label{eqn:mstep}
-\end{equation}
-This procedure is repeated with new randomly-ordered minibatches until convergence.
-If we set $\lambda=1$ and replace each minibatch with the entire dataset,
-% and factorise the update for $V_j$ -- % unclear to someone new?
-then the update corresponds to the original batch fitting method.
-Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}.
-There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
-% symbolically equivalent but
-more numerically stable form.
-
-\subsection{Stochastic Gradient Descent}
-
-An alternative to EM-based methods is to optimise the log-likelihood directly.
-The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
-Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
-\begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
-V_j = L_jL_j^T, \quad
-(L_j)_{qq} = \exp({\tilde{L}_q}).
-\end{equation}
-Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
-% TODO we didn't invent this style of approach! Should find a citation or two where this sort
-% of thing is done.
-
-For a standard Gaussian mixture model, gradient based optimization has a scaling advantage over EM\@.
-There is no need to form the $D\!\times\!D$ covariance matrix $V_j$, since the Gaussian density can be evaluated directly from the Cholesky factor $L_j$ in $O(D^2)$, whereas an EM update is $O(D^3)$.
-Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
-We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
-In total there were 2 million rows.
-Where data were missing, we set the field to zero and the noise variance to a large value.
-We set the projection $R_i$ to the identity matrix for every sample.
-This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used a range of mixture component sizes $K$.
-In practice we would want to select a value of $K$ by cross-validation.
-
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
-While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
-We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
-Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
-
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
-The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
-Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
-Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
-
-\begin{table}{}
-  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
-Average over 10 runs with standard deviation.}
-  \label{results-table}
-  \centering
-  \begin{tabular}{lcccc}
-      \toprule
-      Method     & K &  Validation     & Test\\
-      \midrule
-      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
-      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
-       & 256 & $-25.76 \pm 0.02$ & - \\
-       & 512 & $-25.67 \pm 0.01$ & $-25.66 \pm 0.01$ \\
-      \midrule
-      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
-       & 128 & $-25.91 \pm 0.01$ & - \\
-       & 256 & $-25.83 \pm 0.00$ & - \\
-       & 512 & $-25.80 \pm 0.00$ & $-25.79 \pm 0.00$ \\
-      \midrule
-      SGD & 64 & $-25.89 \pm 0.02$ & - \\
-       & 128 & $-25.77 \pm 0.02$ & - \\
-       & 256 & $-25.67 \pm 0.02$ & - \\
-       & 512 & $-25.59 \pm 0.02$ & $-25.57 \pm 0.02$ \\
-      \bottomrule
-  \end{tabular}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{figures/learning.pdf}
-  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
-Epochs rescaled by average training time.
-Error bars not visible.
-\textbf{Right}: Training time as a function of mixture components $K$.
-Error bars indicate $\pm$ 2 standard deviations.}
-  \label{fig:training}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics{figures/density.pdf}
-  \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
-  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
-  \label{fig:projection}
-\end{figure}
-
+%!TEX root = ../aistats_paper.tex
 \section{Discussion}
@@
 In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-
-\bibliography{references}
-
-\appendix
-
-\section{Stable Covariance Update}
-\label{apx:variance-rewrite}
-
-In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
-Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
-For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
-
-First, we define an adjustment operation,
-\begin{align}
-  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
-  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
-\end{align}
-which can be thought of as recentering a scaled variance around a new mean.
-Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
-
-In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
-We also compute minibatch-specific parameters using exact sums over the minibatch:
-\begin{equation}
-q_b = \sum_i^M r_i, \quad
-\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
-\end{equation}{}
-We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
-\begin{align}
-V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
-&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
-&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
-&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
-\end{align}
-Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
-Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
-
-\section{Experiment Details}
-\label{apx:repro}
-
-Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
-
-\subsection{Dataset}
-
-From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
-Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
-Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
-For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
-
-\subsection{Initialisation}
-For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-% TODO cite scikit learn too if using their implementation?
-Covariances $V_j$ were set to the identity matrix.
-
-\subsection{Training}
-
-All methods were trained for a total of twenty epochs.
-Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
-
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
-Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
-For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
-For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
-As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
-If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
-
-\end{document}
+Both minibatch methods will allow us to fit larger models going forwards.

### 2019-09-25 aistats_paper/sections/experiments.tex
Hack workshop paper into AISTATS format.

--- 2019-09-15 ml4ps_paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2019
-
-% ready for submission
-% \usepackage{neurips_2019}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2019}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[autolanguage]{numprint}
-
-\setcitestyle{numbers}
-\setcitestyle{square}
-\setcitestyle{comma}
-\bibliographystyle{abbrvnat}
-\setlength{\bibsep}{4pt plus 2pt minus 2pt}
-
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bm}{\mathbf{m}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
-\newcommand{\bd}{\mathbf{d}}
-\newcommand{\bs}{\mathbf{s}}
-
-\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
-
-\title{Scalable Extreme Deconvolution}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{
-  James A. Ritchie\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{james.ritchie@ed.ac.uk} \\
-  \And
-  Iain Murray\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{i.murray@ed.ac.uk} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
-% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog, which contains noisy observations of a billion stars.
-We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting of the model, whilst being able to scale to much larger datasets.
-% TODO do we really want to end with scaling to much larger datasets?
-Not seeing test of scaling with N.
-And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
-(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
-And if want to make other claims, plan future expts accordingly.
-
-\end{abstract}
-
-\section{Introduction}
-
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
-The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
-
-The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-Previous work using an extreme deconvolution model on the Gaia catalogue %had to
-worked with a subset of the data and restricted the number of mixture components,
-%as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
-but the intention is to fit models with the full dataset
-\cite{andersonImprovingGaiaParallax2018}.
-The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
-To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
-
-In this work, we develop two minibatch methods for fitting the extreme deconvolution model
-%that can scale to the full dataset size, % -- not demonstrated here?
-based on 1) an online variation of the Expectation-Maximisation (EM) algorithm, and 2) a gradient optimizer.
-Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
-
-\section{Background}
-
-The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
-\begin{equation}
-  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
-\end{equation}
-We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
-\begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
-\end{equation}
-Because the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
-The total log-likehood of the model is
-\begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
-\end{equation}
-
-Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-
-\section{Methods}
-
-\subsection{Minibatch Expectation-Maximisation}
-\label{sec:minibatch-em}
-The original method of fitting the extreme deconvolution model used a modification of the Expectation-Maximisation (EM) algorithm for mixture models~\cite{dempsterMaximumLikelihoodIncomplete1977}.
-Here we describe a minibatch version of this algorithm based on~\citet{cappeOnlineExpectationMaximization2009}'s online EM algorithm for latent data models.
-At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
-\begin{equation}
-r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
-\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
-B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
-\end{equation}
-The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
-The $\bb_{ij}$ and $B_{ij}$ terms result from the fact that $\bx_i$ and $\bv_i$ are jointly Gaussian, so the distribution of $\bv_i$ conditioned on $\bx_i$ is also Gaussian with mean $\bb_{ij}$ and covariance $B_{ij}$.
-The expected sufficient statistics are then summed together over the minibatch,
-\begin{equation}
-q_{jt} = \sum_i r_{ijt}, \quad
-\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
-S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
-\end{equation}
-Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
-\begin{equation}
-\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
-\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
-\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
-\end{equation}
-Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
-\begin{equation}
-\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
-\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
-V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
-\label{eqn:mstep}
-\end{equation}
-This procedure is repeated with new randomly-ordered minibatches until convergence.
-If we set $\lambda=1$ and replace each minibatch with the entire dataset,
-% and factorise the update for $V_j$ -- % unclear to someone new?
-then the update corresponds to the original batch fitting method.
-Numerically however, the update for $V_j$, as written in \eqref{eqn:mstep}, is inadvisable compared to the batch update given in \cite{bovyExtremeDeconvolutionInferring2011}.
-There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
-% symbolically equivalent but
-more numerically stable form.
-
-\subsection{Stochastic Gradient Descent}
-
-An alternative to EM-based methods is to optimise the log-likelihood directly.
-The optimization is constrained, because the mixture weights $a_j$ are positive and sum to $1$, and the covariances $V_j$ are positive definite.
-Directly fitting the log-likelihood with unconstrained gradient-based optimisers requires a transformation of the parameters to remove the constraints.
-The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
-\begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
-V_j = L_jL_j^T, \quad
-(L_j)_{qq} = \exp({\tilde{L}_q}).
-\end{equation}
-Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
-% TODO we didn't invent this style of approach! Should find a citation or two where this sort
-% of thing is done.
-
-For a standard Gaussian mixture model, gradient based optimization has a scaling advantage over EM\@.
-There is no need to form the $D\!\times\!D$ covariance matrix $V_j$, since the Gaussian density can be evaluated directly from the Cholesky factor $L_j$ in $O(D^2)$, whereas an EM update is $O(D^3)$.
-Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
-
+%!TEX root = ../aistats_paper.tex
 \section{Experiments}
@@
 Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
-Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
 
@@
 
-\begin{table}{}
+\begin{table*}{}
   \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
@@
   \end{tabular}
-\end{table}
+\end{table*}
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:training}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:projection}
-\end{figure}
-
-\section{Discussion}
-
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
-% and being able to work with datasets that cannot fit in memory.
-% TODO I commented out as just an implementation detail? Not seeing an actual test of batch (can be accumulated in chunks) vs minibatch opt.
-The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
-
-Further improvements to our approaches are possible.
-The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
-It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
-Our approaches also add more free parameters to be selected, including learning rate and batch size.
-This adds scope for hyperparameter optimisation to improve the log-likelihood.
-% TODO for longer/final version. Is split-merge actually worthwhile? Seems worth
-% doing quick pilot study to see if actually worth investigating further.
-
-In this preliminary study the SGD method provided the best log-likelihood values, was faster to train, and scaled better with component size $K$.
-In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-
-\bibliography{references}
-
-\appendix
-
-\section{Stable Covariance Update}
-\label{apx:variance-rewrite}
-
-In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
-Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
-For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
-
-First, we define an adjustment operation,
-\begin{align}
-  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
-  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
-\end{align}
-which can be thought of as recentering a scaled variance around a new mean.
-Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
-
-In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
-We also compute minibatch-specific parameters using exact sums over the minibatch:
-\begin{equation}
-q_b = \sum_i^M r_i, \quad
-\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
-\end{equation}{}
-We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
-\begin{align}
-V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
-&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
-&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
-&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
-\end{align}
-Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
-Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
-
-\section{Experiment Details}
-\label{apx:repro}
-
-Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
-
-\subsection{Dataset}
-
-From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
-Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
-Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
-For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
-
-\subsection{Initialisation}
-For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-% TODO cite scikit learn too if using their implementation?
-Covariances $V_j$ were set to the identity matrix.
-
-\subsection{Training}
-
-All methods were trained for a total of twenty epochs.
-Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was set to $10^{-2}$.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of 10 for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
-
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
-Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
-For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
-For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
-As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
-If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
-
-\end{document}
+\end{figure*}

### 2019-09-25 aistats_paper/sections/methods.tex
Hack workshop paper into AISTATS format.

--- 2019-09-23 ml4ps_paper/paper.tex
@@
-\documentclass{article}
-
-% if you need to pass options to natbib, use, e.g.:
-%     \PassOptionsToPackage{numbers, compress}{natbib}
-% before loading neurips_2019
-
-% ready for submission
-% \usepackage{neurips_2019}
-
-% to compile a preprint version, e.g., for submission to arXiv, add add the
-% [preprint] option:
-%     \usepackage[preprint]{neurips_2019}
-
-% to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
-
-% to avoid loading the natbib package, add option nonatbib:
-%     \usepackage[nonatbib]{neurips_2019}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{amsmath}
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[autolanguage]{numprint}
-
-\setcitestyle{numbers}
-\setcitestyle{square}
-\setcitestyle{comma}
-\bibliographystyle{abbrvnat}
-\setlength{\bibsep}{4pt plus 2pt minus 2pt}
-
-\newcommand{\bx}{\mathbf{x}}
-\newcommand{\bv}{\mathbf{v}}
-\newcommand{\bm}{\mathbf{m}}
-\newcommand{\bb}{\mathbf{b}}
-\newcommand{\bc}{\mathbf{c}}
-\newcommand{\bd}{\mathbf{d}}
-\newcommand{\bs}{\mathbf{s}}
-
-\title{Scalable Extreme Deconvolution}
-
-% The \author macro works with any number of authors. There are two commands
-% used to separate the names and addresses of multiple authors: \And and \AND.
-%
-% Using \And between authors leaves it to LaTeX to determine where to break the
-% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
-% authors names on the first line, and the last on the second line, try using
-% \AND instead of \And before the third author name.
-
-\author{
-  James A. Ritchie\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{james.ritchie@ed.ac.uk} \\
-  \And
-  Iain Murray\\
-  School of Informatics\\
-  University of Edinburgh\\
-   \texttt{i.murray@ed.ac.uk} \\
-}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-The Extreme Deconvolution method fits a probability density to a dataset where each observation has Gaussian noise added with a known sample-specific covariance, originally intended for use with astronomical datasets.
-% The existing method of fitting the model cannot scale to new larger astronomy datasets. % -- too strong. Probably feasible, and compute only gets cheaper
-The existing fitting method is batch EM, which would not normally be applied to large datasets such as the Gaia catalog containing noisy observations of a billion stars.
-We propose two minibatch variants of extreme deconvolution, based on an online variation of the EM algorithm, and direct gradient-based optimisation of the log-likelihood, both of which can run on GPUs.
-We demonstrate that these methods provide faster fitting, whilst being able to scale to much larger models for use with larger datasets.
-% TODO do we really want to end with scaling to much larger datasets?
-Not seeing test of scaling with N.
-And Bovy's code could be tweaked to accumulate updates without having everything in RAM?
-(And we can get EC2 machines with 1TB of RAM these days...) -- revist what contributions actually are.
-And if want to make other claims, plan future expts accordingly.
-% Something like larger modesl required for larger datasets.
-
-\end{abstract}
-
-\section{Introduction}
-
-Extreme deconvolution is a method that fits Gaussian Mixture Models (GMMs) to noisy data where we know the covariance of the Gaussian noise added to each sample~\cite{bovyExtremeDeconvolutionInferring2011}.
-The method was originally developed to perform probabilistic density estimation on the dataset of stellar velocities produced by the Hipparcos satellite~\cite{perrymanHipparcosCatalogue1997}.
-The Hipparcos catalogue consists of astrometric solutions (positions and velocities on the sky) and photometry (light intensity) for 118,218 stars, with associated noise covariances provided for each entry.
-
-The successor to the Hipparcos mission, Gaia, aims to produce an even larger catalogue, with entries for an estimated 1 billion astronomical objects~\cite{collaborationGaiaMission2016}.
-Previous work using an extreme deconvolution model on the Gaia catalogue %had to
-worked with a subset of the data and restricted the number of mixture components,
-%as the existing method of fitting the models cannot work with very large datasets % -- I don't think the reference says this part?
-but the intention is to fit models with the full dataset
-\cite{andersonImprovingGaiaParallax2018}.
-The existing extreme deconvolution algorithm makes a full pass over the dataset before it can update parameters, and the reference implementation requires all the data to fit in memory.
-To fit such large datasets in reasonable time, we would normally use stochastic or online methods, with updates based on minibatches of data to make the methods practical on GPUs \cite{bottou2018}.
-
-In this work, we develop two minibatch methods for fitting the extreme deconvolution model
-%that can scale to the full dataset size, % -- not demonstrated here?
-based on 1) an online variation of the Expectation-Maximisation (EM) algorithm, and 2) a gradient optimizer.
-Our implementations can run on GPUs, and provide comparable density estimates to the existing method, whilst being much faster to train.
-
-\section{Background}
-
-The aim of extreme deconvolution is to perform density estimation on a noisy $d$-dimensional dataset $\{\bx_i\}_{i=0}^N$, where $\bx_i$ was generated by adding zero-mean Gaussian noise $\epsilon_i$ with known per-datapoint covariance $S_i$ to a projection $R_i$ of a true value $\bv_i$,
-\begin{equation}
-  \bx_i = R_i\bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i).
-\end{equation}
-We assume that $\bv_i$ can be modelled by a mixture of Gaussians with $K$ components,
-\begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
-\end{equation}
-As the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
-The total log-likehood of the model is
-\begin{equation}
-\mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij}), \quad T_{ij} = R_iV_jR_i^T + S_i.
-\end{equation}
-
-Missing data can be handled either by making $R_i$ rank-deficient, or by setting elements of the covariance matrix $S_i$ to very large values.
-
+%!TEX root = ../aistats_paper.tex
 \section{Methods}
@@
 At each iteration $t$, we compute the sufficient statistics of the latent data $\bv_i$ for each component $j$ in the minibatch of size $M$, using our current estimate of the parameters,
-\begin{equation}
-r_{ij} = \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \ 
-\bb_{ij} = m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \ 
-B_{ij} = V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
-\end{equation}
+\begin{align}
+r_{ij} &= \frac{\alpha_{j} \,\mathcal{N}(\bx_i \mid R_i\bm_j, T_{ij})}{\sum_k \alpha_k \,\mathcal{N}(\bx_i \mid R_i\bm_k, T_{ik})}, \\ 
+\bb_{ij} &= m_j + V_j R_i^T T_{ij}^{-1}(\bx_i - R_i \bm_j), \\
+B_{ij} &= V_j - V_j R_i^T T_{ij}^{-1}R_iV_j.
+\end{align}
 The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
@@
 The expected sufficient statistics are then summed together over the minibatch,
-\begin{equation}
-q_{jt} = \sum_i r_{ijt}, \quad
-\bs_{jt} = \sum_i r_{ijt} \bb_{ijt}, \quad
-S_{jt} = \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
-\end{equation}
+\begin{align}
+q_{jt} &= \sum_i r_{ijt}, \\
+\bs_{jt} &= \sum_i r_{ijt} \bb_{ijt}, \\
+S_{jt} &= \sum_i r_{ijt} [\bb_{ijt}\bb_{ijt}^T + B_{ijt}].
+\end{align}
 Stochastic estimates $\hat{\phi}_{jt} $ of the sums of sufficient statistics over the whole dataset are then updated with a sufficiently small step size $\lambda$,
-\begin{equation}
-\hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\quad
-\phi_{jt} = \{q_{jt}, \bs_{jt}, S_{jt} \},\quad
-\hat{\phi}_{jt} = \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
-\end{equation}
+\begin{align}
+\hat{\phi}_{jt} &= (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt},\\
+\phi_{jt} &= \{q_{jt}, \bs_{jt}, S_{jt} \},\\
+\hat{\phi}_{jt} &= \{\hat{q}_{jt}, \hat{\bs}_{jt}, \hat{S}_{jt} \}. \label{eq:sums}
+\end{align}
 Finally, we normalise the updated sums of expected sufficient statistics to get updated estimates of the parameters,
-\begin{equation}
-\alpha_{j} = \frac{\hat{q}_{jt}}{M}, \quad
-\bm_{j} = \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \quad
-V_{j} = \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
+\begin{align}
+\alpha_{j} &= \frac{\hat{q}_{jt}}{M}, \\
+\bm_{j} &= \frac{\hat{\bs}_{jt}}{\hat{q}_{jt}}, \\
+V_{j} &= \frac{\hat{S}_{jt}}{\hat{q}_{jt}} - \bm_{j} \bm_{j}^T.
 \label{eqn:mstep}
-\end{equation}
+\end{align}
 This procedure is repeated with new randomly-ordered minibatches until convergence.
@@
 There is likely to be catastrophic cancellation if the variances of the components are small relative to the means, especially if single precision floats are used, as is standard with GPU computation.
-In Appendix~\ref{apx:variance-rewrite} we show how the minibatch version of this update can be rewritten in a
-% symbolically equivalent but
-more numerically stable form.
 
@@
 The mixture weights $\alpha_j$ can be parameterised by taking the softmax of an unconstrained vector $\mathbf{z}$, and the covariances $V_j$ represented by its lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
-\begin{equation}
-\alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \quad
-V_j = L_jL_j^T, \quad
-(L_j)_{qq} = \exp({\tilde{L}_q}).
-\end{equation}
+\begin{align}
+\alpha_j &= \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}, \\
+V_j &= L_jL_j^T, \\
+(L_j)_{qq} &= \exp({\tilde{L}_q}).
+\end{align}
 Having removed the constraints, we can optimise the likelihood using any standard minibatch gradient-based optimiser.
@@
 Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
-
-\section{Experiments}
-\label{sec:experiments}
-
-We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
-We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
-In total there were 2 million rows.
-Where data were missing, we set the field to zero and the noise variance to a large value.
-We set the projection $R_i$ to the identity matrix for every sample.
-This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used a range of mixture component sizes $K$.
-In practice we would want to select a value of $K$ by cross-validation.
-
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
-While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
-We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
-Details required for reproducibility are provided in Appendix~\ref{apx:repro}.
-
-Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
-The values are similar, but not exactly comparable, as the effect of regularisation differs for each method.
-Figure~\ref{fig:training} plots the training log-likelihood against time-rescaled epoch for $K=256$, and training time as function of mixture components $K$.
-Figure~\ref{fig:projection} shows a 2-D projection from an example model with $K=256$ fitted with the minibatch-EM method.
-
-\begin{table}{}
-  \caption{Average validation log-likelihoods for the Gaia data subset for different numbers of mixture components $K$, with average test log-likelihood for the best value of $K$ by validation.
-Average over 10 runs with standard deviation.}
-  \label{results-table}
-  \centering
-  \begin{tabular}{lcccc}
-      \toprule
-      Method     & K &  Validation     & Test\\
-      \midrule
-      Existing EM & 64 & $-26.10 \pm 0.03$ & - \\
-      \citet{bovyExtremeDeconvolutionInferring2011} & 128 & $-25.96 \pm 0.04$ & - \\
-       & 256 & $-25.76 \pm 0.02$ & - \\
-       & 512 & $-25.67 \pm 0.01$ & $-25.66 \pm 0.01$ \\
-      \midrule
-      Minibatch EM & 64 & $-26.05 \pm 0.01$ & - \\
-       & 128 & $-25.91 \pm 0.01$ & - \\
-       & 256 & $-25.83 \pm 0.00$ & - \\
-       & 512 & $-25.80 \pm 0.00$ & $-25.79 \pm 0.00$ \\
-      \midrule
-      SGD & 64 & $-25.89 \pm 0.02$ & - \\
-       & 128 & $-25.77 \pm 0.02$ & - \\
-       & 256 & $-25.67 \pm 0.02$ & - \\
-       & 512 & $-25.59 \pm 0.02$ & $-25.57 \pm 0.02$ \\
-      \bottomrule
-  \end{tabular}
-\end{table}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{figures/learning.pdf}
-  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
-Epochs rescaled by average training time.
-Error bars not visible.
-\textbf{Right}: Training time as a function of mixture components $K$.
-Error bars indicate $\pm$ 2 standard deviations.}
-  \label{fig:training}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics{figures/density.pdf}
-  \caption{Density plot showing a 2-D projection of 100000 samples drawn from a model with $K=256$ and fitted with the minibatch EM method.
-  The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
-  \label{fig:projection}
-\end{figure}
-
-\section{Discussion}
-
-Our results have shown that both of our proposed methods perform comparably to the existing method of fitting extreme deconvolution models, whilst converging faster.
-The results also show that using GPU-based computation speeds up fitting considerably, allowing sublinear scaling of training time with mixture component size $K$.
-
-Further improvements to our approaches are possible.
-The original paper presents a method of getting out of local maxima by splitting and merging clusters, with the split-merge criteria evaluated on the whole dataset.
-It should be possible to replace the criteria with stochastic estimates, which would permit them to be used with both the SGD and minibatch EM methods.
-Our approaches also add more free parameters to be selected, including learning rate and batch size.
-This adds scope for hyperparameter optimisation to improve the log-likelihood.
-
-In this preliminary study the SGD method provided the best log-likelihood values, was faster to train, and scaled better with component size $K$.
-In addition, we found SGD to be more numerically stable than minibatch-EM during training.
-Both minibatch methods will allow us to fit larger models going forwards.
-
-% TODO again, always aim to question claims and have deep understanding. Is fast
-% to train per epoch and/or convergence? Could Adam be applied to the
-% minibatch-EM updates?
-% z = (1-\lambda) z + \lambda z'
-%   = z - \lambda (z - z')
-% interpret (z-z') as a gradient?
-%
-% Do we understand why the scaling with K is better. Implementation detail that
-% might change, or some matrix operation that's intrinsically more annoying?
-%
-% Would be nice to pin down numerical issues further. Although could decide not
-% worth it, if it wouldn't be better even if infinite precision were free.
-
-\bibliography{references}
-
-\appendix
-
-\section{Stable Covariance Update}
-\label{apx:variance-rewrite}
-
-In Section~\ref{sec:minibatch-em} describing our minibatch-EM method, we noted that the M-step update for the variance of each component $V_j$ as presented in Equation~\ref{eqn:mstep} would be prone to catastrophic cancellation as a result of taking a small difference between large values using single-precision floats.
-Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
-For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
-
-First, we define an adjustment operation,
-\begin{align}
-  \text{adjust}(V, s, \bc, \bd) &= sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T + \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1} \\
-  &= s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
-\end{align}
-which can be thought of as recentering a scaled variance around a new mean.
-Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
-
-In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
-We also compute minibatch-specific parameters using exact sums over the minibatch:
-\begin{equation}
-q_b = \sum_i^M r_i, \quad
-\bm_b = \frac{\sum_i^M r_i \bx_i}{q_b},\quad V_b = \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
-\end{equation}{}
-We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
-\begin{align}
-V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) + \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}\\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
-&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
-&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
-&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
-\end{align}
-Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
-Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
-
-\section{Experiment Details}
-\label{apx:repro}
-
-Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
-
-\subsection{Dataset}
-
-From the Gaia DR2 source table we selected the columns \texttt{RA}, \texttt{DEC}, \texttt{PARALLAX}, \texttt{PMRA}, \texttt{PMDA}, \texttt{BP\_RP} and \texttt{PHOT\_G\_MEAN\_MAG} to assemble the observed dataset $\{\bx_i\}_{i=0}^N$~\cite{brownGaiaDataRelease2018}.
-Random subsampling was done by selecting rows with the value of the \texttt{RANDOM\_INDEX} column less than $\numprint{2000000}$.
-Noise covariance matrices $S_i$ were assembled using the corresponding error and correlation columns for each variable.
-Where a column of a row was marked as missing, the corresponding element of $\bx_i$ was set to zero, the corresponding diagonal element of $S_i$ was set to $10^{12}$, and the corresponding off-diagonal elements set to zero.
-For columns which do not have associated noise, the corresponding diagonal elements of $S_i$ were set to a value of $10^{-2}$, and corresponding off-diagonal elements to zero.
-
-\subsection{Initialisation}
-For each method, initialisation of the means and weights was done using the estimated counts and centroids after 10 epochs of minibatch k-means clustering~\cite{sculleyWebscaleKmeansClustering2010}.
-Covariances $V_j$ were set to the identity matrix.
-
-\subsection{Training}
-
-All methods were trained for a total of twenty epochs.
-Both minibatch methods used a batch size of $500$.
-For the minibatch-EM method, the step size $\lambda$ was started at $10^{-2}$ and reduced by a factor of two after ten epochs.
-For the SGD method, we used the Adam optimiser with a learning of $10^{-2}$ for the first ten epochs, reducing by a factor of ten for the last ten epochs, and all other parameters set to the suggested defaults~\cite{kingmaAdamMethodStochastic2014}.
-
-For numerical stability, a very small amount of regularisation was applied to the covariances $V_j$.
-Using the original implementation, we set the regularisation constant $w = 10^{-3}$.
-For the minibatch-EM method, we added diagonal matrix $wI$ directly to each covariance matrix after updating them.
-For the SGD method, we added a penalty term $\sum_j \frac{w}{\mathrm{Trace}[V_j]}$ to the loss function.
-As noted in Section~\ref{sec:experiments}, the effect of $w$ is not comparable across methods.
-If we were using larger values of $w$ to prevent overfitting rather than just avoiding numerical instability, $w$ would be tuned specifically for each method.
-
-\end{document}

### 2019-09-25 aistats_paper/figures/learning.py

@@ -1,0 +1,66 @@
+import os
+import numpy as np
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes, mark_inset
+import seaborn as sns
+
+from results_table import get_table
+
+sns.set()
+sns.set_context('paper')
+
+fig, axes = plt.subplots(1, 2, figsize=(6, 2)) 
+
+table = get_table('../results/')
+
+axes[0].set_xlabel('Time-scaled Epoch')
+axes[0].set_ylabel('Avg. Log-likelihood')
+
+labels = ('Existing EM', 'Minibatch EM', 'SGD')
+
+result_files = os.listdir('../results/')
+baseline_curves = np.array([
+    np.loadtxt('../results/' + f) for f in result_files if f.startswith(
+        'baseline_256'
+    ) and f.endswith('loglike.log')
+])
+
+linestyles = ('-', '--', '-.')
+
+for l, ls, result in zip(labels, linestyles, table.values()):
+    if l != 'Existing EM':
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
+        y = result[256][2].mean(axis=0)
+        y_err = 1.96 * result[256][2].std(axis=0)
+    else:
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
+        y = baseline_curves.mean(axis=0)
+        y_err = 1.96 * baseline_curves.std(axis=0)
+
+    axes[0].plot(x, y, linestyle=ls)
+    axes[0].fill_between(x, y + y_err, y - y_err, alpha=0.5)
+
+axes[0].set_xscale('log')
+axes[0].set_ylim(-28, -25)
+
+
+k = np.array([64, 128, 256, 512])
+
+axes[1].set_xlabel(r'Mixture Components $K$')
+axes[1].set_ylabel(r'Time (minutes)')
+axes[1].set_xticks(k)
+
+for l, ls, result in zip(labels, linestyles, table.values()):
+    t = np.array([[r[1].mean(), r[1].std()] for r in result.values()])
+    print(t.shape)
+    axes[1].errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l, linestyle=ls)
+
+axes[1].legend(loc='lower left', bbox_to_anchor=(1.01, 0))
+axes[1].set_ylim(-1, 430)
+
+fig.tight_layout()
+
+fig.savefig('figures/learning.pdf')
+
+plt.show()

### 2019-09-25 aistats_paper/figures/ra_dec_projection.py

@@ -1,0 +1,46 @@
+import numpy as np
+import torch
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+import corner
+
+sns.set()
+sns.set_context('paper')
+
+N = 100000
+
+weights, means, covars = torch.load(
+    '../results/em_256_507993_params.pkl',
+    map_location=torch.device('cpu')
+)
+
+X = torch.distributions.MultivariateNormal(
+    loc=means, covariance_matrix=covars
+).sample((N,)).numpy()
+
+j = torch.distributions.Categorical(
+    probs=weights.squeeze()
+).sample((N,)).numpy()
+
+X = X[np.arange(N), j, :]
+
+fig, ax = plt.subplots(figsize=(4, 2))
+
+figure = corner.hist2d(X[:, 0], X[:, 1], bins=100)
+
+ax.set_xlim(0, 360)
+ax.set_xticks(np.arange(0, 361, 60))
+
+ax.set_ylim(-90, 90)
+ax.set_yticks(np.arange(-90, 91, 30))
+
+ax.set_xlabel(r'Right Ascension ($\circ$)')
+ax.set_ylabel(r'Declination ($\circ$)')
+
+fig.tight_layout()
+
+fig.savefig('figures/density.pdf')
+
+plt.show()

### 2019-09-26 aistats_paper/sections/experiments.tex
More WIP

@@
 We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-To evaluate each method, we used a random sample of rows from the Gaia DR2 source table~\cite{brownGaiaDataRelease2018}.
+Initial experiements were conducted using synthetic data sampled from a known mixture of Gaussians.
+
+\subsection{Synthetic Data}
+
+\todo{Description of synthetic setup}
+
+\todo{Table of NLL results?}
+
+\todo{Plot: Example 2D solution}
+
+\todo{Plot: Obvious local maxima}
+
+\todo{Plot: Denegenerate minibatch EM solution}
+
+\subsection{Gaia Catalogue}
+
+To evaluate each method with a real application, we used a random sample of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
 We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
@@
 This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-We used a range of mixture component sizes $K$.
-In practice we would want to select a value of $K$ by cross-validation.
 
@@
 Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
+
+\subsubsection{Variable K}
 
@@
 \end{figure*}
+
+\subsubsection{Variable N}

### 2019-09-26 aistats_paper/sections/methods.tex
More WIP

@@
 
+Here we present an alternative update for $V_j$ that is less prone to numerical instability, and show that it is equivalent to Equation~\ref{eqn:mstep}.
+For clarity we drop the component indicator $j$ from the parameters, and add indicators $t$ and $t-1$ to distinguish between current and previous estimates of parameters.
+
+First, we define an adjustment operation,
+\begin{align}
+  \begin{split}
+  \text{adjust}(V, s, \bc, \bd) =& \ sV + \frac{1}{2}(\bc - \bd)(\bc + \bd)^T \\
+  &+ \frac{1}{2}(\bc + \bd)(\bc - \bd)^T  \label{eq:recentre1}
+  \end{split} \\
+  =& \ s(V + \bc\bc^T) - \bd\bd^T,\label{eq:recentre2}
+\end{align}
+which can be thought of as recentering a scaled variance around a new mean.
+Equation~\ref{eq:recentre1} is how we actually compute the adjustment, to minimise taking small differences between large values, whilst Equation~\ref{eq:recentre2} shows the identity we are interested in.
+
+In the M-step at iteration $t$ of our minibatch EM approach, we compute estimates of $\hat{q}_t$, $\alpha_t$ and $\bm_t$ as before using Equations~\ref{eq:sums} and~\ref{eqn:mstep}.
+We also compute minibatch-specific parameters using exact sums over the minibatch:
+\begin{align}
+q_b &= \sum_i^M r_i, \\
+\bm_b &= \frac{\sum_i^M r_i \bx_i}{q_b},\\
+V_b &= \frac{\sum_i^M r_i[(\bx_i - \bb_i)(\bx_i - \bb_i)^T  + B_i]}{q_b}
+\end{align}
+We then compute our new estimate of the variance $V_t$ as a function of the previous estimates $\{\hat{q}_{t-1}, \bm_{t-1}, V_{t-1} \}$, the minibatch values $\{q_b, \bm_b, V_b\}$, and the current estimates $\{\hat{q}_t,\bm_t \}$:
+\begin{align}
+\begin{split}
+V_{t} &= (1 - \lambda)\,\text{adjust}(V_{t-1}, \frac{\hat{q}_{t-1}}{\hat{q}_t}, \bm_{t-1}, \bm_t) \\
+&+ \lambda\, \text{adjust}(V_{b}, \frac{q_b}{\hat{q}_t}, \bm_{b}, \bm_{t}) \label{eq:update-computed}
+\end{split} \\
+&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
+&= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
+&= \frac{(1- \lambda)\hat{S}_{t-1} + \lambda S_t}{\hat{q}_t} - \bm_t\bm_t^T \\
+&= \frac{\hat{S}_t}{\hat{q}_t} - \bm_t\bm_t^T \label{eq:update-equiv}
+\end{align}
+Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.
+Whilst we found this update worked better in practice than a direct implementation, numerical instability is still possible if the standard deviations of the components are small enough relative to the means, and further work is needed to determine if a more stable update can be performed.
+
 \subsection{Stochastic Gradient Descent}
@@
 Unfortunately SGD updates are also $O(D^3)$ for the extreme deconvolution model, as we need to form the covariance $T_{ij}$ for each datapoint.
+
+\subsection{Stochastic Split-Merge}
+
+\todo{Add description of proposed stochastic split-merge}

### 2019-09-27 aistats_paper/sections/experiments.tex
More updates.

@@
 We implemented both minibatch approaches in PyTorch, and compared against the reference implementation from \citet{bovyExtremeDeconvolutionInferring2011}.
-Initial experiements were conducted using synthetic data sampled from a known mixture of Gaussians.
+Initial experiments were conducted using synthetic data sampled from a known mixture of Gaussians.
+We then evaluated the different methods on a real application using a dataset constructed from the Gaia catalogue.
+
+The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
+While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes and dataset sizes illustrate how the methods scale.
+Full details of our experimental setup for reproducibility are available in the supplementary material.
 
@@
 
-To evaluate each method with a real application, we used a random sample of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
+To evaluate each method with a real application, we used random samples of rows from the Gaia DR2 source table~\citep{brownGaiaDataRelease2018}.
 We selected the 5 primary astrometric features, along with the BP-RP colour and mean magnitude in the G-band.
-In total there were 2 million rows.
+We set the projection $R_i$ to the identity matrix for every sample.
 Where data were missing, we set the field to zero and the noise variance to a large value.
-We set the projection $R_i$ to the identity matrix for every sample.
-This preliminary study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
-
-The existing EM method ran on CPU, whilst the minibatch EM and SGD methods ran on GPU\@.
-While the absolute times depend strongly on hardware and fine implementation details, they give a sense of the sort of times possible on current workstations, and the relative times across model sizes illustrate how the methods scale.
-We used a validation set comprising $10\%$ of the rows when developing our experiments.
-Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
+This study uses only a small fraction of the full dataset size, but this allows us to fit the training data into memory, a requirement for use with the original implementation of extreme deconvolution.
 
@@
 
+In our initial experiments we evaluated the effect of varying the number of mixture components $K$.
+A fixed subsample of $2,000,000$ rows was used.
+We used a validation set comprising $10\%$ of the rows when developing our experiments.
+Final model performance was evaluated on a different held-out test set also comprising 10\% of the rows at the last stage, with no parameter selection or development done based on this set.
 Table~\ref{results-table} reports the validation and test log-likelihoods for each method.
@@
 \subsubsection{Variable N}
+
+\todo{Description of setup.}
+
+\todo{Table of results for different values of N}
+
+\todo{Plot of validation LL against cost-scaled epochs}
+
+\todo{Plot of training time against N}

### 2019-09-27 aistats_paper/sections/methods.tex
More updates.

@@
 \end{align}
+\todo{Expand out where $\bb_{ij}$ and $B_{ij}$ come from, possibly in background?.}
 The $r_{ij}$ term is the posterior probability of datapoint $\bx_i$ coming from component $j$.
@@
 \end{split} \\
-&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] + \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right] \\
+\begin{split}
+&= (1 - \lambda) \left[\frac{\hat{q}_{t-1}}{\hat{q}_t} \left(V_{t-1} + \bm_{t-1}\bm_{t-1}^T \right) -\bm_t \bm_t^T \right] \\
+&+ \lambda\,\left[\frac{q_b}{\hat{q}_t} \left(V_{b} + \bm_{b}\bm_{b}^T \right) -\bm_t \bm_t^T \right]
+\end{split} \\
 &= (1 - \lambda)\,\left[\frac{\hat{S}_{t-1}}{\hat{q}_t} -\bm_t \bm_t^T \right] + \lambda\, \left[\frac{S_t}{\hat{q}_t} -\bm_t \bm_t^T \right] \\
@@
 \end{align}
+\todo{Fix typesetting here.}
 Again, Equation~\ref{eq:update-computed} is how we actually compute the update to minimise numerical errors, whilst Equation~\ref{eq:update-equiv} shows that the update is equivalent to the covariance update defined in Equation~\ref{eqn:mstep}.

### 2019-10-20 ml4ps_paper/paper.tex
Change to camera-ready and edits per reviewer comments.

@@
 % to compile a camera-ready version, add the [final] option, e.g.:
-\usepackage{neurips_2019}
+\usepackage[final]{neurips_2019_ml4ps}
 
@@
 \begin{equation}
-p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
+p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K,
 \end{equation}
+parameterised by mixture weight $\alpha_j$, mean $\bm_j$ and covariance $V_j$.
 As the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture.
@@
   \includegraphics[width=\textwidth]{figures/learning.pdf}
-  \caption{\textbf{Left}: Average training log-likelihood as a function of training epoch for models with $K=256$.
+  \caption{\textbf{Left}: Average training log-likelihood as a function of training on the Gaia subset for models with $K=256$.
 Epochs rescaled by average training time.
@@
 Both minibatch methods will allow us to fit larger models going forwards.
+
+\subsubsection*{Acknowledgments}
+We are grateful to David W. Hogg for providing us with the problem and advising us on our experimental setup.
+Our experiments made use of AstroPy~\citep{astropy:2013,astropy:2018}, corner.py~\citep{corner}, Numpy~\citep{numpy}, Pandas~\citep{mckinney-proc-scipy-2010}, PyTorch~\citep{paszke2017automatic} and Scikit-learn~\citep{scikit-learn}.
+This work was supported in part by the EPSRC Centre for Doctoral Training in Data Science, funded by the UK Engineering and Physical Sciences Research Council (grant EP/L016427/1) and the University of Edinburgh.
 

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_baseline.sh
Variable N fixes.

@@ -1,0 +1,19 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --mem=14000  # memory in Mb
+#SBATCH --cpus-per-task=8
+#SBATCH --time=0-20:00:00
+#SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/baseline_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+
+export PYTHONPATH=${PYTHONPATH}:${HOME}/original_xd/py/
+
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 64 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_64_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 128 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_128_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 256 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_baseline.py -c 512 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_em.sh

@@ -1,0 +1,18 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --gres=gpu:1
+#SBATCH --mem=8000  # memory in Mb
+#SBATCH --cpus-per-task=8
+#SBATCH --time=0-10:00:00
+#SBATCH -o output/em-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/em_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 64 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_64_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 128 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_128_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 256 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_em.py -c 512 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_gaia_baseline.py

@@ -1,0 +1,103 @@
+import argparse
+import json
+import time
+
+import numpy as np
+import torch
+import torch.utils.data as data_utils
+
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
+from deconv.gmm.util import minibatch_k_means
+
+from extreme_deconvolution import extreme_deconvolution
+
+
+def fit_gaia_baseline(datafile, output_prefix, K, epochs, w_reg,
+                      k_means_iters):
+    data = np.load(datafile)
+
+    train_data = SGDDeconvDataset(
+        torch.Tensor(data['X_train']),
+        torch.Tensor(data['C_train'])
+    )
+
+    loader = data_utils.DataLoader(
+        train_data,
+        batch_size=5000,
+        num_workers=4,
+        shuffle=True
+    )
+
+    start_time = time.time()
+    counts, centroids = minibatch_k_means(loader, k=K, max_iters=10)
+
+    weights = (counts / counts.sum()).numpy()
+    means = centroids.numpy()
+    covars = np.array(K * [np.eye(7)])
+
+    ll = extreme_deconvolution(
+        data['X_train'],
+        data['C_train'],
+        weights,
+        means,
+        covars,
+        w=w_reg,
+        logfile=str(output_prefix) + '_log',
+        maxiter=epochs
+    )
+
+    end_time = time.time()
+
+    train_score = ll * data['X_train'].shape[0]
+
+    val_ll = extreme_deconvolution(
+        data['X_val'],
+        data['C_val'],
+        weights,
+        means,
+        covars,
+        w=w_reg,
+        likeonly=True
+    )
+    val_score = val_ll * data['X_val'].shape[0]
+
+    print('Training score: {}'.format(train_score))
+    print('Val score: {}'.format(val_score))
+
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+    }
+
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
+    np.savez(
+        output_prefix + '_params.npz',
+        weights=weights,
+        means=means,
+        covar=covars
+    )
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-w', '--w_reg', type=float)
[…]

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_gaia_lim_em.py

@@ -1,0 +1,101 @@
+import argparse
+import json
+import time
+
+import numpy as np
+import torch
+
+from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
+from deconv.gmm.data import DeconvDataset
+
+
+def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w_reg,
+                    k_means_iters, lr_step, lr_gamma, use_cuda):
+    data = np.load(datafile)
+
+    if use_cuda:
+        device = torch.device('cuda')
+    else:
+        device = torch.device('cpu')
+
+    train_data = DeconvDataset(
+        torch.Tensor(data['X_train']),
+        torch.Tensor(data['C_train'])
+    )
+
+    val_data = DeconvDataset(
+        torch.Tensor(data['X_val']),
+        torch.Tensor(data['C_val'])
+    )
+
+    gmm = OnlineDeconvGMM(
+        K,
+        7,
+        device=device,
+        batch_size=batch_size,
+        restarts=1,
+        w=w_reg,
+        step_size=step_size,
+        epochs=epochs,
+        k_means_iters=k_means_iters,
+        lr_step=lr_step,
+        lr_gamma=lr_gamma
+    )
+
+    start_time = time.time()
+    gmm.fit(train_data, val_data=val_data, verbose=True)
+    end_time = time.time()
+
+    train_score = gmm.score_batch(train_data)
+    val_score = gmm.score_batch(val_data)
+
+    print('Training score: {}'.format(train_score))
+    print('Val score: {}'.format(val_score))
+
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+        'train_curve': gmm.train_ll_curve,
+        'val_curve': gmm.val_ll_curve
+    }
+
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
+    torch.save(
+        (gmm.weights, gmm.means, gmm.covars),
+        output_prefix + '_params.pkl'
+    )
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-b', '--batch-size', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-s', '--step-size', type=float)
+    parser.add_argument('-w', '--w_reg', type=float)
[…]

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_gaia_lim_sgd.py

@@ -1,0 +1,96 @@
+import argparse
+import json
+import time
+
+import numpy as np
+import torch
+
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM
+from deconv.gmm.data import DeconvDataset
+
+
+def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
+                     w_reg, k_means_iters, lr_step, lr_gamma,
+                     use_cuda):
+    data = np.load(datafile)
+
+    if use_cuda:
+        device = torch.device('cuda')
+    else:
+        device = torch.device('cpu')
+
+    train_data = DeconvDataset(
+        torch.Tensor(data['X_train']),
+        torch.Tensor(data['C_train'])
+    )
+
+    val_data = DeconvDataset(
+        torch.Tensor(data['X_val']),
+        torch.Tensor(data['C_val'])
+    )
+
+    gmm = SGDDeconvGMM(
+        K,
+        7,
+        device=device,
+        batch_size=batch_size,
+        epochs=epochs,
+        w=w_reg,
+        restarts=1,
+        k_means_iters=k_means_iters,
+        lr=lr,
+        lr_step=lr_step,
+        lr_gamma=lr_gamma
+    )
+    start_time = time.time()
+    gmm.fit(train_data, val_data=val_data, verbose=True)
+    end_time = time.time()
+
+    train_score = gmm.score_batch(train_data)
+    val_score = gmm.score_batch(val_data)
+
+    print('Training score: {}'.format(train_score))
+    print('Val score: {}'.format(val_score))
+
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+        'train_curve': gmm.train_loss_curve,
+        'val_curve': gmm.val_loss_curve
+    }
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
+    torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-b', '--batch-size', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-l', '--learning-rate', type=float)
+    parser.add_argument('-w', '--w_reg', type=float)
+    parser.add_argument('-k', '--k-means-iters', type=int)
[…]

### 2019-10-30 deconv/experiments/gaia/variable_k/fit_sgd.sh

@@ -1,0 +1,18 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --gres=gpu:1
+#SBATCH --mem=8000  # memory in Mb
+#SBATCH --cpus-per-task=8
+#SBATCH --time=0-05:00:00
+#SBATCH -o output/sgd-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/sgd_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+
+python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 64 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_64_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 128 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_128_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 256 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_256_${SLURM_JOBID}
+python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 512 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/variable_k/score_test_em.py

@@ -1,0 +1,66 @@
+import argparse
+import json
+import os
+
+import numpy as np
+import torch
+
+from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
+from deconv.gmm.data import DeconvDataset
+
+
+def score_em(datafile, results_dir, output_file):
+    data = np.load(datafile)
+
+    test_data = SGDDeconvDataset(
+        torch.Tensor(data['X_test']),
+        torch.Tensor(data['C_test'])
+    )
+
+    rf = os.listdir(results_dir)
+
+    param_files = [
+        f for f in rf if f.startswith('em_512') and f.endswith('.pkl')
+    ]
+
+    gmm = OnlineDeconvGMM(
+        512,
+        7,
+        batch_size=500
+    )
+
+    scores = []
+
+    for p in param_files:
+        weights, means, covars = torch.load(
+            results_dir + p,
+            map_location=torch.device('cpu')
+        )
+        gmm.weights = weights
+        gmm.means = means
+        gmm.covars = covars
+
+        test_score = gmm.score_batch(test_data)
+        print(test_score)
+
+        scores.append(test_score)
+
+    print('Test Score: {} +- {}'.format(
+        np.mean(scores),
+        np.std(scores)
+    ))
+
+    json.dump(scores, open(output_file, 'w'))
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('datafile')
+    parser.add_argument('results_dir')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    score_em(args.datafile, args.results_dir, args.output_file)

### 2019-10-30 deconv/experiments/gaia/variable_k/score_test_sgd.py

@@ -1,0 +1,64 @@
+import argparse
+import json
+import os
+
+import numpy as np
+import torch
+
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM
+from decon.gmm.data import DeconvDataset
+
+
+def score_sgd(datafile, results_dir, output_file):
+    data = np.load(datafile)
+
+    test_data = SGDDeconvDataset(
+        torch.Tensor(data['X_test']),
+        torch.Tensor(data['C_test'])
+    )
+
+    rf = os.listdir(results_dir)
+
+    param_files = [
+        f for f in rf if f.startswith('sgd_512') and f.endswith('.pkl')
+    ]
+
+    gmm = SGDDeconvGMM(
+        512,
+        7,
+        batch_size=500
+    )
+
+    scores = []
+
+    for p in param_files:
+        state_dict = torch.load(
+            results_dir + p,
+            map_location=torch.device('cpu')
+        )
+        gmm.module.load_state_dict(state_dict)
+
+        test_score = gmm.score_batch(test_data)
+        print(test_score)
+
+        scores.append(test_score)
+
+    print('Test Score: {} +- {}'.format(
+        np.mean(scores),
+        np.std(scores)
+    ))
+
+    json.dump(scores, open(output_file, 'w'))
+
+
+if __name__ == '__main__':
+
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('datafile')
+    parser.add_argument('results_dir')
+    parser.add_argument('output_file')
+
+    args = parser.parse_args()
+
+    score_sgd(args.datafile, args.results_dir, args.output_file)

### 2019-10-30 deconv/experiments/gaia/variable_n/fit_gaia_sgd.py

@@ -1,0 +1,89 @@
+import argparse
+import json
+import time
+
+import numpy as np
+import torch
+
+from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM
+from deconv.gmm.data import H5DeconvDataset
+
+
+def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
+                     w_reg, k_means_iters, lr_step, lr_gamma, train_limit,
+                     use_cuda):
+    if use_cuda:
+        device = torch.device('cuda')
+    else:
+        device = torch.device('cpu')
+
+    train_data = H5DeconvDataset(datafile, 'train', limit=train_limit)
+    val_data = H5DeconvDataset(datafile, 'val', limit=25e6)
+
+    gmm = SGDDeconvGMM(
+        K,
+        7,
+        device=device,
+        batch_size=batch_size,
+        epochs=epochs,
+        w=w_reg,
+        restarts=1,
+        k_means_iters=k_means_iters,
+        lr=lr,
+        lr_step=lr_step,
+        lr_gamma=lr_gamma
+    )
+    start_time = time.time()
+    gmm.fit(train_data, val_data=val_data, verbose=True)
+    end_time = time.time()
+
+    train_score = gmm.score_batch(train_data)
+    val_score = gmm.score_batch(val_data)
+
+    print('Training score: {}'.format(train_score))
+    print('Val score: {}'.format(val_score))
+
+    results = {
+        'start_time': start_time,
+        'end_time': end_time,
+        'train_score': train_score,
+        'val_score': val_score,
+        'train_curve': gmm.train_loss_curve,
+        'val_curve': gmm.val_loss_curve
+    }
+    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
+    torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser()
+
+    parser.add_argument('-c', '--components', type=int)
+    parser.add_argument('-b', '--batch-size', type=int)
+    parser.add_argument('-e', '--epochs', type=int)
+    parser.add_argument('-l', '--learning-rate', type=float)
+    parser.add_argument('-w', '--w_reg', type=float)
+    parser.add_argument('-k', '--k-means-iters', type=int)
+    parser.add_argument('--lr-step', type=int)
+    parser.add_argument('--lr-gamma', type=float)
[…]

### 2019-10-30 deconv/experiments/gaia/variable_n/fit_sgd.sh

@@ -1,0 +1,15 @@
+#!/bin/sh
+#SBATCH -N 1      # nodes requested
+#SBATCH -n 1      # tasks requested
+#SBATCH --gres=gpu:1
+#SBATCH --mem=8000  # memory in Mb
+#SBATCH --cpus-per-task=12
+#SBATCH --time=2-00:00:00
+#SBATCH -o output/sgd-%A_%a.txt  # send stdout to outfile
+#SBATCH -e output/sgd_error-%A_%a.txt  # send stderr to errfile
+#SBATCH --partition=apollo
+
+source ~/.bashrc
+conda activate deconv
+
+python deconv/experiments/gaia/variable_n/fit_gaia_sgd.py -c 512 -b 500 -e 10 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --train-limit 25000000 --use-cuda /disk/scratch/s0904254/gaia_source.h5 results/variable_n/sgd_512_25_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/variable_n/preproccesing.py

@@ -1,0 +1,155 @@
+import argparse
+import os
+
+import h5py
+import numpy as np
+import pandas as pd
+
+from astropy.table import Table
+from sklearn.model_selection import train_test_split
+import tqdm
+
+np.random.seed(90115)
+
+columns = [
+    'ra',
+    'dec',
+    'parallax',
+    'pmra',
+    'pmdec',
+    'bp_rp',
+    'phot_g_mean_mag'
+]
+
+error_columns = [
+    'ra_error',
+    'dec_error',
+    'parallax_error',
+    'pmra_error',
+    'pmdec_error',
+]
+
+corr_map = {
+    'ra_dec_corr': [0, 1],
+    'ra_parallax_corr': [0, 2],
+    'ra_pmra_corr': [0, 3],
+    'ra_pmdec_corr': [0, 4],
+    'dec_parallax_corr': [1, 2],
+    'dec_pmra_corr': [1, 3],
+    'dec_pmdec_corr': [1, 4],
+    'parallax_pmra_corr': [2, 3],
+    'parallax_pmdec_corr': [2, 4],
+    'pmra_pmdec_corr': [3, 4]
+}
+
+
+def get_covar(row):
+    return np.diag(row[error_columns].fillna(1e12).to_numpy(dtype=np.float32))
+
+
+def vot_to_pandas(vot_file):
+
+    tb = Table.read(vot_file)
+
+    df = tb[columns + error_columns + list(corr_map.keys())].to_pandas()
+    return df
+
+
+def pandas_to_numpy(df):
+    df.insert(12, column='phot_g_mean_mag_error', value=0.01)
+    df.insert(12, column='bp_rp_error', value=0.01)
+
+    ec = error_columns.copy()
+    ec.insert(5, 'phot_g_mean_mag_error')
+    ec.insert(5, 'bp_rp_error')
+
+    X = df[columns].fillna(0.0).to_numpy(dtype=np.float32)
+    C = np.zeros((len(df), 7, 7), dtype=np.float32)
+    diag = np.arange(7)
+    C[:, diag, diag] = df[ec].fillna(1e6).to_numpy(
+        dtype=np.float32
+    )
+
+    for column, (i, j) in corr_map.items():
+        C[:, i, j] = df[column].fillna(0).to_numpy(dtype=np.float32)
+        C[:, i, j] *= (C[:, i, i] * C[:, j, j])
+        C[:, j, i] = C[:, i, j]
+
[…]

### 2019-10-30 deconv/gmm/data.py

@@ -3,6 +3,7 @@ import torch.utils.data as data_utils
 
 import h5py
 
+
 class DeconvDataset(data_utils.Dataset):
 
     def __init__(self, X, noise_covars):
@@ -18,16 +19,25 @@ class DeconvDataset(data_utils.Dataset):
 
 class H5DeconvDataset(data_utils.Dataset):
 
-    def __init__(self, filepath, key):
+    def __init__(self, filepath, key, limit=None):
 
-        self.store = h5py.File(filepath, 'r')
-        self.group = self.store[key]
+        self.filepath = filepath
+        self.key = key
+        self.limit = limit
+        self.group = None
 
     def __len__(self):
-        return self.group['X'].shape[0]
+        if self.limit:
+            return self.limit
+        else:
+            store = h5py.File(self.filepath, 'r')
+            return store[self.key]['X'].shape[0]
 
     def __getitem__(self, i):
+        if self.group is None:
+            store = h5py.File(self.filepath, 'r')
+            self.group = store[self.key]
         return (
-            self.group['X'][i, :],
-            self.group['C'][i, :, :]
+            torch.from_numpy(self.group['X'][i, :]),
+            torch.from_numpy(self.group['C'][i, :, :])
         )

### 2019-10-30 deconv/gmm/sgd_gmm.py

@@ -116,7 +116,7 @@ class BaseSGDGMM(ABC):
         init_loader = data_utils.DataLoader(
             data,
             batch_size=self.batch_size * self.k_means_factor,
-            num_workers=4,
+            num_workers=8,
             shuffle=True,
             pin_memory=True
         )
@@ -124,7 +124,7 @@ class BaseSGDGMM(ABC):
         loader = data_utils.DataLoader(
             data,
             batch_size=self.batch_size,
-            num_workers=4,
+            num_workers=8,
             shuffle=True,
             pin_memory=True
         )

### 2019-10-30 deconv/experiments/gaia/fit_baseline.sh
More changes.

@@ -1,19 +1,0 @@
-#!/bin/sh
-#SBATCH -N 1      # nodes requested
-#SBATCH -n 1      # tasks requested
-#SBATCH --mem=14000  # memory in Mb
-#SBATCH --cpus-per-task=8
-#SBATCH --time=0-20:00:00
-#SBATCH -o output/baseline-%A_%a.txt  # send stdout to outfile
-#SBATCH -e output/baseline_error-%A_%a.txt  # send stderr to errfile
-#SBATCH --partition=apollo
-
-source ~/.bashrc
-conda activate deconv
-
-export PYTHONPATH=${PYTHONPATH}:${HOME}/original_xd/py/
-
-python deconv/experiments/gaia/fit_gaia_baseline.py -c 64 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_64_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_baseline.py -c 128 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_128_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_baseline.py -c 256 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_256_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_baseline.py -c 512 -e 20 -w=0.001 -k 10 data/gaia_sample_mag.npz results/baseline_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/fit_em.sh

@@ -1,18 +1,0 @@
-#!/bin/sh
-#SBATCH -N 1      # nodes requested
-#SBATCH -n 1      # tasks requested
-#SBATCH --gres=gpu:1
-#SBATCH --mem=8000  # memory in Mb
-#SBATCH --cpus-per-task=8
-#SBATCH --time=0-10:00:00
-#SBATCH -o output/em-%A_%a.txt  # send stdout to outfile
-#SBATCH -e output/em_error-%A_%a.txt  # send stderr to errfile
-#SBATCH --partition=apollo
-
-source ~/.bashrc
-conda activate deconv
-
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 64 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_64_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 128 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_128_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 256 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_256_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_em.py -c 512 -b 500 -e 20 -s 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.5 --use-cuda data/gaia_sample_mag.npz results/em_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/fit_gaia_baseline.py

@@ -1,103 +1,0 @@
-import argparse
-import json
-import time
-
-import numpy as np
-import torch
-import torch.utils.data as data_utils
-
-from deconv.gmm.sgd_deconv_gmm import SGDDeconvDataset
-from deconv.gmm.util import minibatch_k_means
-
-from extreme_deconvolution import extreme_deconvolution
-
-
-def fit_gaia_baseline(datafile, output_prefix, K, epochs, w_reg,
-                      k_means_iters):
-    data = np.load(datafile)
-
-    train_data = SGDDeconvDataset(
-        torch.Tensor(data['X_train']),
-        torch.Tensor(data['C_train'])
-    )
-
-    loader = data_utils.DataLoader(
-        train_data,
-        batch_size=5000,
-        num_workers=4,
-        shuffle=True
-    )
-
-    start_time = time.time()
-    counts, centroids = minibatch_k_means(loader, k=K, max_iters=10)
-
-    weights = (counts / counts.sum()).numpy()
-    means = centroids.numpy()
-    covars = np.array(K * [np.eye(7)])
-
-    ll = extreme_deconvolution(
-        data['X_train'],
-        data['C_train'],
-        weights,
-        means,
-        covars,
-        w=w_reg,
-        logfile=str(output_prefix) + '_log',
-        maxiter=epochs
-    )
-
-    end_time = time.time()
-
-    train_score = ll * data['X_train'].shape[0]
-
-    val_ll = extreme_deconvolution(
-        data['X_val'],
-        data['C_val'],
-        weights,
-        means,
-        covars,
-        w=w_reg,
-        likeonly=True
-    )
-    val_score = val_ll * data['X_val'].shape[0]
-
-    print('Training score: {}'.format(train_score))
-    print('Val score: {}'.format(val_score))
-
-    results = {
-        'start_time': start_time,
-        'end_time': end_time,
-        'train_score': train_score,
-        'val_score': val_score,
-    }
-
-    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
-    np.savez(
-        output_prefix + '_params.npz',
-        weights=weights,
-        means=means,
-        covar=covars
-    )
-
-
-if __name__ == '__main__':
-
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('-c', '--components', type=int)
-    parser.add_argument('-e', '--epochs', type=int)
-    parser.add_argument('-w', '--w_reg', type=float)
[…]

### 2019-10-30 deconv/experiments/gaia/fit_gaia_lim_em.py

@@ -1,101 +1,0 @@
-import argparse
-import json
-import time
-
-import numpy as np
-import torch
-
-from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
-from deconv.gmm.data import DeconvDataset
-
-
-def fit_gaia_lim_em(datafile, output_prefix, K, batch_size, epochs, step_size, w_reg,
-                    k_means_iters, lr_step, lr_gamma, use_cuda):
-    data = np.load(datafile)
-
-    if use_cuda:
-        device = torch.device('cuda')
-    else:
-        device = torch.device('cpu')
-
-    train_data = DeconvDataset(
-        torch.Tensor(data['X_train']),
-        torch.Tensor(data['C_train'])
-    )
-
-    val_data = DeconvDataset(
-        torch.Tensor(data['X_val']),
-        torch.Tensor(data['C_val'])
-    )
-
-    gmm = OnlineDeconvGMM(
-        K,
-        7,
-        device=device,
-        batch_size=batch_size,
-        restarts=1,
-        w=w_reg,
-        step_size=step_size,
-        epochs=epochs,
-        k_means_iters=k_means_iters,
-        lr_step=lr_step,
-        lr_gamma=lr_gamma
-    )
-
-    start_time = time.time()
-    gmm.fit(train_data, val_data=val_data, verbose=True)
-    end_time = time.time()
-
-    train_score = gmm.score_batch(train_data)
-    val_score = gmm.score_batch(val_data)
-
-    print('Training score: {}'.format(train_score))
-    print('Val score: {}'.format(val_score))
-
-    results = {
-        'start_time': start_time,
-        'end_time': end_time,
-        'train_score': train_score,
-        'val_score': val_score,
-        'train_curve': gmm.train_ll_curve,
-        'val_curve': gmm.val_ll_curve
-    }
-
-    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
-    torch.save(
-        (gmm.weights, gmm.means, gmm.covars),
-        output_prefix + '_params.pkl'
-    )
-
-if __name__ == '__main__':
-
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('-c', '--components', type=int)
-    parser.add_argument('-b', '--batch-size', type=int)
-    parser.add_argument('-e', '--epochs', type=int)
-    parser.add_argument('-s', '--step-size', type=float)
-    parser.add_argument('-w', '--w_reg', type=float)
[…]

### 2019-10-30 deconv/experiments/gaia/fit_gaia_lim_sgd.py

@@ -1,96 +1,0 @@
-import argparse
-import json
-import time
-
-import numpy as np
-import torch
-
-from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM
-from deconv.gmm.data import DeconvDataset
-
-
-def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
-                     w_reg, k_means_iters, lr_step, lr_gamma,
-                     use_cuda):
-    data = np.load(datafile)
-
-    if use_cuda:
-        device = torch.device('cuda')
-    else:
-        device = torch.device('cpu')
-
-    train_data = DeconvDataset(
-        torch.Tensor(data['X_train']),
-        torch.Tensor(data['C_train'])
-    )
-
-    val_data = DeconvDataset(
-        torch.Tensor(data['X_val']),
-        torch.Tensor(data['C_val'])
-    )
-
-    gmm = SGDDeconvGMM(
-        K,
-        7,
-        device=device,
-        batch_size=batch_size,
-        epochs=epochs,
-        w=w_reg,
-        restarts=1,
-        k_means_iters=k_means_iters,
-        lr=lr,
-        lr_step=lr_step,
-        lr_gamma=lr_gamma
-    )
-    start_time = time.time()
-    gmm.fit(train_data, val_data=val_data, verbose=True)
-    end_time = time.time()
-
-    train_score = gmm.score_batch(train_data)
-    val_score = gmm.score_batch(val_data)
-
-    print('Training score: {}'.format(train_score))
-    print('Val score: {}'.format(val_score))
-
-    results = {
-        'start_time': start_time,
-        'end_time': end_time,
-        'train_score': train_score,
-        'val_score': val_score,
-        'train_curve': gmm.train_loss_curve,
-        'val_curve': gmm.val_loss_curve
-    }
-    json.dump(results, open(str(output_prefix) + '_results.json', mode='w'))
-    torch.save(gmm.module.state_dict(), output_prefix + '_params.pkl')
-
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('-c', '--components', type=int)
-    parser.add_argument('-b', '--batch-size', type=int)
-    parser.add_argument('-e', '--epochs', type=int)
-    parser.add_argument('-l', '--learning-rate', type=float)
-    parser.add_argument('-w', '--w_reg', type=float)
-    parser.add_argument('-k', '--k-means-iters', type=int)
[…]

### 2019-10-30 deconv/experiments/gaia/fit_sgd.sh

@@ -1,18 +1,0 @@
-#!/bin/sh
-#SBATCH -N 1      # nodes requested
-#SBATCH -n 1      # tasks requested
-#SBATCH --gres=gpu:1
-#SBATCH --mem=8000  # memory in Mb
-#SBATCH --cpus-per-task=8
-#SBATCH --time=0-05:00:00
-#SBATCH -o output/sgd-%A_%a.txt  # send stdout to outfile
-#SBATCH -e output/sgd_error-%A_%a.txt  # send stderr to errfile
-#SBATCH --partition=apollo
-
-source ~/.bashrc
-conda activate deconv
-
-python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 64 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_64_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 128 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_128_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 256 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_256_${SLURM_JOBID}
-python deconv/experiments/gaia/fit_gaia_lim_sgd.py -c 512 -b 500 -e 20 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --use-cuda data/gaia_sample_mag.npz results/sgd_512_${SLURM_JOBID}

### 2019-10-30 deconv/experiments/gaia/preproccesing.py

@@ -1,155 +1,0 @@
-import argparse
-import os
-
-import h5py
-import numpy as np
-import pandas as pd
-
-from astropy.table import Table
-from sklearn.model_selection import train_test_split
-import tqdm
-
-np.random.seed(90115)
-
-columns = [
-    'ra',
-    'dec',
-    'parallax',
-    'pmra',
-    'pmdec',
-    'bp_rp',
-    'phot_g_mean_mag'
-]
-
-error_columns = [
-    'ra_error',
-    'dec_error',
-    'parallax_error',
-    'pmra_error',
-    'pmdec_error',
-]
-
-corr_map = {
-    'ra_dec_corr': [0, 1],
-    'ra_parallax_corr': [0, 2],
-    'ra_pmra_corr': [0, 3],
-    'ra_pmdec_corr': [0, 4],
-    'dec_parallax_corr': [1, 2],
-    'dec_pmra_corr': [1, 3],
-    'dec_pmdec_corr': [1, 4],
-    'parallax_pmra_corr': [2, 3],
-    'parallax_pmdec_corr': [2, 4],
-    'pmra_pmdec_corr': [3, 4]
-}
-
-
-def get_covar(row):
-    return np.diag(row[error_columns].fillna(1e12).to_numpy(dtype=np.float32))
-
-
-def vot_to_pandas(vot_file):
-
-    tb = Table.read(vot_file)
-
-    df = tb[columns + error_columns + list(corr_map.keys())].to_pandas()
-    return df
-
-
-def pandas_to_numpy(df):
-    df.insert(12, column='phot_g_mean_mag_error', value=0.01)
-    df.insert(12, column='bp_rp_error', value=0.01)
-
-    ec = error_columns.copy()
-    ec.insert(5, 'phot_g_mean_mag_error')
-    ec.insert(5, 'bp_rp_error')
-
-    X = df[columns].fillna(0.0).to_numpy(dtype=np.float32)
-    C = np.zeros((len(df), 7, 7), dtype=np.float32)
-    diag = np.arange(7)
-    C[:, diag, diag] = df[ec].fillna(1e6).to_numpy(
-        dtype=np.float32
-    )
-
-    for column, (i, j) in corr_map.items():
-        C[:, i, j] = df[column].fillna(0).to_numpy(dtype=np.float32)
-        C[:, i, j] *= (C[:, i, i] * C[:, j, j])
-        C[:, j, i] = C[:, i, j]
-
[…]

### 2019-10-30 deconv/experiments/gaia/score_test_baseline.py

@@ -1,57 +1,0 @@
-import argparse
-import json
-import os
-
-import numpy as np
-
-from extreme_deconvolution import extreme_deconvolution
-
-
-def score_baseline(datafile, results_dir, output_file):
-    data = np.load(datafile)
-
-    rf = os.listdir(results_dir)
-
-    param_files = [
-        f for f in rf if f.startswith('baseline_512') and f.endswith('.npz')
-    ]
-
-    scores = []
-
-    for p in param_files:
-        params = np.load(results_dir + p)
-        weights = params['weights']
-        means = params['means']
-        covars = params['covar']
-
-        test_score = extreme_deconvolution(
-            data['X_test'],
-            data['C_test'],
-            weights,
-            means,
-            covars,
-            likeonly=True
-        )
-        print(test_score)
-
-        scores.append(test_score)
-
-    print('Test Score: {} +- {}'.format(
-        np.mean(scores),
-        np.std(scores)
-    ))
-
-    json.dump(scores, open(output_file, 'w'))
-
-
-if __name__ == '__main__':
-
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('datafile')
-    parser.add_argument('results_dir')
-    parser.add_argument('output_file')
-
-    args = parser.parse_args()
-
-    score_baseline(args.datafile, args.results_dir, args.output_file)

### 2019-10-30 deconv/experiments/gaia/score_test_em.py

@@ -1,66 +1,0 @@
-import argparse
-import json
-import os
-
-import numpy as np
-import torch
-
-from deconv.gmm.online_deconv_gmm import OnlineDeconvGMM
-from deconv.gmm.data import DeconvDataset
-
-
-def score_em(datafile, results_dir, output_file):
-    data = np.load(datafile)
-
-    test_data = SGDDeconvDataset(
-        torch.Tensor(data['X_test']),
-        torch.Tensor(data['C_test'])
-    )
-
-    rf = os.listdir(results_dir)
-
-    param_files = [
-        f for f in rf if f.startswith('em_512') and f.endswith('.pkl')
-    ]
-
-    gmm = OnlineDeconvGMM(
-        512,
-        7,
-        batch_size=500
-    )
-
-    scores = []
-
-    for p in param_files:
-        weights, means, covars = torch.load(
-            results_dir + p,
-            map_location=torch.device('cpu')
-        )
-        gmm.weights = weights
-        gmm.means = means
-        gmm.covars = covars
-
-        test_score = gmm.score_batch(test_data)
-        print(test_score)
-
-        scores.append(test_score)
-
-    print('Test Score: {} +- {}'.format(
-        np.mean(scores),
-        np.std(scores)
-    ))
-
-    json.dump(scores, open(output_file, 'w'))
-
-
-if __name__ == '__main__':
-
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('datafile')
-    parser.add_argument('results_dir')
-    parser.add_argument('output_file')
-
-    args = parser.parse_args()
-
-    score_em(args.datafile, args.results_dir, args.output_file)

### 2019-10-30 deconv/experiments/gaia/score_test_sgd.py

@@ -1,64 +1,0 @@
-import argparse
-import json
-import os
-
-import numpy as np
-import torch
-
-from deconv.gmm.sgd_deconv_gmm import SGDDeconvGMM
-from decon.gmm.data import DeconvDataset
-
-
-def score_sgd(datafile, results_dir, output_file):
-    data = np.load(datafile)
-
-    test_data = SGDDeconvDataset(
-        torch.Tensor(data['X_test']),
-        torch.Tensor(data['C_test'])
-    )
-
-    rf = os.listdir(results_dir)
-
-    param_files = [
-        f for f in rf if f.startswith('sgd_512') and f.endswith('.pkl')
-    ]
-
-    gmm = SGDDeconvGMM(
-        512,
-        7,
-        batch_size=500
-    )
-
-    scores = []
-
-    for p in param_files:
-        state_dict = torch.load(
-            results_dir + p,
-            map_location=torch.device('cpu')
-        )
-        gmm.module.load_state_dict(state_dict)
-
-        test_score = gmm.score_batch(test_data)
-        print(test_score)
-
-        scores.append(test_score)
-
-    print('Test Score: {} +- {}'.format(
-        np.mean(scores),
-        np.std(scores)
-    ))
-
-    json.dump(scores, open(output_file, 'w'))
-
-
-if __name__ == '__main__':
-
-    parser = argparse.ArgumentParser()
-
-    parser.add_argument('datafile')
-    parser.add_argument('results_dir')
-    parser.add_argument('output_file')
-
-    args = parser.parse_args()
-
-    score_sgd(args.datafile, args.results_dir, args.output_file)

### 2019-10-30 deconv/experiments/gaia/variable_n/fit_gaia_sgd.py

@@ -18,7 +18,7 @@ def fit_gaia_lim_sgd(datafile, output_prefix, K, batch_size, epochs, lr,
         device = torch.device('cpu')
 
     train_data = H5DeconvDataset(datafile, 'train', limit=train_limit)
-    val_data = H5DeconvDataset(datafile, 'val', limit=25e6)
+    val_data = H5DeconvDataset(datafile, 'val', limit=int(25e6))
 
     gmm = SGDDeconvGMM(
         K,

### 2019-10-30 deconv/experiments/gaia/variable_n/fit_sgd.sh

@@ -12,4 +12,4 @@
 source ~/.bashrc
 conda activate deconv
 
-python deconv/experiments/gaia/variable_n/fit_gaia_sgd.py -c 512 -b 500 -e 10 -l 0.01 -w=0.001 -k 10 --lr-step 10 --lr-gamma 0.1 --train-limit 25000000 --use-cuda /disk/scratch/s0904254/gaia_source.h5 results/variable_n/sgd_512_25_${SLURM_JOBID}
+python deconv/experiments/gaia/variable_n/fit_gaia_sgd.py -c 512 -b 500 -e 10 -l 0.01 -w=0.001 -k 1 --lr-step 10 --lr-gamma 0.1 --train-limit 25000000 --use-cuda /disk/scratch/s0904254/gaia_source.h5 results/variable_n/sgd_512_25_${SLURM_JOBID}

### 2019-10-30 deconv/gmm/data.py

@@ -38,6 +38,6 @@ class H5DeconvDataset(data_utils.Dataset):
             store = h5py.File(self.filepath, 'r')
             self.group = store[self.key]
         return (
-            torch.from_numpy(self.group['X'][i, :]),
-            torch.from_numpy(self.group['C'][i, :, :])
+            self.group['X'][i, :],
+            self.group['C'][i, :, :]
         )

### 2019-10-30 deconv/gmm/sgd_gmm.py

@@ -117,7 +117,7 @@ class BaseSGDGMM(ABC):
             data,
             batch_size=self.batch_size * self.k_means_factor,
             num_workers=8,
-            shuffle=True,
+            shuffle=False,
             pin_memory=True
         )
 

### 2019-10-30 deconv/experiments/gaia/preproccesing.py
Checkout preprocessing.

@@ -1,13 +1,10 @@
 import argparse
-import os
 
-import h5py
 import numpy as np
 import pandas as pd
 
 from astropy.table import Table
 from sklearn.model_selection import train_test_split
-import tqdm
 
 np.random.seed(90115)
 
@@ -55,18 +52,16 @@ def vot_to_pandas(vot_file):
     return df
 
 
-def pandas_to_numpy(df):
+def pandas_to_numpy(df, output_file):
     df.insert(12, column='phot_g_mean_mag_error', value=0.01)
     df.insert(12, column='bp_rp_error', value=0.01)
-
-    ec = error_columns.copy()
-    ec.insert(5, 'phot_g_mean_mag_error')
-    ec.insert(5, 'bp_rp_error')
+    error_columns.insert(5, 'phot_g_mean_mag_error')
+    error_columns.insert(5, 'bp_rp_error')
 
     X = df[columns].fillna(0.0).to_numpy(dtype=np.float32)
     C = np.zeros((len(df), 7, 7), dtype=np.float32)
     diag = np.arange(7)
-    C[:, diag, diag] = df[ec].fillna(1e6).to_numpy(
+    C[:, diag, diag] = df[error_columns].fillna(1e6).to_numpy(
         dtype=np.float32
     )
 
@@ -84,15 +79,6 @@ def pandas_to_numpy(df):
         X_test, C_test, test_size=0.5, random_state=84253
     )
 
-    return (
-        (X_train, C_train),
-        (X_val, C_val),
-        (X_test, C_test)
-    )
-
-
-def numpy_to_file(data, output_file):
-    (X_train, C_train), (X_val, C_val), (X_test, C_test) = data
     np.savez(
         output_file,
         X_train=X_train,
@@ -103,53 +89,13 @@ def numpy_to_file(data, output_file):
         C_test=C_test
     )
 
-
-def process_csv(csv_dir, output_file):
-    file_list = os.listdir(csv_dir)
-
-    store = h5py.File(output_file, 'w')
-
-    train_data = store.create_group('train')
-    val_data = store.create_group('val')
-    test_data = store.create_group('test')
-
-    groups = (train_data, val_data, test_data)
-
-    for g in groups:
-        g.create_dataset(
-            'X', (0, 7),
-            maxshape=(None, 7),
-            dtype=np.float32,
-            chunks=(512, 7),
-            compression='gzip'
-        )
-        g.create_dataset(
-            'C', (0, 7, 7),
[…]

### 2019-10-30 deconv/experiments/gaia/fit_gaia_baseline.py
Add some rough comments.

@@ -1,3 +1,4 @@
+"""Script to fit the baseline model"""
 import argparse
 import json
 import time

### 2019-10-30 deconv/experiments/gaia/fit_gaia_lim_em.py

@@ -1,3 +1,4 @@
+"""Script to fit the minibatch EM model."""
 import argparse
 import json
 import time

### 2019-10-30 deconv/experiments/gaia/preproccesing.py

@@ -1,7 +1,12 @@
+"""
+Script to preprocess data for the Gaia experiments.
+
+Converts a VOT file from a query on the Gaia DR2 source table
+into a Numpy .npz file for use with the experiments.
+"""
 import argparse
 
 import numpy as np
-import pandas as pd
 
 from astropy.table import Table
 from sklearn.model_selection import train_test_split
@@ -88,6 +93,7 @@ def pandas_to_numpy(df, output_file):
         X_test=X_test,
         C_test=C_test
     )
+
 
 if __name__ == '__main__':
 

### 2019-10-30 deconv/gmm/base.py

@@ -9,6 +9,7 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 
 class BaseGMM(ABC):
+    """ABC for GMMs fitted with EM-type methods."""
 
     def __init__(self, components, dimensions, epochs=100,
                  w=1e-6, tol=1e-9, restarts=5, device=None):

### 2019-10-30 deconv/gmm/data.py

@@ -1,9 +1,9 @@
 import torch
 import torch.utils.data as data_utils
 
-import h5py
 
 class DeconvDataset(data_utils.Dataset):
+    """Dataset with values and noise covariances."""
 
     def __init__(self, X, noise_covars):
         self.X = X
@@ -14,20 +14,3 @@ class DeconvDataset(data_utils.Dataset):
 
     def __getitem__(self, i):
         return (self.X[i, :], self.noise_covars[i, :, :])
-
-
-class H5DeconvDataset(data_utils.Dataset):
-
-    def __init__(self, filepath, key):
-
-        self.store = h5py.File(filepath, 'r')
-        self.group = self.store[key]
-
-    def __len__(self):
-        return self.group['X'].shape[0]
-
-    def __getitem__(self, i):
-        return (
-            self.group['X'][i, :],
-            self.group['C'][i, :, :]
-        )

### 2019-10-30 deconv/gmm/deconv_gmm.py

@@ -9,6 +9,7 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 
 class DeconvGMM(BaseGMM):
+    """Implementation of a deconvolving GMM fitted with standard EM."""
 
     def __init__(self, components, dimensions, epochs=1000,
                  w=1e-6, tol=1e-6, restarts=5, device=None):

### 2019-10-30 deconv/gmm/gmm.py

@@ -7,6 +7,7 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 
 class GMM(BaseGMM):
+    """Implementation of a standard GMM fitted with EM."""
 
     def _init_expectations(self, data):
         X = data[0]

### 2019-10-30 deconv/gmm/online_deconv_gmm.py

@@ -6,6 +6,7 @@ from .util import minibatch_k_means
 
 
 class OnlineDeconvGMM(DeconvGMM):
+    """Implementation of a deconvolving GMM fitted with minibatch-EM."""
 
     def __init__(self, components, dimensions, epochs=1000, w=1e-6,
                  tol=1e-6, step_size=0.1, batch_size=100, restarts=5,

### 2019-10-30 deconv/gmm/plotting.py

@@ -4,7 +4,7 @@ import matplotlib as mpl
 
 def plot_covariance(mean, cov, ax, alpha=0.5, color=None):
     """
-    Plots a Gaussian convariance on a Matplotlib axis
+    Plot a Gaussian convariance on a Matplotlib axis.
 
     Adapted from https://scikit-learn.org/stable/auto_examples/
     mixture/plot_gmm_covariances.html

### 2019-10-30 deconv/gmm/sgd_deconv_gmm.py

@@ -1,15 +1,13 @@
 import torch
 import torch.distributions as dist
-import torch.nn as nn
-import torch.utils.data as data_utils
 
 from .sgd_gmm import SGDGMMModule, BaseSGDGMM
-from .util import minibatch_k_means
 
 mvn = dist.multivariate_normal.MultivariateNormal
 
 
 class SGDDeconvGMMModule(SGDGMMModule):
+    """Implementation of a deconvolving GMM as a PyTorch nn module."""
 
     def forward(self, data):
         x, noise_covars = data
@@ -28,20 +26,8 @@ class SGDDeconvGMMModule(SGDGMMModule):
         return -1 * torch.sum(log_prob)
 
 
-class SGDDeconvDataset(data_utils.Dataset):
-
-    def __init__(self, X, noise_covars):
-        self.X = X
-        self.noise_covars = noise_covars
-
-    def __len__(self):
-        return self.X.shape[0]
-
-    def __getitem__(self, i):
-        return (self.X[i, :], self.noise_covars[i, :, :])
-
-
 class SGDDeconvGMM(BaseSGDGMM):
+    """Concrete implementation of class to fit a deconvolving GMM with SGD."""
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
                  batch_size=64, tol=1e-6, w=1e-3, restarts=5,

### 2019-10-30 deconv/gmm/sgd_gmm.py

@@ -12,6 +12,7 @@ mvn = dist.multivariate_normal.MultivariateNormal
 
 
 class SGDGMMModule(nn.Module):
+    """Implementation of a standard GMM as a PyTorch nn module."""
 
     def __init__(self, components, dimensions, w, device=None):
         super().__init__()
@@ -63,6 +64,7 @@ class SGDGMMModule(nn.Module):
 
 
 class BaseSGDGMM(ABC):
+    """ABC for fitting a PyTorch nn-based GMM."""
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
                  batch_size=64, tol=1e-6, restarts=5, max_no_improvement=20,
@@ -110,7 +112,7 @@ class BaseSGDGMM(ABC):
         return l.sum()
 
     def fit(self, data, val_data=None, verbose=False, interval=1):
-
+        """Fit the GMM to data."""
         n_total = len(data)
 
         init_loader = data_utils.DataLoader(
@@ -131,7 +133,7 @@ class BaseSGDGMM(ABC):
 
         best_loss = float('inf')
 
-        for j in range(self.restarts):
+        for q in range(self.restarts):
 
             self.init_params(init_loader)
 
@@ -248,6 +250,7 @@ class BaseSGDGMM(ABC):
 
 
 class SGDGMM(BaseSGDGMM):
+    """Concrete implementation of class to fit a standard GMM with SGD."""
 
     def __init__(self, components, dimensions, epochs=10000, lr=1e-3,
                  batch_size=64, tol=1e-6, w=1e-3, device=None):

### 2019-10-30 deconv/gmm/util.py

@@ -2,7 +2,7 @@ import torch
 
 
 def k_means(X, k, max_iters=50, tol=1e-9, device=None):
-
+        """Do standard k-means clustering."""
         n, d = X.shape
 
         x_min = torch.min(X, dim=0)[0]
@@ -34,7 +34,11 @@ def k_means(X, k, max_iters=50, tol=1e-9, device=None):
 
 
 def minibatch_k_means(loader, k, max_iters=50, tol=1e-3, device=None):
+    """
+    Do minibatch version of k-means
 
+    Based on https://www.eecs.tufts.edu/~dsculley/papers/fastkmeans.pdf
+    """
     centroids = next(iter(loader))[0][:k].to(device)
     counts = torch.ones(k, device=device)
 

### 2019-10-30 ml4ps_paper/paper.tex
Final paper tweaks.

@@
 \subsubsection*{Acknowledgments}
-We are grateful to David W. Hogg for providing us with the problem and advising us on our experimental setup.
-Our experiments made use of AstroPy~\citep{astropy:2013,astropy:2018}, corner.py~\citep{corner}, Numpy~\citep{numpy}, Pandas~\citep{mckinney-proc-scipy-2010}, PyTorch~\citep{paszke2017automatic} and Scikit-learn~\citep{scikit-learn}.
+We are grateful to David W. Hogg for providing us with the problem and advising us on the Gaia data.
+Our experiments made use of AstroPy~\citep{astropy:2013,astropy:2018}, corner.py~\citep{corner}, Matplotlib~\citep{Hunter:2007}, Numpy~\citep{numpy}, Pandas~\citep{mckinney-proc-scipy-2010}, PyTorch~\citep{paszke2017automatic} and Scikit-learn~\citep{scikit-learn}.
 This work was supported in part by the EPSRC Centre for Doctoral Training in Data Science, funded by the UK Engineering and Physical Sciences Research Council (grant EP/L016427/1) and the University of Edinburgh.
@@
 Here we provide specific details of our experiments for reproducibility.
-Code used to run the experiments will be made available.
+Code used to run the experiments is available at \url{https://github.com/bayesiains/scalable_xd}.
 

### 2019-11-18 poster/poster.tex
Add poster base.

% Gemini theme
% https://github.com/anishathalye/gemini
%
% We try to keep this Overleaf template in sync with the canonical source on
% GitHub, but it's recommended that you obtain the template directly from
% GitHub to ensure that you are using the latest version.

\documentclass[final]{beamer}

% ====================
% Packages
% ====================

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[size=custom,width=84.1,height=118.9,scale=1.175]{beamerposter}
\usetheme{gemini}
\usecolortheme{edinburgh}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage[numbers]{natbib}
\usetikzlibrary{shapes, arrows, positioning, fit, calc, arrows.meta}
\usetikzlibrary{bayesnet, positioning}
\tikzset{>=latex}

\bibliographystyle{abbrv}

% ====================
% Lengths
% ====================

% If you have N columns, choose \sepwidth and \colwidth such that
% (N+1)*\sepwidth + N*\colwidth = \paperwidth
\newlength{\sepwidth}
\newlength{\colwidth}
\setlength{\sepwidth}{0.0333\paperwidth}
\setlength{\colwidth}{0.45\paperwidth}

\newcommand{\separatorcolumn}{\begin{column}{\sepwidth}\end{column}}

\definecolor{phicolor}{HTML}{0074D9}
\definecolor{gammacolor}{HTML}{FF851B}

% Tikz options
\tikzset{%
  var/.style      = {draw, thick, circle, minimum size=0.6cm, inner sep=0},
  block/.style    = {draw, thick, rectangle, rounded corners},
  customline/.style    = {draw opacity=0.4, line width=1}
}
\tikzstyle{arrow} = [draw,line width=0.8pt, black, >=triangle 45]

\definecolor{b}{rgb}{0, 0, 0}
\definecolor{g}{rgb}{0, 0, 0}
\definecolor{r}{rgb}{0, 0, 0}
\definecolor{m}{rgb}{0, 0, 0}

% ====================
% Title
% ====================

\title{Scalable Extreme Deconvolution}

\author{James A. Ritchie, Iain Murray}

\institute[]{School of Informatics, University of Edinburgh}

% ====================
% Body
% ====================

\begin{document}

\addtobeamertemplate{headline}{} 
{
    \begin{tikzpicture}[remember picture,overlay] 
    \node [anchor=north west, inner sep=3cm] at ([xshift=0.cm, yshift=1.cm]current page.north west)     
    {\includegraphics[height=4.cm]{figures/uoe-informatics-white.png}}; 
    \end{tikzpicture} 
}

\addtobeamertemplate{headline}{} 
{
    \begin{tikzpicture}[remember picture,overlay] 
    \node [anchor=north east, inner sep=3cm] at ([xshift=0.cm, yshift=1.cm]current page.north east)     
    {\includegraphics[height=3.cm]{figures/cdt-inverted.eps}}; 
    \end{tikzpicture} 
}

\addtobeamertemplate{headline}{} 
{
    \begin{tikzpicture}[remember picture,overlay] 
    \node [anchor=north east, inner sep=3cm] at ([xshift=0.cm, yshift=-3.cm]current page.north east)     
    {\includegraphics[height=3.cm]{figures/epsrc_logo.pdf}}; 
    \end{tikzpicture} 
}

\begin{frame}[t]
\begin{columns}[t]
\separatorcolumn

\begin{column}{\colwidth}

    \begin{alertblock}{Summary}
    \end{alertblock}

    \begin{block}{Column 1}
    \end{block}

\end{column}

\separatorcolumn

\begin{column}{\colwidth}

    \begin{block}{Column 2}
    \end{block}

    \begin{block}{Acknowledgements}
    \end{block}

    \begin{block}{References}

    \nocite{*}
    \footnotesize{\bibliography{poster}}

    \end{block}

\end{column}

\separatorcolumn
\end{columns}
\end{frame}

\end{document}

### 2020-01-04 poster/poster.tex
Poster tweaks.
Final poster commits.

@@
 \usepackage{lmodern}
-\usepackage[size=custom,width=84.1,height=118.9,scale=1.175]{beamerposter}
+\usepackage[size=custom,width=84.1,height=118.9,scale=1.2]{beamerposter}
 \usetheme{gemini}
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{abbrvnat}
 
@@
 \newlength{\colwidth}
+\newlength{\doublewidth}
 \setlength{\sepwidth}{0.0333\paperwidth}
 \setlength{\colwidth}{0.45\paperwidth}
+\setlength{\doublewidth}{0.933\paperwidth}
+
+\setlength{\abovecaptionskip}{0.5ex}
 
@@
 \definecolor{m}{rgb}{0, 0, 0}
+
+\newcommand{\bx}{\mathbf{x}}
+\newcommand{\bv}{\mathbf{v}}
+\newcommand{\bm}{\mathbf{m}}
+\newcommand{\bb}{\mathbf{b}}
+\newcommand{\bc}{\mathbf{c}}
+\newcommand{\bd}{\mathbf{d}}
+\newcommand{\bs}{\mathbf{s}}
 
@@
 \begin{frame}[t]
+
+\begin{columns}[t]
+
+\separatorcolumn
+\begin{column}{\doublewidth}
+\vspace{-3ex}
+\begin{figure}
+        \centering
+        \includegraphics{figures/density.pdf}
+        \caption{Marginalised log-density plot of an extreme deconvolution model with $K=256$ components, fitted with our minibatch EM method to a subset of the Gaia astronomical catalogue.
+The plot shows the estimated density of star positions on the sky, and has correctly recovered the structure of the Milky Way and the Magellanic Clouds.}
+\end{figure}
+\vspace*{2ex}
+\end{column}
+
+\separatorcolumn
+
+\end{columns}
+
 \begin{columns}[t]
@@
     \begin{alertblock}{Summary}
+        \begin{itemize} 
+            \item \textbf{Extreme Deconvolution} is a method for fitting \textbf{Gassian Mixture Models} to noisy data, originally intended for use with astronomical datasets.
+            \item We propose two \textbf{stochastic minibatch methods} for fitting such models which can run on GPUs. 
+            \item Our methods \textbf{converge faster} and take less time to fit with \textbf{larger models} compared to the existing batch method.
+        \end{itemize}
     \end{alertblock}
 
-    \begin{block}{Column 1}
+    \begin{block}{Extreme Deconvolution}
+        We have noisy $d$-dimensional data $\{\bx_i\}_{i=0}^N$,
+        \begin{equation}
+            \bx_i = \bv_i + \epsilon_i,\quad  \epsilon_i \sim \mathcal{N}(\mathbf{0}, S_i),
+        \end{equation}
+        where we know the noise covariance $S_i$.
+        We want to fit a Gaussian Mixture Model (GMM) with $K$ components to $\bv$,
+        \begin{equation}
+        p(\bv_i \mid \theta) = \sum_j^K \alpha_j \,\mathcal{N}(\bv_i \mid \bm_j, V_j), \quad \theta = \{\alpha_j, \bm_j, V_j\}_{j=1} ^ K.
+        \end{equation}
+        As the noise model is Gaussian and the model of the underlying density is a mixture of Gaussians, the probability of $\bx_i$ is also a Gaussian mixture with total log-likelihood
+        \begin{equation}
+        \mathcal{L}(\theta) = \sum_i^N \log \sum_j^K \alpha_j\,\mathcal{N}(\bx_i \mid \bm_j, T_{ij}), \quad T_{ij} = V_j + S_i.
+        \end{equation}
+        \citet{bovyExtremeDeconvolutionInferring2011} present a batch method for fitting this model using an Expectation-Maximisation (EM) method, originally intended for astronomical datasets.
+        We propose two new methods for fitting this model that scale to larger models for use with larger datasets.
+    \end{block}
+
+    \begin{block}{Method 1: Minibatch Expectation-Maximisation}
+        \begin{itemize}
+        \item Following~\citet{cappeOnlineExpectationMaximization2009}, we develop a minibatch version of the EM procedure.
+        \item At each iteration $t$ compute sums of expected sufficient statistics $\phi_{jt}$ per mixture component $j$ over a minibatch.
+        \item Stochastic estimates $\hat{\phi}_{jt}$ then updated with a sufficiently small step size $\lambda$,
+        \begin{equation}
+        \hat{\phi}_{jt} = (1 - \lambda)\hat{\phi}_{j(t-1)} + \lambda \phi_{jt}.
+        \end{equation}
+        \item Normalise $\hat{\phi}_{jt}$ after each iteration to update the parameters.
+        \item If we set $\lambda=1$ and replace each minibatch with the entire dataset, then the update corresponds to the original batch fitting method.
+        \end{itemize}
+    \end{block}
+
+    \begin{block}{Method 2: Stochastic Gradient Descent}
+        \begin{itemize}
+            \item We can also optimize the likelihood directly with respect to the parameters $\theta$.
+            \item Following \citet{williams1996}, we transform the parameters to remove constraints.
+            \begin{itemize}
+                \item For the mixture weights $\alpha_j$, we take a softmax of an unconstrained vector $\mathbf{z}$,
+                \begin{equation}
+                \alpha_j = \frac{e^{z_j}}{\sum_{k=1}^K e^{z_k}}.
+                \end{equation}
+                \item For the covariances $V_j$, we represent them as a lower triangular Cholesky decomposition $L_j$, where the diagonal elements $qq$ of $L_j$ are constrained positive by taking the exponential of unconstrained elements $\tilde{L}_q$,
+                \begin{equation}
+                V_j = L_jL_j^T, \quad
+                (L_j)_{qq} = \exp({\tilde{L}_q}).
+                \end{equation}
+            \end{itemize}
+            \item With no constraints, we can use any standard minibatch gradient-based optimiser.
+        \end{itemize}
     \end{block}
@@
 \begin{column}{\colwidth}
+    \vskip0.5ex
+    \begin{block}{Experiments}
 
-    \begin{block}{Column 2}
-    \end{block}
+        \begin{itemize}
+        \item We fitted extreme deconvolution models to a random subset of data from the Gaia DR2 Catalogue~\cite{brownGaiaDataRelease2018}.
+        \item Our methods running on GPUs were compared against the reference implementation provided by~\citet{bovyExtremeDeconvolutionInferring2011} running on CPU.
+        \item Both of our methods converged faster and took less time to fit, with slightly improved test log-likelihoods.
+        \item The SGD method was more numerically stable than the minibatch EM method.
+        \item More experiments are needed to demonstrate the utility of our methods for real astronomical problems.
+        \end{itemize}
 
-    \begin{block}{Acknowledgements}
+        \begin{figure}
+        \centering
+        \includegraphics{figures/plots/epochs.pdf}
+        \caption{Average training log-likelihood as a function of training epoch for models with $K=256$ components.
+Epochs rescaled by average training time.
+Error bars indicate $\pm$ 2 standard deviations.}
+        \end{figure}
+
+        \begin{figure}
+        \centering
+        \includegraphics{figures/plots/learning.pdf}
+        \caption{Training time as a function of mixture components $K$. Error bars indicate $\pm$ 2 standard deviations.}
+        \end{figure}
     \end{block}
@@
 
-    \nocite{*}
     \footnotesize{\bibliography{poster}}

### 2020-01-04 poster/figures/plots/learning.py

@@ -1,0 +1,71 @@
+import os
+import numpy as np
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes, mark_inset
+import seaborn as sns
+
+from results_table import get_table
+
+sns.set()
+sns.set_context('poster')
+sns.set_palette('colorblind')
+
+fig, ax = plt.subplots(1, 1, figsize=(12, 6)) 
+
+table = get_table('../results/')
+
+ax.set_xlabel('Time-scaled Epoch')
+ax.set_ylabel('Avg. Log-likelihood')
+
+labels = ('Existing EM', 'Minibatch EM', 'SGD')
+
+result_files = os.listdir('../results/')
+baseline_curves = np.array([
+    np.loadtxt('../results/' + f) for f in result_files if f.startswith(
+        'baseline_256'
+    ) and f.endswith('loglike.log')
+])
+
+linestyles = ('-', '--', '-.')
+
+for l, ls, result in zip(labels, linestyles, table.values()):
+    if l != 'Existing EM':
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
+        y = result[256][2].mean(axis=0)
+        y_err = 1.96 * result[256][2].std(axis=0)
+    else:
+        x = np.arange(1, 21) * (result[256][1].mean() / 20)
+        y = baseline_curves.mean(axis=0)
+        y_err = 1.96 * baseline_curves.std(axis=0)
+
+    ax.plot(x, y, linestyle=ls, label=l)
+    ax.fill_between(x, y + y_err, y - y_err, alpha=0.5)
+
+ax.set_xscale('log')
+ax.set_ylim(-28, -25)
+ax.legend()
+
+fig.tight_layout()
+
+fig.savefig('figures/plots/epochs.pdf')
+
+fig, ax = plt.subplots(1, 1, figsize=(12, 6))
+
+k = np.array([64, 128, 256, 512])
+
+ax.set_xlabel(r'Mixture Components $K$')
+ax.set_ylabel(r'Time (minutes)')
+ax.set_xticks(k)
+
+for l, ls, result in zip(labels, linestyles, table.values()):
+    t = np.array([[r[1].mean(), r[1].std()] for r in result.values()])
+    print(t.shape)
+    ax.errorbar(k, t[:, 0], 1.96 * t[:, 1], label=l, linestyle=ls)
+
[…]

### 2020-01-04 poster/figures/plots/ra_dec_projection.py

@@ -1,0 +1,62 @@
+import numpy as np
+import torch
+
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+import seaborn as sns
+
+sns.set()
+sns.set_context('poster')
+
+weights, means, covars = torch.load(
+    '../results/em_256_508798_params.pkl',
+    map_location=torch.device('cpu')
+)
+
+x_width = 6000
+y_width = 3000
+
+x = np.linspace(0, 360, num=x_width, dtype=np.float32)
+y = np.linspace(-90, 90, num=y_width, dtype=np.float32)
+
+xx, yy = np.meshgrid(x, y)
+
+d = torch.tensor(np.concatenate((xx[:, :, None], yy[:, :, None]), axis=-1))
+
+z = np.zeros((y_width, x_width))
+
+mvn = torch.distributions.MultivariateNormal(
+    loc=means[:, :2], covariance_matrix=covars[:, :2, :2]
+)
+
+for i in range(x_width):
+    z[:, i] = torch.logsumexp(
+        mvn.log_prob(d[:, i, None, :]) + torch.log(weights[:, 0]),
+        axis=-1
+    )
+
+
+fig, ax = plt.subplots(figsize=(24, 12))
+
+cmap = sns.cubehelix_palette(8, as_cmap=True)
+
+im = ax.imshow(z, extent=[0, 360, -90, 90], origin='lower')
+
+divider = make_axes_locatable(ax)
+cax = divider.append_axes("right", size="5%", pad=0.05)
+
+ax.set_xlim(0, 360)
+ax.set_xticks(np.arange(0, 361, 60))
+
+ax.set_ylim(-90, 90)
+ax.set_yticks(np.arange(-90, 91, 30))
+
+ax.grid(False)
+fig.colorbar(im, cax=cax)
+
+ax.set_xlabel(r'Right Ascension ($\circ$)')
+ax.set_ylabel(r'Declination ($\circ$)')
+
+fig.tight_layout()
+
+fig.savefig('figures/density.pdf')

### 2020-01-04 poster/figures/plots/results_table.py

@@ -1,0 +1,119 @@
+import argparse
+import json
+import os
+
+import numpy as np
+
+MODELS = [
+    'baseline',
+    'em',
+    'sgd'
+]
+
+SIZES = [
+    64,
+    128,
+    256,
+    512
+]
+
+def get_results(files, model, size, file_dir):
+    return [
+        json.load(open(str(file_dir) + f)) for f in files if f.startswith(
+            '{}_{}'.format(model, size)
+        ) and f.endswith('.json')
+    ]
+
+
+def get_table(results_dir):
+
+    results_files = os.listdir(str(results_dir))
+
+    table = {}
+
+    for model in MODELS:
+        table[model] = {}
+        for size in SIZES:
+            results = get_results(results_files, model, size, results_dir)
+            scores = np.array([r['val_score'] for r in results]) / 200000
+            times = np.array(
+                [r['end_time'] - r['start_time'] for r in results]
+            ) / 60
+            if model != 'baseline':
+                curve = np.array([r['train_curve'] for r in results]) / 1600000
+            else:
+                curve = None
+
+            if model == 'sgd':
+                scores = scores * -1
+                curve = curve * -1
+
+            if size == 512:
+                test_scores = np.array(json.load(
+                    open(results_dir + '{}_test_scores.json'.format(
+                        model
+                    ))
+                ))
+
+                if model != 'baseline':
+                    test_scores /= 200000
+
+                if model == 'sgd':
+                    test_scores *= -1
+
+                table[model][size] = (
+                    scores, times, curve, test_scores
+                )
+            else:
+                table[model][size] = (scores, times, curve)
+
+    return table
+
+def print_table(table):
+
+    print(r'\toprule')
+    print(r'Method     & K &  Validation     & Test\\')
+
+    labels = (
+        'Existing EM',
+        'Minibatch EM',
+        'SGD'
+    )
+
+    for l, m in zip(labels, table.values()):
+        print(r'\midrule')
+        for i, (k, r) in enumerate(m.items()):
+
+            if k == 512:
+                s = ' & {} & ${:.2f} \pm {:.2f}$ & ${:.2f} \pm {:.2f}$ \\\\'.format(
+                    k,
+                    r[0].mean(),
+                    r[0].std(),
+                    r[3].mean(),
+                    r[3].std()
+                )
+            else:
+                s = ' & {} & ${:.2f} \pm {:.2f}$ & - \\\\'.format(
[…]
